---
title: "CDISC Analysis Results Standard (ARS) — Interactive Training"
subtitle: "Chapter 6: R Ecosystem and Tools"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Chapter 6 maps the open-source R ecosystem to the ARS pipeline: {cards} and
  {cardx} cover the ARD layer; {gtsummary} and {rtables} cover the display
  layer; {datasetjson} covers dataset exchange — but ARM-TS parsing,
  generation, and validation are entirely uncovered. You will build a full
  demographics table from ARD → gtsummary on real pharmaverse data and
  enumerate what the ecosystem still lacks.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
suppressWarnings(suppressMessages({
  has_cardx     <- requireNamespace("cardx",     quietly = TRUE)
  has_gtsummary <- requireNamespace("gtsummary", quietly = TRUE)
  if (has_cardx)     library(cardx)
  if (has_gtsummary) library(gtsummary)
}))
library(jsonlite)
library(glue)

tutorial_options(exercise.timelimit = 60)
gradethis_setup()

# Load pharmaverse ADaM datasets
data("adsl",  package = "pharmaverseadam")
data("adae",  package = "pharmaverseadam")
# pharmaverseadam 1.4+ replaced the generic "adtte" with TA-specific
# datasets; use the oncology one and join subject-level vars in from
# ADSL. pharmaverseadam ADSL lacks ITTFL, so derive it from ARM (a
# subject is "randomised / ITT" iff ARM is non-empty).
data("adtte_onco", package = "pharmaverseadam")
adtte <- adtte_onco %>%
  left_join(
    adsl %>%
      mutate(ITTFL = ifelse(!is.na(ARM) & ARM != "", "Y", "N")) %>%
      select(USUBJID, TRT01A, ITTFL),
    by = "USUBJID"
  )
data("adlb",  package = "pharmaverseadam")

adsl_safe <- adsl %>% filter(SAFFL == "Y")
adsl_itt  <- adsl %>% filter(!is.na(ARM) & ARM != "")
adae_teae <- adae %>% filter(TRTEMFL == "Y")

re_json <- jsonlite::read_json("data/reporting_event.json")
```

---

## Chapter 6: R Ecosystem and Tools

### 6.1 The Coverage Picture

The R ecosystem for ARS covers **three parts of the pipeline well** and **one part poorly**.
Understanding exactly what each tool does — and where it stops — is essential before
designing `arsbridge`.

```
SAP ──→ ARM-TS ──→ ADaM ──→ ARD ──→ Display
          ↑                  ↑          ↑
     [NOT COVERED]     [{cards}]   [{gtsummary}]
     [arsbridge]       [{cardx}]   [{rtables}]
          ↑                           ↑
     [NOT COVERED]            [{datasetjson}]
     [arsbridge]               (exchange format)
```

The diagonal arrows from `[NOT COVERED]` show where `arsbridge` will sit.

---

### 6.2 `{cards}` — CDISC Analysis Results Data

**Repository:** github.com/insightsengineering/cards
**Maintainers:** Roche, GSK, Novartis (Insights Engineering)
**On CRAN:** yes
**Purpose:** generate ARD data frames from R data

**Core functions**

| Function | Purpose | Example use |
|----------|---------|-------------|
| `ard_continuous()` | n, mean, sd, median, min, max, quantiles | Age, lab values, vital signs |
| `ard_categorical()` | Frequency tables | AE preferred terms, sex, race |
| `ard_dichotomous()` | Binary counts and percentages | Any-TEAE flag, response flag |
| `ard_missing()` | Missing-data summaries | Missing lab values |
| `ard_total_n()` | Total N per group | Denominators for percentages |
| `ard_stack()` | Combine multiple ARDs | Build a full table's ARD |
| `ard_fmt_args()` | Set formatting arguments | Apply display formats |

**What `{cards}` does NOT do**

- It does **not** read or parse ARM-TS JSON
- It does **not** know about AnalysisSets — you apply population filters manually before
  calling it
- It does **not** know about `GroupingFactor` definitions — you pass the grouping variable
  directly as `by =`
- It does **not** link the ARD back to the analysis specification
- It does **not** produce ARM-TS from annotated shells

---

### 6.3 `{cardx}` — Extended ARD Utilities

**Repository:** github.com/insightsengineering/cardx
**Purpose:** complex statistical analyses for ARD generation

`{cardx}` extends `{cards}` with statistics that need extra packages.

| Function | Purpose |
|----------|---------|
| `ard_survival_survfit()` | Kaplan-Meier estimates at fixed timepoints |
| `ard_regression()` | Logistic / Cox regression model summaries |
| `ard_stats_t_test()` | t-test ARD |
| `ard_stats_wilcox_test()` | Wilcoxon ARD |
| `ard_stats_chisq_test()` | Chi-square test ARD |
| `ard_stats_fisher_test()` | Fisher's exact test |
| `ard_cmh_test()` | Cochran-Mantel-Haenszel (stratified) |

Like `{cards}`, it reads neither ARM-TS nor any analysis specification — you still apply
population and data-subset filters manually.

---

### 6.4 `{gtsummary}` — Tables from ARDs

**Repository:** github.com/ddsjoberg/gtsummary
**Purpose:** render clinical-trial tables from ARD data frames

In recent versions (3.x), `{gtsummary}` has been refactored to use ARD as its core data
structure. The pattern is `ard_stack(...) %>% tbl_ard_summary(...)`.

**What `{gtsummary}` does NOT do**

- It does not produce ARM-TS
- It does not link `OutputDisplay` sections (title, footnotes) to the formal model
- It does not validate that the table matches what ARM-TS specifies

---

### 6.5 `{datasetjson}` — CDISC Dataset-JSON Exchange Format

**Repository:** github.com/atorus-research/datasetjson
**On CRAN:** yes (v0.3.0)
**Purpose:** read/write CDISC Dataset-JSON for regulatory submission

Dataset-JSON is the new exchange format the FDA is evaluating as a **replacement for SAS XPT**.
It serializes SDTM/ADaM datasets as JSON with CDISC-compliant metadata. The R Consortium
used it for the **FDA Pilot 4 submission (fall 2025)**.

**Relevance to ARS.** Dataset-JSON handles the **dataset exchange layer**. ARM-TS is a
separate JSON structure that lives **alongside** Dataset-JSON — ARM-TS encodes the *analysis
specification*, Dataset-JSON encodes the *datasets*.

---

### 6.6 The ARM-TS Parser Gap

**None** of the above tools read or write ARM-TS JSON. There is no R function in the open
source ecosystem that does any of the following:

- Parse an ARM-TS JSON file into an R list/object
- Extract `AnalysisSet` conditions and apply them as dplyr filters
- Generate an ARM-TS JSON from a structured annotated shell
- Render an ARM-TS JSON as a human-readable annotated shell
- Validate that an ARD is consistent with its ARM-TS specification

The CDISC model YAML (`ars_ldm.yaml`) defines the schema. But no R package implements:

```r
parse_ars("reporting_event.json")        # → R list of ARS objects
ars_to_shell(re, output_id = "OUT001")   # → Excel annotated shell
shell_to_ars("annotated_shell.xlsx")     # → ARM-TS JSON
validate_ard(re, ard)                    # → traceability report
```

This is what `arsbridge` builds. Chapter 7 covers the design in detail.

---

### 6.7 `siera` (Clymb Clinical) — Closed-Source Comparison

For completeness: `siera` is a commercial R package from **Clymb Clinical** that:

- Automates ARD generation
- Claims traceability linking
- Integrates with their eTFL portal

It is proprietary, not on CRAN, and requires a commercial license. It is the closest thing
to what `arsbridge` would do — but it is **not open source**, not pharmaverse-aligned, and is
developed by the same team (Bhavin Busa) who leads the ARS standard. The open-source
community version of this capability does not exist.

---

### 6.8 Tool Coverage Summary

| Capability | Tool | Coverage |
|------------|------|----------|
| Generate ARD from ADaM | `{cards}` + `{cardx}` | ✅ Good |
| Render tables from ARD | `{gtsummary}`, `{rtables}` | ✅ Good |
| Exchange datasets as JSON | `{datasetjson}` | ✅ Good |
| Parse ARM-TS JSON into R | — | ❌ Missing |
| Generate ARM-TS from annotated shell | — | ❌ Missing |
| Render ARM-TS as annotated shell | — | ❌ Missing |
| Validate ARD against ARM-TS | — | ❌ Missing |
| Apply AnalysisSet from ARM-TS to data | — | ❌ Missing |

The bottom **four** rows are `arsbridge`.

---

## Hands-On: Use the Existing Stack End-to-End

The exercises below assemble the parts that **do** exist (`{cards}` → `{cardx}` →
`{gtsummary}`) — then enumerate what's still missing.

---

## Exercise 6.1 — Build a Multi-Variable ARD with `ard_stack()` {data-progressive=FALSE}

Real demographic tables combine continuous and categorical variables. Use `ard_stack()` to
build an ARD that covers **AGE (continuous)** and **SEX (categorical)** by treatment arm in
the Safety Population.

```{r ch6-ex1, exercise=TRUE, exercise.lines=18}
ard_demog <- ard_stack(
  # Continuous: AGE
  ard_continuous(
    data      = adsl_safe,
    by        = ___,             # treatment arm
    variables = ___              # which variable?
  ),
  # Categorical: SEX
  ard_categorical(
    data      = adsl_safe,
    by        = ___,             # treatment arm
    variables = ___              # which variable?
  )
)

cat("ard_demog rows:", nrow(ard_demog), "\n")
cat("Distinct variables in ARD:\n")
unique(ard_demog$variable)
```

```{r ch6-ex1-hint-1}
# by        = TRT01A
# variables = AGE   (for continuous)
# variables = SEX   (for categorical)
```

```{r ch6-ex1-solution}
ard_demog <- ard_stack(
  ard_continuous(
    data = adsl_safe, by = TRT01A, variables = AGE
  ),
  ard_categorical(
    data = adsl_safe, by = TRT01A, variables = SEX
  )
)
cat("ard_demog rows:", nrow(ard_demog), "\n")
unique(ard_demog$variable)
# ard_stack() unions multiple ARDs into one tall data frame.
# This is the unit you hand to {gtsummary} to render a full demographics table.
```

---

## Exercise 6.2 — Render the ARD as a `{gtsummary}` Table {data-progressive=FALSE}

Wrap `ard_demog` into a formatted demographics table with `tbl_ard_summary()` and add a
caption. (The cell renders the table as text — but in a real Rmd it would be a fully styled
HTML/RTF table.)

```{r ch6-ex2-setup, include=FALSE}
ard_demog <- ard_stack(
  ard_continuous(data = adsl_safe, by = TRT01A, variables = AGE),
  ard_categorical(data = adsl_safe, by = TRT01A, variables = SEX)
)
```

```{r ch6-ex2, exercise=TRUE, exercise.lines=18, exercise.setup="ch6-ex2-setup"}
if (!requireNamespace("gtsummary", quietly = TRUE)) {
  cat("Install gtsummary to run this exercise: install.packages('gtsummary')\n")
} else {
  library(gtsummary)

  tbl <- tbl_ard_summary(
    cards     = ard_demog,
    by        = ___,                   # the grouping variable
    variables = c(___, ___)            # the two analysis variables
  ) %>%
    modify_caption("Table 14.1.1 — Demographic Characteristics (Safety Population)")

  tbl
}
```

```{r ch6-ex2-hint-1}
# by        = TRT01A
# variables = c(AGE, SEX)
```

```{r ch6-ex2-solution}
if (!requireNamespace("gtsummary", quietly = TRUE)) {
  cat("Install gtsummary to run this exercise.\n")
} else {
  library(gtsummary)
  tbl <- tbl_ard_summary(
    cards     = ard_demog,
    by        = TRT01A,
    variables = c(AGE, SEX)
  ) %>%
    modify_caption("Table 14.1.1 — Demographic Characteristics (Safety Population)")
  tbl
}
# This is the current state of the art end-to-end in open source R:
# {cards} produces the ARD, {gtsummary} renders the table.
# What's NOT here: any link back to ARM-TS — no record of WHY the Safety Pop
# was chosen, no SAP reference, no AnalysisSet ID, no analysis purpose flag.
# That's the arsbridge gap.
```

---

## Exercise 6.3 — Identify the Tool for Each Capability {data-progressive=FALSE}

Fill in the right tool (or `"MISSING / arsbridge"`) for each ARS-related capability.

```{r ch6-ex3, exercise=TRUE, exercise.lines=20}
tool_coverage <- tribble(
  ~capability,                                        ~tool,
  "Generate ARD for continuous variables",            ___,
  "Generate KM ARD at fixed timepoints",              ___,
  "Render an ARD as a clinical table",                ___,
  "Write an ADaM dataset as Dataset-JSON for FDA",    ___,
  "Parse an ARM-TS JSON file into R",                 ___,
  "Generate ARM-TS from an annotated shell",          ___,
  "Validate ARD against ARM-TS",                      ___
)

tool_coverage
```

```{r ch6-ex3-hint-1}
# Choose from:
#   "{cards}"
#   "{cardx}"
#   "{gtsummary}"
#   "{datasetjson}"
#   "MISSING / arsbridge"
```

```{r ch6-ex3-solution}
tool_coverage <- tribble(
  ~capability,                                        ~tool,
  "Generate ARD for continuous variables",            "{cards}",
  "Generate KM ARD at fixed timepoints",              "{cardx}",
  "Render an ARD as a clinical table",                "{gtsummary}",
  "Write an ADaM dataset as Dataset-JSON for FDA",    "{datasetjson}",
  "Parse an ARM-TS JSON file into R",                 "MISSING / arsbridge",
  "Generate ARM-TS from an annotated shell",          "MISSING / arsbridge",
  "Validate ARD against ARM-TS",                      "MISSING / arsbridge"
)
tool_coverage
# Three rows of MISSING — and each of them is one arsbridge function:
#   parse_ars(), shell_to_ars(), validate_ard().
```

---

## Exercise 6.4 — Read the AnalysisSet Manually (the gap exercise) {data-progressive=FALSE}

`{cards}` does **not** read ARM-TS. So today, even though the AnalysisSet `AS002` is sitting
right there in `re_json`, you (the programmer) have to **manually translate it** into a
filter. Do exactly that — and notice how mechanical it is. This mechanical step is what
`arsbridge::parse_ars()` will automate.

```{r ch6-ex4, exercise=TRUE, exercise.lines=18}
# Pull the AS002 condition out of the parsed ARM-TS
cond <- re_json$analysisSets[[2]]$condition
cond

# Apply it MANUALLY as a dplyr filter on adsl,
# then hand the filtered data to ard_continuous() for AGE by TRT01A
adsl_filtered <- adsl %>%
  filter(.data[[cond$___]] == cond$value[[1]])     # which field carries the variable name?

ard <- ard_continuous(
  data = adsl_filtered, by = TRT01A, variables = AGE
)

cat("Rows in filtered ADSL:", nrow(adsl_filtered), "\n")
cat("Rows in ARD:          ", nrow(ard), "\n")
```

```{r ch6-ex4-hint-1}
# cond has these fields: dataset, variable, comparator, value
# You want the VARIABLE name to build a dplyr filter on that column
```

```{r ch6-ex4-solution}
cond <- re_json$analysisSets[[2]]$condition
adsl_filtered <- adsl %>%
  filter(.data[[cond$variable]] == cond$value[[1]])

ard <- ard_continuous(
  data = adsl_filtered, by = TRT01A, variables = AGE
)
cat("Rows in filtered ADSL:", nrow(adsl_filtered), "\n")
cat("Rows in ARD:          ", nrow(ard), "\n")
# That filter() call is mechanical translation from a WhereClause.
# arsbridge::parse_ars() will return a $analysis_sets[["AS002"]]$filter_fn
# that you can call directly: adsl %>% filter_fn() — no manual rewriting.
```

---

## Chapter 6 Knowledge Check

```{r ch6-quiz, echo=FALSE}
quiz(
  caption = "Chapter 6 — Check Your Understanding",

  question(
    "Which R package generates ARD data frames from ADaM datasets?",
    answer("{gtsummary}"),
    answer("{cards}", correct = TRUE),
    answer("{datasetjson}"),
    answer("{admiral}"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which package extends {cards} with Kaplan-Meier, regression, and inferential statistics?",
    answer("{cardx}", correct = TRUE),
    answer("{rtables}"),
    answer("{gtsummary}"),
    answer("{tfrmt}"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which package handles dataset exchange (e.g., the FDA Pilot 4 submission format)?",
    answer("{cards}"),
    answer("{datasetjson}", correct = TRUE),
    answer("{admiral}"),
    answer("{gtsummary}"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "What capability is MISSING from the current open-source R ecosystem for ARS?",
    answer("Generating ARDs from continuous variables"),
    answer("Creating tables from ARD"),
    answer(
      "Parsing ARM-TS JSON, applying its AnalysisSets, and validating ARD traceability against it",
      correct = TRUE
    ),
    answer("Reading SAS XPT files"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "When using {cards} today, who is responsible for applying the AnalysisSet (population filter)?",
    answer("{cards} reads the ARM-TS and applies it automatically"),
    answer("The PROGRAMMER applies it manually with dplyr::filter() before calling ard_*()", correct = TRUE),
    answer("{gtsummary} applies it during table rendering"),
    answer("It is applied by the regulator after submission"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Why does the open-source ARM-TS gap matter even though `siera` exists?",
    answer(
      "siera is commercial and closed-source; pharmaverse and academic users need an open-source equivalent",
      correct = TRUE
    ),
    answer("siera does not generate ARDs"),
    answer("siera cannot read SAS datasets"),
    answer("siera does not exist"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which arsbridge function would replace the manual `filter(.data[[cond$variable]] == cond$value)` pattern from Exercise 6.4?",
    answer("ars_to_shell()"),
    answer("parse_ars() (returns ready-to-use filter functions per AnalysisSet)", correct = TRUE),
    answer("validate_ard()"),
    answer("ars_summary()"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

---

## Chapter 6 Summary {.congratulations}

You now know:

- **`{cards}` + `{cardx}`** generate ARDs from ADaM — fully covered
- **`{gtsummary}` and `{rtables}`** render tables from ARDs — fully covered
- **`{datasetjson}`** handles dataset exchange (the format used in FDA Pilot 4) — fully covered
- **The ARM-TS layer** — parsing, generating from shells, rendering as shells, validating
  ARD against ARM-TS — is **entirely uncovered** in open-source R
- **`siera`** is the only existing tool in this space but is commercial, closed, and tied to
  one vendor
- You've assembled the existing stack end-to-end (`ard_stack()` → `tbl_ard_summary()`) and
  felt firsthand the mechanical work that `arsbridge::parse_ars()` will automate

**Next:** Chapter 7 — Package Design Roadmap (the five `arsbridge` functions, the build
sequence, and the launch plan)
