---
title: "CDISC Analysis Results Standard (ARS): Interactive Training"
subtitle: "Chapter 2: The ARS Data Model"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Chapter 2 walks through the complete ARS object hierarchy: ReportingEvent,
  AnalysisSets, DataSubsets, AnalysisGroupings, Methods, Analyses, Outputs,
  and the OperationResult (ARD). You will explore a real ARM-TS JSON,
  translate WhereClauses into dplyr filters, and build object structures
  by hand.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(jsonlite)
library(glue)

# learnr global options
tutorial_options(
  exercise.timelimit = 60
)

# gradethis setup
gradethis_setup()

# Load pharmaverse ADaM datasets
data("adsl",  package = "pharmaverseadam")
data("adae",  package = "pharmaverseadam")
# pharmaverseadam ADSL lacks ITTFL; derive it from ARM (a subject is
# "randomised / ITT" iff ARM is non-empty). Exercises in this chapter
# filter adsl on ITTFL directly, so add it to adsl itself.
adsl <- adsl %>%
  mutate(ITTFL = ifelse(!is.na(ARM) & ARM != "", "Y", "N"))
# pharmaverseadam 1.4+ replaced the generic "adtte" with TA-specific
# datasets; use the oncology one and join subject-level vars in from ADSL.
data("adtte_onco", package = "pharmaverseadam")
adtte <- adtte_onco %>%
  left_join(
    adsl %>% select(USUBJID, TRT01A, ITTFL),
    by = "USUBJID"
  )
data("adlb",  package = "pharmaverseadam")

# Pre-filter datasets used throughout the tutorial
adsl_safe <- adsl %>% filter(SAFFL == "Y")
adsl_itt  <- adsl %>% filter(ITTFL == "Y")
adae_teae <- adae %>% filter(TRTEMFL == "Y")

# Load sample ARM-TS JSON
re_json <- jsonlite::read_json("data/reporting_event.json")
```

---

## Chapter 2: The ARS Data Model

### 2.1 Overview: Why a Logical Model?

The ARS data model is a **logical model**. It defines objects and their relationships,
independent of any specific file format. In practice, ARS is serialized as JSON (or YAML),
but the model itself is format-agnostic. You could store an ARM-TS in a relational database,
in YAML, or in protobuf, and it would still be the same logical structure.

The model has a single root object: the `ReportingEvent`. Everything else hangs off it.

Think of the `ReportingEvent` as one CSR-worth of work, or one interim analysis, or one
regulatory deliverable. Inside it lives every population, every method, every analysis,
and every output associated with that deliverable.

---

### 2.2 The Two Layers Revisited

Before diving into objects, anchor yourself in the two layers introduced in Chapter 1:

```
+---------------------------------------------------------+
|                     ARM-TS Layer                        |
|  (What was planned and how analyses were performed)     |
|                                                         |
|  ReportingEvent                                         |
|  ├── AnalysisSets      (population filters / flags)     |
|  ├── DataSubsets       (data record filters)            |
|  ├── AnalysisGroupings (treatment, sex, visit...)       |
|  ├── Methods           (statistical operations)         |
|  ├── Analyses          (dataset + variable + method)    |
|  └── Outputs           (displays / tables)              |
+---------------------------------------------------------+

+---------------------------------------------------------+
|                      ARD Layer                          |
|  (What was actually computed)                           |
|                                                         |
|  Lives inside Analysis.results                          |
|  OperationResult: operationId + resultGroups +          |
|                   rawValue + formattedValue             |
+---------------------------------------------------------+
```

Here is a detail that surprises most people: the ARD is **not** a separate file in the
formal model. It lives as the `results` slot inside each `Analysis` object. In practice,
teams often extract it and store it as a separate flat dataset (a data frame) for use with
tools like `{cards}`. Chapter 4 covers that workflow in depth.

---

### 2.3 The Complete Object Hierarchy

Here is every class in the ARS model, organized by function. Do not try to memorize this
on the first pass. Just orient yourself; the exercises will make you walk it by hand.

**Root**

```
ReportingEvent
```

The top-level container. Everything for one reporting requirement (for example, one CSR or
one interim analysis) lives inside or is referenced from a `ReportingEvent`.

**Structural / Organizational Objects**

```
ReportingEvent
├── ListOfContents          : structured table of contents for analyses and outputs
│   └── NestedList
│       └── OrderedListItem : points to an Analysis or Output, with level/order
│
├── ReferenceDocument       : external file (SAP PDF, program file, etc.)
│
├── TerminologyExtension    : sponsor-defined terms extending CDISC enumerations
│   └── SponsorTerm
│
└── AnalysisOutputCategorization  : implementer-defined category system
    └── AnalysisOutputCategory    : a single category (e.g., "Safety", "Efficacy")
```

**Population and Data Selection Objects**

These define **who** is included and **which data records** count. This maps directly to
the WHERE clauses in your annotated shells. Note the split carefully, because it matters
throughout the standard and the quiz will test it:

- An **AnalysisSet** filters *subjects*. It is a subject-level population, evaluated
  against ADSL: ITT, Per-Protocol, Safety.
- A **DataSubset** filters *data records* within an analysis dataset: one visit, one
  parameter, one time window. A subject can pass the AnalysisSet but still have individual
  records excluded by a DataSubset.

```
ReportingEvent
├── AnalysisSets            : subject populations (ITT, PP, Safety, etc.)
│   └── WhereClause         : the filter logic (e.g., ADSL.ITTFL EQ 'Y')
│
├── DataSubsets             : data record filters (e.g., a specific visit, PARAMCD)
│   └── WhereClause
│
└── AnalysisGroupings       : factors for subdividing (treatment arm, sex, visit)
    └── GroupingFactor
        └── Group           : one level within the factor (e.g., "Treatment A")
            └── WhereClause
```

**Method Objects**

These define **how** analyses are performed. One method can be reused across many analyses.
That reuse is a core design principle, and we will come back to it in Section 2.6.

```
ReportingEvent
└── Methods
    └── AnalysisMethod       : a named set of statistical operations
        ├── Operations       : one or more statistical operations
        │   └── Operation    : each produces one result
        │       └── ReferencedOperationRelationship  (e.g., NUMERATOR / DENOMINATOR)
        │
        └── AnalysisProgrammingCodeTemplate  : reusable code template with parameters
            └── TemplateCodeParameter
```

**Analysis Objects**

Each `Analysis` is one specific analysis in the study. It ties together a dataset, a
variable, a population, optional data filters, optional groupings, and a method.

```
ReportingEvent
└── Analyses
    └── Analysis
        ├── dataset          : e.g., "ADTTE"
        ├── variable         : e.g., "AVAL"
        ├── analysisSetId    : references an AnalysisSet (population)
        ├── dataSubsetId     : references a DataSubset (optional)
        ├── orderedGroupings : references GroupingFactors (treatment, visit, etc.)
        ├── methodId         : references an AnalysisMethod
        ├── reason           : SPECIFIED IN PROTOCOL / SPECIFIED IN SAP / DATA DRIVEN
        ├── purpose          : PRIMARY / SECONDARY / EXPLORATORY OUTCOME MEASURE
        ├── documentRefs     : links to SAP, program files
        ├── programmingCode  : actual R/SAS code or reference to file
        └── results          : the ARD (OperationResults)
            └── OperationResult
                ├── operationId     : which operation this result is for
                ├── resultGroups    : which group combination
                ├── rawValue        : unformatted result
                └── formattedValue  : result formatted per resultPattern
```

**Output / Display Objects**

These define the **display**: the table itself, its title, footnotes, and how analyses map
into it.

```
ReportingEvent
└── Outputs
    └── Output
        ├── fileSpecifications  : output file format (PDF, RTF, TXT)
        ├── programmingCode     : code to generate the display
        └── displays            : ordered list of displays in the output
            └── OrderedDisplay
                └── OutputDisplay
                    ├── displayTitle      : the table title
                    └── displaySections   : title, header, footnote, abbreviation, footer
                        └── DisplaySection
                            └── OrderedDisplaySubSection
                                └── DisplaySubSection (text)
```

---

### 2.4 The WhereClause System

The `WhereClause` is the most important structural pattern in ARS, because it appears in
**three different contexts**: `AnalysisSet`, `DataSubset`, and `Group`. Master this one
pattern and you understand 60% of the model.

**Simple Condition**

```json
{
  "condition": {
    "dataset": "ADSL",
    "variable": "ITTFL",
    "comparator": "EQ",
    "value": ["Y"]
  }
}
```

This directly maps to `WHERE ADSL.ITTFL = 'Y'` in SAS, or `filter(ITTFL == "Y")` in R.

**Compound Expression**

Real filters are rarely a single condition. A compound expression joins multiple
WhereClauses with a logical operator:

```json
{
  "compoundExpression": {
    "logicalOperator": "AND",
    "whereClauses": [
      {
        "condition": {
          "dataset": "ADSL",
          "variable": "ITTFL",
          "comparator": "EQ",
          "value": ["Y"]
        }
      },
      {
        "condition": {
          "dataset": "ADSL",
          "variable": "AGE",
          "comparator": "GE",
          "value": ["18"]
        }
      }
    ]
  }
}
```

- **Logical operators:** `AND`, `OR`, `NOT`. These three, and only these three.
- **Comparators:** `EQ`, `NE`, `GT`, `GE`, `LT`, `LE`, `IN`, `NOTIN`

Each comparator has a mechanical R equivalent. This mapping is worth internalizing now,
because translating WhereClauses into dplyr filters is exactly what you will do in the
exercises (and what `arsbridge` automates):

| ARS comparator | dplyr equivalent |
|:---------------|:-----------------|
| `EQ`  | `==` |
| `NE`  | `!=` |
| `GT` / `GE` | `>` / `>=` |
| `LT` / `LE` | `<` / `<=` |
| `IN`  | `%in%` |
| `NOTIN` | `!(x %in% ...)` |

**Referenced WhereClause**

A `WhereClause` can also reference another already-defined clause by its ID, for example
reusing an `AnalysisSet` as part of a `Group` definition. This avoids repeating complex
logic and keeps a single source of truth for each filter.

---

### 2.5 The Controlled Terminology

ARS keeps its controlled vocabulary deliberately small: **four extensible enumerations**.
These are the only controlled terms in the model.

| Enumeration | Values | What it encodes |
|-------------|--------|-----------------|
| `AnalysisReasonEnum` | SPECIFIED IN PROTOCOL, SPECIFIED IN SAP, DATA DRIVEN, REQUESTED BY REGULATORY AGENCY | When/why the analysis was planned |
| `AnalysisPurposeEnum` | PRIMARY OUTCOME MEASURE, SECONDARY OUTCOME MEASURE, EXPLORATORY OUTCOME MEASURE | Role of the analysis in the study |
| `OperationRoleEnum` | NUMERATOR, DENOMINATOR | Relationship between operations (for % calculations) |
| `OutputFileTypeEnum` | pdf, rtf, txt | File format of output |

All four are **extensible**: sponsors can add their own terms via `TerminologyExtension`.

One structural detail you will need in Exercise 2.5: in the JSON, fields like `reason` and
`purpose` are not plain strings. Each is a small object holding a `controlledTerm` field
(and, for sponsor extensions, a `sponsorTermId` instead). So in parsed R you reach the
value with `an$reason$controlledTerm`, not `an$reason`.

---

### 2.6 Key Design Principles

**Principle 1: Reuse over repetition**

Methods, AnalysisSets, DataSubsets, and Groups are defined once and referenced by ID from
multiple Analyses. Here is the pattern in JSON, because you will write it constantly.

**Define once**, in the top-level `analysisSets` collection:

```json
"analysisSets": [
  { "id": "AS001",
    "name": "ITT Population",
    "condition": { "dataset": "ADSL", "variable": "ITTFL",
                   "comparator": "EQ", "value": ["Y"] } }
]
```

**Reference everywhere**, from any number of analyses. Note that the analyses carry only
the ID string, never a copy of the condition:

```json
"analyses": [
  { "id": "AN001", "name": "Age Summary",
    "analysisSetId": "AS001",  "...": "..." },

  { "id": "AN002", "name": "Weight Summary",
    "analysisSetId": "AS001",  "...": "..." },

  { "id": "AN003", "name": "Response Rate",
    "analysisSetId": "AS001",  "...": "..." }
]
```

If 50 analyses use the ITT population, there is still exactly one `AS001`, and 50 lines
saying `"analysisSetId": "AS001"`. This is how the model enforces consistency: change the
definition in one place and every analysis that references it follows. It is also why, in
production, a wrong population filter is a one-line fix instead of a 50-analysis hunt.

**Principle 2: Separation of specification from results**

The specification (what to do) and the results (what was computed) live in the same
`Analysis` object. The `results` slot contains `OperationResult` objects, and that is the
ARD. The full context of a result is always navigable from the result itself.

**Principle 3: Ordered everything**

Lists in ARS are ordered explicitly using `level` and `order` integer fields. The
`ListOfContents` uses this to define the table of contents. `OrderedGroupings` uses it to
define the sequence of grouping factors. This matters for display rendering: the model
never relies on implicit list position.

**Principle 4: Extensibility**

Where CDISC controlled terms do not cover a sponsor's needs, the `TerminologyExtension`
mechanism allows adding custom terms while remaining conformant with the standard.

---

### 2.7 Entity-Relationship Summary

```
ReportingEvent (1)
    |
    ├─── (many) AnalysisSet          <- referenced by Analysis.analysisSetId
    ├─── (many) DataSubset           <- referenced by Analysis.dataSubsetId
    ├─── (many) GroupingFactor       <- referenced by OrderedGroupingFactor.groupingId
    │            └── (many) Group    <- referenced by ResultGroup.groupId
    ├─── (many) AnalysisMethod       <- referenced by Analysis.methodId
    │            └── (many) Operation <- referenced by OperationResult.operationId
    ├─── (many) Analysis             <- referenced by OrderedListItem.analysisId
    │            └── (many) OperationResult (ARD lives here)
    ├─── (many) Output               <- referenced by OrderedListItem.outputId
    │            └── (many) OutputDisplay
    └─── (many) ReferenceDocument    <- referenced by DocumentReference.referenceDocumentId
```

Read the arrows as "is referenced by". Every relationship in the model is an ID reference,
never a copy. That is Principle 1 made concrete.

---

## Hands-On: A Real ARM-TS in Your Hands

### Orient to the sample ARM-TS

The sample ARM-TS JSON (`reporting_event.json`) is already loaded as `re_json`, the same
file you met in Chapter 1. Remember from that chapter: parsed JSON becomes a nested R list,
and you reach named fields with `$`.

Take a look at the top-level structure before the exercises. `names()` lists the field
names of the root ReportingEvent; each one corresponds to a branch of the hierarchy diagram
in Section 2.3.

```{r ch2-preview, exercise=FALSE, echo=TRUE}
# Top-level fields of the ReportingEvent
names(re_json)
```

```{r ch2-preview-counts, exercise=FALSE, echo=TRUE}
# How many of each object type are in this sample?
cat("ReferenceDocuments:  ", length(re_json$referenceDocuments),  "\n")
cat("AnalysisSets:        ", length(re_json$analysisSets),        "\n")
cat("AnalysisGroupings:   ", length(re_json$analysisGroupings),   "\n")
cat("Methods:             ", length(re_json$methods),             "\n")
cat("Analyses:            ", length(re_json$analyses),            "\n")
```

One more navigation rule before you start: fields that hold *many* objects (like
`analysisSets` or `analyses`) parse as unnamed lists, so you index into them with
`[[1]]`, `[[2]]`, and so on. `re_json$analysisSets[[1]]` is the first AnalysisSet object,
and from there `$` works again: `re_json$analysisSets[[1]]$id`.

---

## Exercise 2.1: Parse and Explore the ARM-TS {data-progressive=FALSE}

**Step 1: the annotated shell.** The shell package header you write today:

> *Study: CDISC Pilot Study. Deliverable: CSR. Contains: populations, methods, analyses.*

**Step 2: the ARM-TS encoding.** That header is the top level of the ReportingEvent:

```json
{
  "id": "RE001",
  "name": "CDISC Pilot Study CSR",
  "analysisSets":      [ "..." ],
  "analysisGroupings": [ "..." ],
  "methods":           [ "..." ],
  "analyses":          [ "..." ]
}
```

**Step 3: your turn.** Walk the object hierarchy. The structure mirrors the diagram in
Section 2.3; every named field you see on the JSON is an object class in the model.

```{r ch2-ex1, exercise=TRUE, exercise.lines=12}
# re_json is the parsed ARM-TS, i.e. the root ReportingEvent

# 1. Print the reporting event name
re_json$___

# 2. How many analyses are defined?
length(re_json$___)

# 3. What is the ID of the first analysis set?
re_json$analysisSets[[1]]$___
```

```{r ch2-ex1-hint-1}
# Top-level fields are: id, version, name, referenceDocuments,
# analysisSets, analysisGroupings, methods, analyses
# Inside an AnalysisSet you'll find: id, name, condition
```

```{r ch2-ex1-solution}
re_json$name
length(re_json$analyses)
re_json$analysisSets[[1]]$id
# "CDISC Pilot Study CSR"
# 1 analysis in this minimal example
# "AS001", the ITT Population
```

---

## Exercise 2.2: WhereClause to dplyr Filter {data-progressive=FALSE}

This is the key bridge exercise of the chapter.

**Step 1: the annotated shell.** The population note on every safety table:

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; text-align:center;">Table 14.3.1: Summary of Adverse Events</div>
<div style="text-align:center; font-family:sans-serif; margin-bottom:8px;">Safety Population &nbsp;(<span style="color:#b3400e; font-weight:bold;">ADSL.SAFFL='Y'</span>)</div>
<table style="border-collapse:collapse; width:100%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left;"></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Treatment A</th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Treatment B</th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px; color:#888; font-style:italic;" colspan="3">[analyses...]</td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">Filter applied: <span style="color:#b3400e;">ADSL.SAFFL EQ 'Y'</span> — maps to ARS AnalysisSet AS002</div>
</div>

**Step 2: the ARM-TS encoding.** That note is the AnalysisSet `AS002`:

```json
{
  "id": "AS002",
  "name": "Safety Population",
  "condition": {
    "dataset": "ADSL",
    "variable": "SAFFL",
    "comparator": "EQ",
    "value": ["Y"]
  }
}
```

**Step 3: your turn.** Translate the condition into a dplyr filter and count the subjects.
Per the Section 2.4 mapping table, `EQ` becomes `==`. The rest is reading the condition
fields: variable on the left, value on the right.

```{r ch2-ex2, exercise=TRUE, exercise.lines=10}
# Translate the WhereClause into a dplyr filter on adsl,
# then count how many subjects are in the Safety Population

adsl %>%
  filter(___ == ___) %>%   # apply the WhereClause condition
  nrow()
```

```{r ch2-ex2-hint-1}
# The condition reads: ADSL.SAFFL EQ 'Y'
# In dplyr: filter(SAFFL == "Y")
```

```{r ch2-ex2-solution}
adsl %>%
  filter(SAFFL == "Y") %>%
  nrow()
# This is exactly what arsbridge::parse_ars() will do automatically:
# extract the WhereClause and generate this filter expression for you.
```

```{r ch2-ex2-check}
grade_result(
  pass_if(~ .result == nrow(dplyr::filter(adsl, SAFFL == "Y")),
          "Correct. That's the Safety Population. A WhereClause maps mechanically to a dplyr filter; this is the core of arsbridge::parse_ars()."),
  fail_if(~ TRUE,
          "Try: adsl %>% filter(SAFFL == 'Y') %>% nrow()")
)
```

---

## Exercise 2.3: Read a Compound WhereClause {data-progressive=FALSE}

Real WhereClauses are often compound, joined by `AND`, `OR`, or `NOT` (Section 2.4).

**Step 1: the annotated shell.** A two-part population note:

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; text-align:center;">Table 14.X.X: Efficacy Summary</div>
<div style="text-align:center; font-family:sans-serif; margin-bottom:6px;">Adult Intent-to-Treat Population (<span style="color:#b3400e; font-weight:bold;">ADSL.ITTFL='Y' AND ADSL.AGE&gt;=18</span>)</div>
<table style="border-collapse:collapse; width:100%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left;"></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Treatment A</th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Placebo</th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px; color:#888; font-style:italic;" colspan="3">[analyses...]</td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">Compound filter: <span style="color:#b3400e;">ADSL.ITTFL EQ 'Y'</span> AND <span style="color:#b3400e;">ADSL.AGE GE 18</span> — one compoundExpression, logicalOperator = "AND"</div>
</div>

**Step 2: the ARM-TS encoding.** Two conditions under one logical operator:

```json
{
  "compoundExpression": {
    "logicalOperator": "AND",
    "whereClauses": [
      { "condition": { "dataset": "ADSL", "variable": "ITTFL",
                       "comparator": "EQ", "value": ["Y"] } },
      { "condition": { "dataset": "ADSL", "variable": "AGE",
                       "comparator": "GE", "value": ["18"] } }
    ]
  }
}
```

**Step 3: your turn.** An `AND` of two conditions means a row must satisfy both; in dplyr
that is `&` inside one `filter()` call, or equivalently two chained `filter()` calls.
Note the second comparator: `GE` maps to `>=`, not `==`.

```{r ch2-ex3, exercise=TRUE, exercise.lines=10}
# Filter adsl to subjects who are BOTH in the ITT population AND aged 18+
# Use `&` (or two filter() calls) to combine conditions.

adsl %>%
  filter(___ == ___ & ___ >= ___) %>%
  nrow()
```

```{r ch2-ex3-hint-1}
# Two conditions, joined by AND:
#   ITTFL == "Y"
#   AGE   >= 18
```

```{r ch2-ex3-solution}
adsl %>%
  filter(ITTFL == "Y" & AGE >= 18) %>%
  nrow()
# In ARM-TS compoundExpression form, this is:
#   logicalOperator = "AND"
#   whereClauses[1] = ADSL.ITTFL EQ 'Y'
#   whereClauses[2] = ADSL.AGE   GE  18
```

---

## Exercise 2.4: Identify the Object Types {data-progressive=FALSE}

Now flip direction. Instead of translating ARS to R, take phrases straight out of an
annotated shell and name the ARS object each one becomes. This is how you will think when
reading a shell and translating it into ARM-TS: every annotation decomposes into one of
six object types you met in Section 2.3.

The scaffold below uses `tribble()`, a tidyverse helper that builds a small data frame
row by row (column names start with `~`, then each line is one row). You only need to fill
in the right-hand values.

```{r ch2-ex4, exercise=TRUE, exercise.lines=20}
# Each row below is something you might find in an annotated shell.
# Build a small lookup mapping each phrase to its ARS object class.

mappings <- tribble(
  ~annotated_shell_phrase,                       ~ars_object,
  "Population: Safety (SAFFL='Y')",              ___,
  "Visit: Week 24 only (AVISITN == 24)",         ___,
  "By treatment arm",                            ___,
  "n, mean, SD, median, min, max",               ___,
  "AVAL from ADTTE",                             ___,
  "Per SAP Section 9.2.3",                       ___
)

mappings
```

```{r ch2-ex4-hint-1}
# Choose from:
#   "AnalysisSet"        : subject-level population filter
#   "DataSubset"         : data-record-level filter
#   "AnalysisGrouping"   : factor splitting results into groups
#   "AnalysisMethod"     : set of statistical operations
#   "Analysis"           : the dataset + variable + method binding
#   "DocumentReference"  : link to external document
```

```{r ch2-ex4-solution}
mappings <- tribble(
  ~annotated_shell_phrase,                       ~ars_object,
  "Population: Safety (SAFFL='Y')",              "AnalysisSet",
  "Visit: Week 24 only (AVISITN == 24)",         "DataSubset",
  "By treatment arm",                            "AnalysisGrouping",
  "n, mean, SD, median, min, max",               "AnalysisMethod",
  "AVAL from ADTTE",                             "Analysis",
  "Per SAP Section 9.2.3",                       "DocumentReference"
)
mappings
# This six-way mapping is the whole game.
# Every annotated shell row decomposes into these six object types.
```

---

## Exercise 2.5: Walk an Analysis Object {data-progressive=FALSE}

**Step 1: the annotated shell.** One shell row, as your lead writes it:

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; text-align:center;">Table 14.1.1: Demographic Characteristics</div>
<div style="text-align:center; font-family:sans-serif; margin-bottom:8px;">Safety Population &nbsp;(<span style="color:#b3400e; font-weight:bold;">ADSL.SAFFL='Y'</span>)</div>
<table style="border-collapse:collapse; width:100%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left; width:42%;"></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Xanomeline High Dose<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Xanomeline Low Dose<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Placebo<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX</span></th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">Age (years) &nbsp;<span style="color:#b3400e; font-weight:bold;">ADSL.AGE</span></td>
  <td style="border:1px solid #bbb; padding:3px 10px;"></td>
  <td style="border:1px solid #bbb; padding:3px 10px;"></td>
  <td style="border:1px solid #bbb; padding:3px 10px;"></td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;n</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx</td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;Mean (SD)</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x (xxx.xx)</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x (xxx.xx)</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x (xxx.xx)</td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;Median</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x</td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;Min, Max</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx, xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx, xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx, xxx</td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">Grouping: <span style="color:#b3400e;">TRT01A</span> &nbsp;|&nbsp; Method: n / Mean / SD / Median / Min / Max &nbsp;|&nbsp; SAP §8.1 (PRIMARY)</div>
</div>

**Step 2: the ARM-TS encoding.** That row is one `Analysis` object; note that population
and method appear only as ID references (Principle 1):

```json
{
  "id": "AN001",
  "dataset": "ADSL",
  "variable": "AGE",
  "analysisSetId": "AS002",
  "methodId": "M001",
  "reason":  { "controlledTerm": "SPECIFIED IN SAP" },
  "purpose": { "controlledTerm": "PRIMARY OUTCOME MEASURE" }
}
```

**Step 3: your turn.** The first (and only) `Analysis` in `re_json` is exactly this
object. Pull out its key fields to confirm the model matches the diagram in Section 2.3.

Two things to remember from earlier sections:

- Most fields (`dataset`, `variable`, `analysisSetId`, `methodId`) are plain strings, so
  `an$dataset` works directly.
- `reason` and `purpose` are controlled-terminology objects (Section 2.5), so the value
  lives one level down: `an$reason$controlledTerm`.

```{r ch2-ex5, exercise=TRUE, exercise.lines=15}
# Access the first Analysis object
an <- re_json$analyses[[1]]

# Print these fields:
#   - dataset
#   - variable
#   - analysisSetId      (which population?)
#   - methodId           (which method?)
#   - reason$controlledTerm
#   - purpose$controlledTerm

cat("dataset:  ",       an$___,                       "\n")
cat("variable: ",       an$___,                       "\n")
cat("analysisSetId: ",  an$___,                       "\n")
cat("methodId: ",       an$___,                       "\n")
cat("reason:   ",       an$reason$___,                "\n")
cat("purpose:  ",       an$purpose$___,               "\n")
```

```{r ch2-ex5-solution}
an <- re_json$analyses[[1]]
cat("dataset:  ",       an$dataset,                "\n")
cat("variable: ",       an$variable,               "\n")
cat("analysisSetId: ",  an$analysisSetId,          "\n")
cat("methodId: ",       an$methodId,               "\n")
cat("reason:   ",       an$reason$controlledTerm,  "\n")
cat("purpose:  ",       an$purpose$controlledTerm, "\n")
# dataset:  ADSL
# variable: AGE
# analysisSetId: AS002    (Safety Population)
# methodId: M001          (n, mean, SD, median, min, max)
# reason:  SPECIFIED IN SAP
# purpose: PRIMARY OUTCOME MEASURE
```

---

## Exercise 2.6: Read an OperationResult (a peek at the ARD) {data-progressive=FALSE}

**Step 1: the annotated shell.** One filled-in table cell:

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; text-align:center;">Table 14.1.1: Demographic Characteristics [excerpt]</div>
<div style="text-align:center; font-family:sans-serif; margin-bottom:8px;">Safety Population &nbsp;(<span style="color:#b3400e; font-weight:bold;">ADSL.SAFFL='Y'</span>)</div>
<table style="border-collapse:collapse; width:100%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left; width:44%;"></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Xanomeline High Dose &nbsp;<span style="color:#b3400e; font-size:0.85em; font-weight:normal;">ADSL.TRT01A='Xanomeline High Dose'</span></th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">Mean Age &nbsp;<span style="color:#b3400e; font-weight:bold;">ADSL.AGE</span></td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;"><strong>74.4</strong></td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">Format: <span style="color:#b3400e;">XXX.X</span> &nbsp;|&nbsp; rawValue: 74.38095 &nbsp;&rarr;&nbsp; formattedValue: 74.4</div>
</div>

**Step 2: the ARM-TS encoding.** That one cell is one `OperationResult` inside
`Analysis.results`; every field except the two values is an ID reference back to the spec:

```json
{
  "operationId": "O002",
  "resultGroups": [ { "groupingId": "GF001", "groupId": "G001" } ],
  "rawValue": "74.38095",
  "formattedValue": "74.4"
}
```

### Concept first: the anatomy of one result

Inside `Analysis.results` lives the ARD. Each entry is one `OperationResult`, and you
already know its four key fields from the hierarchy diagram:

- `operationId`: which statistical operation produced this value (here, `O002` is the Mean)
- `resultGroups`: which group combination the value belongs to. This is a *list*, because
  a result can sit at the intersection of several groupings (arm by sex, arm by visit).
  Each entry holds a `groupId`, so the first one is `res$resultGroups[[1]]$groupId`.
- `rawValue`: the number at full precision, exactly as computed
- `formattedValue`: the display-ready string, formatted according to the operation's
  `resultPattern` (a pattern like `"XXX.X"` means one decimal place)

Why carry both values? Because they serve different audiences. `rawValue` lets a QC
programmer or a regulator re-compute and compare without rounding noise. `formattedValue`
is what lands in the table. A static PDF gives you only the second one; the ARD preserves
both. This pair is arguably the single most important innovation of the ARD over a
formatted table.

**Step 3: your turn.** Pull the mean age for the Xanomeline High Dose arm out of the
sample; the JSON above is literally `results[[2]]` of the analysis you walked in
Exercise 2.5.

```{r ch2-ex6, exercise=TRUE, exercise.lines=15}
# results is a list of OperationResult objects.
# In this sample, results[[2]] is the Mean for Group G001 (Xanomeline High Dose).

res <- re_json$analyses[[1]]$results[[2]]

cat("operationId:    ", res$___,                              "\n")  # O002 = Mean
cat("groupId:        ", res$resultGroups[[1]]$___,            "\n")  # which group?
cat("rawValue:       ", res$___,                              "\n")  # unformatted
cat("formattedValue: ", res$___,                              "\n")  # display-ready
```

```{r ch2-ex6-solution}
res <- re_json$analyses[[1]]$results[[2]]
cat("operationId:    ", res$operationId,                    "\n")
cat("groupId:        ", res$resultGroups[[1]]$groupId,      "\n")
cat("rawValue:       ", res$rawValue,                       "\n")
cat("formattedValue: ", res$formattedValue,                 "\n")
# operationId:    O002        (Mean)
# groupId:        G001        (Xanomeline High Dose)
# rawValue:       74.38095    (full precision)
# formattedValue: 74.4        (per the operation's resultPattern "XXX.X")
```

---

## Chapter 2 Knowledge Check

Everything below comes from Sections 2.1 through 2.6 and the exercises. If a question
stumps you: the hierarchy and the AnalysisSet vs DataSubset split are in 2.3, WhereClause
operators and comparators in 2.4, the enumerations in 2.5, the design principles in 2.6,
and the rawValue/formattedValue pair in Exercise 2.6.

```{r ch2-quiz, echo=FALSE}
quiz(
  caption = "Chapter 2: Check Your Understanding",

  question(
    "What is the root object in the ARS data model?",
    answer("Analysis"),
    answer("ReportingEvent", correct = TRUE),
    answer("AnalysisMethod"),
    answer("Output"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "What is the difference between an AnalysisSet and a DataSubset?",
    answer(
      "AnalysisSet filters subjects (e.g., SAFFL='Y' in ADSL); DataSubset filters data records (e.g., PARAMCD='OS' in ADTTE)",
      correct = TRUE
    ),
    answer("AnalysisSet is for ADaM; DataSubset is for SDTM"),
    answer("They are the same; both filter rows from any dataset"),
    answer("AnalysisSet applies only to ADSL; DataSubset applies only to BDS datasets"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "In ARM-TS, which logical operators are available for compound WhereClause expressions?",
    answer("AND, OR only"),
    answer("AND, OR, NOT", correct = TRUE),
    answer("AND, OR, NOT, XOR"),
    answer("EQ, NE, IN, NOTIN"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which ARS comparator corresponds to dplyr's `%in%`?",
    answer("EQ"),
    answer("IN", correct = TRUE),
    answer("GE"),
    answer("NE"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Where does the ARD physically live in the ARS data model?",
    answer("As a separate top-level field of ReportingEvent"),
    answer("Inside Analysis.results as a list of OperationResult objects", correct = TRUE),
    answer("In a dedicated ARD file referenced by Output.fileSpecifications"),
    answer("Inside AnalysisMethod.operations"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "What does Principle 1, 'Reuse over repetition', imply for the data model?",
    answer("Methods, AnalysisSets, and Groups are defined once and referenced by ID from many Analyses", correct = TRUE),
    answer("Every Analysis must define its own private Method"),
    answer("Each ReportingEvent must contain only one AnalysisSet"),
    answer("Results from one analysis can be reused as input to another"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which of the following is NOT one of the four extensible CDISC enumerations in ARS?",
    answer("AnalysisReasonEnum"),
    answer("AnalysisPurposeEnum"),
    answer("OperationRoleEnum"),
    answer("StatisticalMethodEnum", correct = TRUE),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Why does each OperationResult have BOTH a rawValue and a formattedValue?",
    answer(
      "rawValue preserves full numeric precision for re-computation/QC; formattedValue is the display-ready string per the operation's resultPattern",
      correct = TRUE
    ),
    answer("rawValue is the SAS output; formattedValue is the R output"),
    answer("rawValue is used in ARM-TS; formattedValue is used in ARD"),
    answer("They are duplicates kept for backward compatibility"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

---

## Chapter 2 Summary {.congratulations}

You now know:

- The ARS model has a single root: **`ReportingEvent`**
- ARM-TS objects define the specification: **AnalysisSets, DataSubsets, Groupings, Methods,
  Analyses, Outputs**
- **AnalysisSet filters subjects; DataSubset filters data records.** Keep that split sharp.
- The ARD lives **inside `Analysis.results`** as a list of `OperationResult` objects, not
  as a separate top-level structure
- **`WhereClause`** is the universal filter mechanism (simple condition, compound
  expression, or reference) and maps mechanically to a dplyr `filter()`
- Objects are defined once and referenced by ID; **reuse is a core design principle**
- Four extensible controlled terminologies cover **reason, purpose, operation role, and
  file type**
- An `OperationResult` carries both `rawValue` (full precision) and `formattedValue`
  (display string)

**Next:** Chapter 3, the ARM-TS Deep Dive: every object with examples and field-by-field
annotated-shell mappings
