---
title: "CDISC Analysis Results Standard (ARS) — Interactive Training"
subtitle: "Chapter 2: The ARS Data Model"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Chapter 2 walks through the complete ARS object hierarchy: ReportingEvent,
  AnalysisSets, DataSubsets, AnalysisGroupings, Methods, Analyses, Outputs,
  and the OperationResult (ARD). You will explore a real ARM-TS JSON,
  translate WhereClauses into dplyr filters, and build object structures
  by hand.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(jsonlite)
library(glue)

# learnr global options
tutorial_options(
  exercise.timelimit = 60
)

# gradethis setup
gradethis_setup()

# Load pharmaverse ADaM datasets
data("adsl",  package = "pharmaverseadam")
data("adae",  package = "pharmaverseadam")
data("adtte", package = "pharmaverseadam")
data("adlb",  package = "pharmaverseadam")

# Pre-filter datasets used throughout the tutorial
adsl_safe <- adsl %>% filter(SAFFL == "Y")
adsl_itt  <- adsl %>% filter(!is.na(ARM) & ARM != "")
adae_teae <- adae %>% filter(TRTEMFL == "Y")

# Load sample ARM-TS JSON
re_json <- jsonlite::read_json("data/reporting_event.json")
```

---

## Chapter 2: The ARS Data Model

### 2.1 Overview — Why a Logical Model?

The ARS data model is a **logical model** — it defines objects and their relationships,
independent of any specific file format. In practice, ARS is serialized as JSON (or YAML),
but the model itself is format-agnostic. You could store an ARM-TS in a relational database,
in YAML, or in protobuf, and it would still be the same logical structure.

The model has a single root object: the `ReportingEvent`. Everything else hangs off it.

Think of the `ReportingEvent` as one CSR-worth of work, or one interim analysis, or one
regulatory deliverable. Inside it lives every population, every method, every analysis,
and every output associated with that deliverable.

---

### 2.2 The Two Layers Revisited

Before diving into objects, anchor yourself in the two layers introduced in Chapter 1:

```
┌─────────────────────────────────────────────────────────┐
│                     ARM-TS Layer                        │
│  (What was planned and how analyses were performed)     │
│                                                         │
│  ReportingEvent                                         │
│  ├── AnalysisSets      (population filters / flags)     │
│  ├── DataSubsets       (data record filters)            │
│  ├── AnalysisGroupings (treatment, sex, visit...)       │
│  ├── Methods           (statistical operations)         │
│  ├── Analyses          (dataset + variable + method)    │
│  └── Outputs           (displays / tables)              │
└─────────────────────────────────────────────────────────┘

┌─────────────────────────────────────────────────────────┐
│                      ARD Layer                          │
│  (What was actually computed)                           │
│                                                         │
│  Lives inside Analysis.results                          │
│  OperationResult: operationId + resultGroups +          │
│                   rawValue + formattedValue             │
└─────────────────────────────────────────────────────────┘
```

The ARD is **not** a separate file in the formal model — it lives as the `results` slot
inside each `Analysis` object. In practice, it is often extracted and stored as a separate
flat dataset (a data frame) for use with tools like `{cards}`. Chapter 4 covers this in depth.

---

### 2.3 The Complete Object Hierarchy

Here is every class in the ARS model, organized by function. Don't try to memorize this —
just orient yourself.

**Root**

```
ReportingEvent
```

The top-level container. Everything for one reporting requirement (e.g., one CSR, one interim
analysis) lives inside or is referenced from a `ReportingEvent`.

**Structural / Organizational Objects**

```
ReportingEvent
├── ListOfContents          — structured table of contents for analyses and outputs
│   └── NestedList
│       └── OrderedListItem — points to an Analysis or Output, with level/order
│
├── ReferenceDocument       — external file (SAP PDF, program file, etc.)
│
├── TerminologyExtension    — sponsor-defined terms extending CDISC enumerations
│   └── SponsorTerm
│
└── AnalysisOutputCategorization  — implementer-defined category system
    └── AnalysisOutputCategory    — a single category (e.g., "Safety", "Efficacy")
```

**Population and Data Selection Objects**

These define **who** and **what data** is included in analyses. This maps directly to WHERE
clauses in your annotated shells.

```
ReportingEvent
├── AnalysisSets            — subject populations (ITT, PP, Safety, etc.)
│   └── WhereClause         — the filter logic (e.g., ADSL.ITTFL EQ 'Y')
│
├── DataSubsets             — data record filters (e.g., a specific visit, PARAMCD)
│   └── WhereClause
│
└── AnalysisGroupings       — factors for subdividing (treatment arm, sex, visit)
    └── GroupingFactor
        └── Group           — one level within the factor (e.g., "Treatment A")
            └── WhereClause
```

**Method Objects**

These define **how** analyses are performed. One method can be reused across many analyses —
this is a core design principle (Principle 1 below).

```
ReportingEvent
└── Methods
    └── AnalysisMethod       — a named set of statistical operations
        ├── Operations        — one or more statistical operations
        │   └── Operation     —  each produces one result
        │       └── ReferencedOperationRelationship  — e.g., NUMERATOR / DENOMINATOR
        │
        └── AnalysisProgrammingCodeTemplate  — reusable code template with parameters
            └── TemplateCodeParameter
```

**Analysis Objects**

Each `Analysis` is one specific analysis in the study. It ties together: a dataset, a variable,
a population, optional data filters, optional groupings, and a method.

```
ReportingEvent
└── Analyses
    └── Analysis
        ├── dataset          — e.g., "ADTTE"
        ├── variable         — e.g., "AVAL"
        ├── analysisSetId    — references an AnalysisSet (population)
        ├── dataSubsetId     — references a DataSubset (optional)
        ├── orderedGroupings — references GroupingFactors (treatment, visit, etc.)
        ├── methodId         — references an AnalysisMethod
        ├── reason           — SPECIFIED IN PROTOCOL / SPECIFIED IN SAP / DATA DRIVEN
        ├── purpose          — PRIMARY / SECONDARY / EXPLORATORY OUTCOME MEASURE
        ├── documentRefs     — links to SAP, program files
        ├── programmingCode  — actual R/SAS code or reference to file
        └── results          — the ARD (OperationResults)
            └── OperationResult
                ├── operationId     — which operation this result is for
                ├── resultGroups    — which group combination
                ├── rawValue        — unformatted result
                └── formattedValue  — result formatted per resultPattern
```

**Output / Display Objects**

These define the **display** — the table itself, its title, footnotes, and how analyses map
into it.

```
ReportingEvent
└── Outputs
    └── Output
        ├── fileSpecifications  — output file format (PDF, RTF, TXT)
        ├── programmingCode     — code to generate the display
        └── displays            — ordered list of displays in the output
            └── OrderedDisplay
                └── OutputDisplay
                    ├── displayTitle      — the table title
                    └── displaySections   — title, header, footnote, abbreviation, footer
                        └── DisplaySection
                            └── OrderedDisplaySubSection
                                └── DisplaySubSection (text)
```

---

### 2.4 The WhereClause System

The `WhereClause` is the most important structural pattern in ARS because it appears in **three
different contexts**: `AnalysisSet`, `DataSubset`, and `Group`. Master this and you understand
60% of the model.

**Simple Condition**

```json
{
  "condition": {
    "dataset": "ADSL",
    "variable": "ITTFL",
    "comparator": "EQ",
    "value": ["Y"]
  }
}
```

This directly maps to: `WHERE ADSL.ITTFL = 'Y'` in SAS, or `filter(ITTFL == "Y")` in R.

**Compound Expression**

```json
{
  "compoundExpression": {
    "logicalOperator": "AND",
    "whereClauses": [
      {
        "condition": {
          "dataset": "ADSL",
          "variable": "ITTFL",
          "comparator": "EQ",
          "value": ["Y"]
        }
      },
      {
        "condition": {
          "dataset": "ADSL",
          "variable": "AGE",
          "comparator": "GE",
          "value": ["18"]
        }
      }
    ]
  }
}
```

- **Logical operators:** `AND`, `OR`, `NOT`
- **Comparators:** `EQ`, `NE`, `GT`, `GE`, `LT`, `LE`, `IN`, `NOTIN`

**Referenced WhereClause**

A `WhereClause` can reference another already-defined clause by its ID (e.g., reusing an
`AnalysisSet` as part of a `Group` definition). This avoids repeating complex logic.

---

### 2.5 The Controlled Terminology

ARS uses **four extensible enumerations**. These are the only controlled terms in the model:

| Enumeration | Values | What it encodes |
|-------------|--------|-----------------|
| `AnalysisReasonEnum` | SPECIFIED IN PROTOCOL, SPECIFIED IN SAP, DATA DRIVEN, REQUESTED BY REGULATORY AGENCY | When/why the analysis was planned |
| `AnalysisPurposeEnum` | PRIMARY OUTCOME MEASURE, SECONDARY OUTCOME MEASURE, EXPLORATORY OUTCOME MEASURE | Role of the analysis in the study |
| `OperationRoleEnum` | NUMERATOR, DENOMINATOR | Relationship between operations (for % calculations) |
| `OutputFileTypeEnum` | pdf, rtf, txt | File format of output |

All four are **extensible** — sponsors can add their own terms via `TerminologyExtension`.

---

### 2.6 Key Design Principles

**Principle 1: Reuse over repetition**

Methods, AnalysisSets, DataSubsets, and Groups are defined once and referenced by ID from
multiple Analyses. If 50 analyses all use the ITT population, there is one `AnalysisSet` with
`id = "AS001"` and all 50 analyses reference `analysisSetId = "AS001"`. This is how the model
enforces consistency.

**Principle 2: Separation of specification from results**

The specification (what to do) and the results (what was computed) live in the same `Analysis`
object. The `results` slot contains `OperationResult` objects — this is the ARD. The full
context of a result is always navigable from the result itself.

**Principle 3: Ordered everything**

Lists in ARS are ordered explicitly using `level` and `order` integer fields. The
`ListOfContents` uses this to define the table of contents. `OrderedGroupings` uses it to
define the sequence of grouping factors. This is important for display rendering.

**Principle 4: Extensibility**

Where CDISC controlled terms don't cover a sponsor's needs, the `TerminologyExtension` mechanism
allows adding custom terms while remaining conformant with the standard.

---

### 2.7 Entity-Relationship Summary

```
ReportingEvent (1)
    |
    ├─── (many) AnalysisSet          ← referenced by Analysis.analysisSetId
    ├─── (many) DataSubset           ← referenced by Analysis.dataSubsetId
    ├─── (many) GroupingFactor       ← referenced by OrderedGroupingFactor.groupingId
    │            └── (many) Group    ← referenced by ResultGroup.groupId
    ├─── (many) AnalysisMethod       ← referenced by Analysis.methodId
    │            └── (many) Operation ← referenced by OperationResult.operationId
    ├─── (many) Analysis             ← referenced by OrderedListItem.analysisId
    │            └── (many) OperationResult (ARD lives here)
    ├─── (many) Output               ← referenced by OrderedListItem.outputId
    │            └── (many) OutputDisplay
    └─── (many) ReferenceDocument    ← referenced by DocumentReference.referenceDocumentId
```

---

## Hands-On: A Real ARM-TS in Your Hands

### Orient to the sample ARM-TS

The sample ARM-TS JSON (`reporting_event.json`) is already loaded as `re_json`. Take a look
at the top-level structure before the exercises.

```{r ch2-preview, exercise=FALSE, echo=TRUE}
# Top-level fields of the ReportingEvent
names(re_json)
```

```{r ch2-preview-counts, exercise=FALSE, echo=TRUE}
# How many of each object type are in this sample?
cat("ReferenceDocuments:  ", length(re_json$referenceDocuments),  "\n")
cat("AnalysisSets:        ", length(re_json$analysisSets),        "\n")
cat("AnalysisGroupings:   ", length(re_json$analysisGroupings),   "\n")
cat("Methods:             ", length(re_json$methods),             "\n")
cat("Analyses:            ", length(re_json$analyses),            "\n")
```

---

## Exercise 2.1 — Parse and Explore the ARM-TS {data-progressive=FALSE}

Walk the object hierarchy. The structure mirrors the diagram in Section 2.3 — every named
field you see on the JSON is an object class in the model.

```{r ch2-ex1, exercise=TRUE, exercise.lines=12}
# re_json is the parsed ARM-TS — the root ReportingEvent

# 1. Print the reporting event name
re_json$___

# 2. How many analyses are defined?
length(re_json$___)

# 3. What is the ID of the first analysis set?
re_json$analysisSets[[1]]$___
```

```{r ch2-ex1-hint-1}
# Top-level fields are: id, version, name, referenceDocuments,
# analysisSets, analysisGroupings, methods, analyses
# Inside an AnalysisSet you'll find: id, name, condition
```

```{r ch2-ex1-solution}
re_json$name
length(re_json$analyses)
re_json$analysisSets[[1]]$id
# "CDISC Pilot Study CSR"
# 1 analysis in this minimal example
# "AS001" — the ITT Population
```

---

## Exercise 2.2 — WhereClause to dplyr Filter {data-progressive=FALSE}

This is the key bridge exercise. The ARM-TS Safety Population AnalysisSet (`AS002`) has the
condition below. **Translate it into a dplyr filter and count the subjects.**

> dataset: `"ADSL"`
> variable: `"SAFFL"`
> comparator: `"EQ"`
> value: `["Y"]`

```{r ch2-ex2, exercise=TRUE, exercise.lines=10}
# Translate the WhereClause into a dplyr filter on adsl,
# then count how many subjects are in the Safety Population

adsl %>%
  filter(___ == ___) %>%   # apply the WhereClause condition
  nrow()
```

```{r ch2-ex2-hint-1}
# The condition reads: ADSL.SAFFL EQ 'Y'
# In dplyr: filter(SAFFL == "Y")
```

```{r ch2-ex2-solution}
adsl %>%
  filter(SAFFL == "Y") %>%
  nrow()
# This is exactly what arsbridge::parse_ars() will do automatically —
# extract the WhereClause and generate this filter expression for you.
```

```{r ch2-ex2-check}
grade_result(
  pass_if(~ .result == nrow(dplyr::filter(adsl, SAFFL == "Y")),
          "Correct — that's the Safety Population. A WhereClause maps mechanically to a dplyr filter; this is the core of arsbridge::parse_ars()."),
  fail_if(~ TRUE,
          "Try: adsl %>% filter(SAFFL == 'Y') %>% nrow()")
)
```

---

## Exercise 2.3 — Read a Compound WhereClause {data-progressive=FALSE}

Real WhereClauses are often compound — joined by `AND`, `OR`, or `NOT`. Translate this
compound expression:

> **AND** of:
> - `ADSL.ITTFL EQ 'Y'`
> - `ADSL.AGE GE 18`

```{r ch2-ex3, exercise=TRUE, exercise.lines=10}
# Filter adsl to subjects who are BOTH in the ITT population AND aged 18+
# Use `&` (or two filter() calls) to combine conditions.

adsl %>%
  filter(___ == ___ & ___ >= ___) %>%
  nrow()
```

```{r ch2-ex3-hint-1}
# Two conditions, joined by AND:
#   ITTFL == "Y"
#   AGE   >= 18
```

```{r ch2-ex3-solution}
adsl %>%
  filter(ITTFL == "Y" & AGE >= 18) %>%
  nrow()
# In ARM-TS compoundExpression form, this is:
#   logicalOperator = "AND"
#   whereClauses[1] = ADSL.ITTFL EQ 'Y'
#   whereClauses[2] = ADSL.AGE   GE  18
```

---

## Exercise 2.4 — Identify the Object Types {data-progressive=FALSE}

Given a scenario, match it to the correct ARS object. This is how you'll think when reading
an annotated shell and translating it into ARM-TS.

```{r ch2-ex4, exercise=TRUE, exercise.lines=20}
# Each row below is something you might find in an annotated shell.
# Build a small lookup mapping each phrase to its ARS object class.

mappings <- tribble(
  ~annotated_shell_phrase,                       ~ars_object,
  "Population: Safety (SAFFL='Y')",              ___,
  "Visit: Week 24 only (AVISITN == 24)",         ___,
  "By treatment arm",                            ___,
  "n, mean, SD, median, min, max",               ___,
  "AVAL from ADTTE",                             ___,
  "Per SAP Section 9.2.3",                       ___
)

mappings
```

```{r ch2-ex4-hint-1}
# Choose from:
#   "AnalysisSet"        — subject-level population filter
#   "DataSubset"         — data-record-level filter
#   "AnalysisGrouping"   — factor splitting results into groups
#   "AnalysisMethod"     — set of statistical operations
#   "Analysis"           — the dataset + variable + method binding
#   "DocumentReference"  — link to external document
```

```{r ch2-ex4-solution}
mappings <- tribble(
  ~annotated_shell_phrase,                       ~ars_object,
  "Population: Safety (SAFFL='Y')",              "AnalysisSet",
  "Visit: Week 24 only (AVISITN == 24)",         "DataSubset",
  "By treatment arm",                            "AnalysisGrouping",
  "n, mean, SD, median, min, max",               "AnalysisMethod",
  "AVAL from ADTTE",                             "Analysis",
  "Per SAP Section 9.2.3",                       "DocumentReference"
)
mappings
# This six-way mapping is the whole game.
# Every annotated shell row decomposes into these six object types.
```

---

## Exercise 2.5 — Walk an Analysis Object {data-progressive=FALSE}

The first (and only) `Analysis` in `re_json` is the Age Summary by Treatment Arm. Pull out
its key fields to confirm the model.

```{r ch2-ex5, exercise=TRUE, exercise.lines=15}
# Access the first Analysis object
an <- re_json$analyses[[1]]

# Print these fields:
#   - dataset
#   - variable
#   - analysisSetId      (which population?)
#   - methodId           (which method?)
#   - reason$controlledTerm
#   - purpose$controlledTerm

cat("dataset:  ",       an$___,                       "\n")
cat("variable: ",       an$___,                       "\n")
cat("analysisSetId: ",  an$___,                       "\n")
cat("methodId: ",       an$___,                       "\n")
cat("reason:   ",       an$reason$___,                "\n")
cat("purpose:  ",       an$purpose$___,               "\n")
```

```{r ch2-ex5-solution}
an <- re_json$analyses[[1]]
cat("dataset:  ",       an$dataset,                "\n")
cat("variable: ",       an$variable,               "\n")
cat("analysisSetId: ",  an$analysisSetId,          "\n")
cat("methodId: ",       an$methodId,               "\n")
cat("reason:   ",       an$reason$controlledTerm,  "\n")
cat("purpose:  ",       an$purpose$controlledTerm, "\n")
# dataset:  ADSL
# variable: AGE
# analysisSetId: AS002    (Safety Population)
# methodId: M001          (n, mean, SD, median, min, max)
# reason:  SPECIFIED IN SAP
# purpose: PRIMARY OUTCOME MEASURE
```

---

## Exercise 2.6 — Read an OperationResult (a peek at the ARD) {data-progressive=FALSE}

Inside `Analysis.results` lives the ARD. Each row is one `OperationResult`. Pull the mean
age for Xanomeline High Dose out of the sample.

```{r ch2-ex6, exercise=TRUE, exercise.lines=15}
# results is a list of OperationResult objects.
# In this sample, results[[2]] is the Mean for Group G001 (Xanomeline High Dose).

res <- re_json$analyses[[1]]$results[[2]]

cat("operationId:    ", res$___,                              "\n")  # O002 = Mean
cat("groupId:        ", res$resultGroups[[1]]$___,            "\n")  # which group?
cat("rawValue:       ", res$___,                              "\n")  # unformatted
cat("formattedValue: ", res$___,                              "\n")  # display-ready
```

```{r ch2-ex6-solution}
res <- re_json$analyses[[1]]$results[[2]]
cat("operationId:    ", res$operationId,                    "\n")
cat("groupId:        ", res$resultGroups[[1]]$groupId,      "\n")
cat("rawValue:       ", res$rawValue,                       "\n")
cat("formattedValue: ", res$formattedValue,                 "\n")
# operationId:    O002        (Mean)
# groupId:        G001        (Xanomeline High Dose)
# rawValue:       74.38095    (full precision)
# formattedValue: 74.4        (per the operation's resultPattern "XXX.X")
# This pair (rawValue + formattedValue) is the single most important
# innovation of the ARD over a static PDF table — both precisions are preserved.
```

---

## Chapter 2 Knowledge Check

```{r ch2-quiz, echo=FALSE}
quiz(
  caption = "Chapter 2 — Check Your Understanding",

  question(
    "What is the root object in the ARS data model?",
    answer("Analysis"),
    answer("ReportingEvent", correct = TRUE),
    answer("AnalysisMethod"),
    answer("Output"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "What is the difference between an AnalysisSet and a DataSubset?",
    answer(
      "AnalysisSet filters subjects (e.g., SAFFL='Y' in ADSL); DataSubset filters data records (e.g., PARAMCD='OS' in ADTTE)",
      correct = TRUE
    ),
    answer("AnalysisSet is for ADaM; DataSubset is for SDTM"),
    answer("They are the same — both filter rows from any dataset"),
    answer("AnalysisSet applies only to ADSL; DataSubset applies only to BDS datasets"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "In ARM-TS, which logical operators are available for compound WhereClause expressions?",
    answer("AND, OR only"),
    answer("AND, OR, NOT", correct = TRUE),
    answer("AND, OR, NOT, XOR"),
    answer("EQ, NE, IN, NOTIN"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which ARS comparator corresponds to dplyr's `%in%`?",
    answer("EQ"),
    answer("IN", correct = TRUE),
    answer("GE"),
    answer("NE"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Where does the ARD physically live in the ARS data model?",
    answer("As a separate top-level field of ReportingEvent"),
    answer("Inside Analysis.results as a list of OperationResult objects", correct = TRUE),
    answer("In a dedicated ARD file referenced by Output.fileSpecifications"),
    answer("Inside AnalysisMethod.operations"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "What does Principle 1 — 'Reuse over repetition' — imply for the data model?",
    answer("Methods, AnalysisSets, and Groups are defined once and referenced by ID from many Analyses", correct = TRUE),
    answer("Every Analysis must define its own private Method"),
    answer("Each ReportingEvent must contain only one AnalysisSet"),
    answer("Results from one analysis can be reused as input to another"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which of the following is NOT one of the four extensible CDISC enumerations in ARS?",
    answer("AnalysisReasonEnum"),
    answer("AnalysisPurposeEnum"),
    answer("OperationRoleEnum"),
    answer("StatisticalMethodEnum", correct = TRUE),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Why does each OperationResult have BOTH a rawValue and a formattedValue?",
    answer(
      "rawValue preserves full numeric precision for re-computation/QC; formattedValue is the display-ready string per the operation's resultPattern",
      correct = TRUE
    ),
    answer("rawValue is the SAS output; formattedValue is the R output"),
    answer("rawValue is used in ARM-TS; formattedValue is used in ARD"),
    answer("They are duplicates kept for backward compatibility"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

---

## Chapter 2 Summary {.congratulations}

You now know:

- The ARS model has a single root: **`ReportingEvent`**
- ARM-TS objects define the specification: **AnalysisSets, DataSubsets, Groupings, Methods,
  Analyses, Outputs**
- ARD lives **inside `Analysis.results`** as a list of `OperationResult` objects — not as a
  separate top-level structure
- **`WhereClause`** is the universal filter mechanism (simple condition, compound expression,
  or reference) and maps mechanically to a dplyr `filter()`
- Objects are defined once and referenced by ID — **reuse is a core design principle**
- Four extensible controlled terminologies cover **reason, purpose, operation role, and file
  type**
- An `OperationResult` carries both `rawValue` (full precision) and `formattedValue` (display
  string)

**Next:** Chapter 3 — ARM-TS Deep Dive (every object with examples and field-by-field
annotated-shell mappings)
