---
title: "CDISC Analysis Results Standard (ARS) — Interactive Training"
subtitle: "Chapter 3: ARM-TS Deep Dive"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Chapter 3 walks through every major ARM-TS object with full JSON examples and
  the equivalent annotated-shell annotation. You will dissect the running case
  study — a safety AE summary by treatment arm — and build ARM-TS objects (an
  AnalysisMethod, an Analysis) by hand from a shell annotation.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(jsonlite)
library(glue)

tutorial_options(exercise.timelimit = 60)
gradethis_setup()

# Load pharmaverse ADaM datasets
data("adsl",  package = "pharmaverseadam")
data("adae",  package = "pharmaverseadam")
# pharmaverseadam 1.4+ replaced the generic "adtte" with TA-specific
# datasets; use the oncology one and join subject-level vars in from
# ADSL. pharmaverseadam ADSL lacks ITTFL, so derive it from ARM (a
# subject is "randomised / ITT" iff ARM is non-empty).
data("adtte_onco", package = "pharmaverseadam")
adtte <- adtte_onco %>%
  left_join(
    adsl %>%
      mutate(ITTFL = ifelse(!is.na(ARM) & ARM != "", "Y", "N")) %>%
      select(USUBJID, TRT01A, ITTFL),
    by = "USUBJID"
  )
data("adlb",  package = "pharmaverseadam")

adsl_safe <- adsl %>% filter(SAFFL == "Y")
adsl_itt  <- adsl %>% filter(!is.na(ARM) & ARM != "")
adae_teae <- adae %>% filter(TRTEMFL == "Y")

re_json <- jsonlite::read_json("data/reporting_event.json")
```

---

## Chapter 3: ARM-TS Deep Dive

This chapter walks through every major ARM-TS object with full JSON examples. Each section
includes the equivalent annotated-shell annotation so you can see exactly how your existing
knowledge maps to the formal model.

The examples use a running case study: **a safety table showing adverse event summary
statistics by treatment arm in the Safety Population** — a table type you have programmed many
times.

---

### 3.1 ReportingEvent

The **root container**. Represents one reporting deliverable — typically one CSR or one
interim analysis.

**Key fields**

| Field | Required | Description |
|-------|----------|-------------|
| `id`, `name` | Yes | Unique identifier and display name |
| `version` | No | Version number (increment on updates) |
| `mainListOfContents` | Yes | Structured table of contents |
| `referenceDocuments` | No | External documents (SAP, programs) |
| `analysisSets`, `dataSubsets`, `analysisGroupings`, `methods`, `analyses`, `outputs` | No | The object collections |

**Annotated-shell equivalent.** The annotated-shell header block — study number, protocol
title, SAP version, date. The `ReportingEvent` is the machine-readable version of that header.

---

### 3.2 ReferenceDocument

Links the reporting event to **external files**. This is how ARS creates a formal link to
the SAP.

```json
{
  "id": "SAP001",
  "name": "Statistical Analysis Plan v2.0",
  "location": "documents/XYZ001_SAP_v2.0.pdf",
  "description": "Final SAP approved 15-Jan-2025"
}
```

**Annotated-shell equivalent.** The *"per Section X.X of SAP"* text you write. In ARM-TS this
becomes a formal `DocumentReference` (with optional `pageRefs`) inside the `Analysis` object,
pointing to the `ReferenceDocument` you define here.

---

### 3.3 AnalysisSet

Defines a **subject population**. Your population flags in ADSL (`ITTFL`, `SAFFL`, `PPSFL`)
become `AnalysisSet` objects.

**Simple condition (single flag)**

```json
{
  "id": "AS002",
  "name": "Safety Population",
  "condition": {
    "dataset": "ADSL",
    "variable": "SAFFL",
    "comparator": "EQ",
    "value": ["Y"]
  }
}
```

**Compound expression (multiple flags)**

```json
{
  "id": "AS003",
  "name": "Per-Protocol Population",
  "compoundExpression": {
    "logicalOperator": "AND",
    "whereClauses": [
      { "condition": { "dataset": "ADSL", "variable": "PPSFL",   "comparator": "EQ", "value": ["Y"] } },
      { "condition": { "dataset": "ADSL", "variable": "COMPLFL", "comparator": "EQ", "value": ["Y"] } }
    ]
  }
}
```

Exactly one of `condition` or `compoundExpression` should be present.

**Annotated-shell equivalent.** *"SAFFL='Y' from ADSL"* or *"Safety Population"*. You define
it **once** as `AS002`, and every analysis referencing the Safety Population just sets
`"analysisSetId": "AS002"` — this is **Principle 1 (reuse)** in action.

---

### 3.4 DataSubset

A filter on **data records** (rows in an analysis dataset), not on subjects.

**Key distinction**

- `AnalysisSet` = *who* is in the analysis (subject-level, usually from ADSL)
- `DataSubset` = *which records* to use (row-level, on the analysis dataset itself)

```json
{
  "id": "DS001",
  "name": "Overall Survival Parameter",
  "condition": {
    "dataset": "ADTTE",
    "variable": "PARAMCD",
    "comparator": "EQ",
    "value": ["OS"]
  }
}
```

**Annotated-shell equivalent.** *"PARAMCD='OS'"* or *"AVISIT='Week 24'"*. The distinction
matters: your population flag selects **subjects**; your `PARAMCD` filter selects **records
within those subjects' data**.

---

### 3.5 GroupingFactor and Group

Defines how to subdivide the analysis. Treatment arm, sex, age group, visit — these are all
grouping factors.

**GroupingFactor — key fields**

| Field | Description |
|-------|-------------|
| `groupingDataset`, `groupingVariable` | Where the grouping variable lives |
| `dataDriven` | `true` = groups derived from data; `false` = groups pre-specified |
| `groups` | Pre-specified groups (when `dataDriven = false`) |

**Pre-specified (treatment arm)**

```json
{
  "id": "GF001",
  "name": "Treatment Arm",
  "groupingDataset": "ADSL",
  "groupingVariable": "TRT01A",
  "dataDriven": false,
  "groups": [
    { "id": "G001", "name": "Drug XYZ 10mg",
      "condition": { "dataset": "ADSL", "variable": "TRT01A", "comparator": "EQ", "value": ["Drug XYZ 10mg"] } },
    { "id": "G002", "name": "Placebo",
      "condition": { "dataset": "ADSL", "variable": "TRT01A", "comparator": "EQ", "value": ["Placebo"] } }
  ]
}
```

**Data-driven (visit)** — groups come from the data:

```json
{
  "id": "GF002",
  "name": "Analysis Visit",
  "groupingDataset": "ADLB",
  "groupingVariable": "AVISIT",
  "dataDriven": true
}
```

**Annotated-shell equivalent.** The **column headers** in your tables. A table with columns
"Drug XYZ 10mg" and "Placebo" reflects a `GroupingFactor` with two pre-specified `Groups`.
Each stored `OperationResult` carries a `resultGroup` linking it back to `G001` or `G002`.

---

### 3.6 AnalysisMethod

Defines one or more **statistical operations** that can be applied to any analysis variable.
Methods are reusable across analyses.

**Operations — key fields**

| Field | Description |
|-------|-------------|
| `id`, `name`, `order` | Identifier, name, position within the method |
| `resultPattern` | Display format (e.g., `"XX.X"`, `"XX.XX"`) |
| `referencedOperationRelationships` | How this operation uses results from another (e.g., NUMERATOR / DENOMINATOR for %) |

**Continuous summary**

```json
{
  "id": "M001",
  "name": "Summary Statistics - Continuous",
  "operations": [
    { "id": "O001", "name": "n",      "order": 1, "resultPattern": "XXX"    },
    { "id": "O002", "name": "Mean",   "order": 2, "resultPattern": "XXX.X"  },
    { "id": "O003", "name": "SD",     "order": 3, "resultPattern": "XXX.XX" },
    { "id": "O004", "name": "Median", "order": 4, "resultPattern": "XXX.X"  },
    { "id": "O005", "name": "Min",    "order": 5, "resultPattern": "XXX"    },
    { "id": "O006", "name": "Max",    "order": 6, "resultPattern": "XXX"    }
  ]
}
```

**Count + percentage (uses ReferencedOperationRelationship)**

```json
{
  "id": "M002",
  "name": "Count and Percentage",
  "operations": [
    { "id": "O007", "name": "Count",      "order": 1, "resultPattern": "XXX" },
    { "id": "O008", "name": "Percentage", "order": 2, "resultPattern": "XX.X",
      "referencedOperationRelationships": [
        { "id": "ROR001",
          "referencedOperationRole": { "controlledTerm": "DENOMINATOR" },
          "operationId": "O009",
          "analysisId": "AN_DENOM_001" }
      ] },
    { "id": "O009", "name": "Denominator", "order": 3, "resultPattern": "XXX" }
  ]
}
```

**Annotated-shell equivalent.** *"n (%), denominator = N in safety population"*. The
`AnalysisMethod` encodes this formally; the `resultPattern` encodes your format string;
the `ReferencedOperationRelationship` with `DENOMINATOR` encodes where the N comes from.

---

### 3.7 Analysis — The Central Object

Each `Analysis` ties together: a dataset, a variable, a population, optional data filter,
optional grouping factors, and a method.

**Key fields**

| Field | Description |
|-------|-------------|
| `dataset`, `variable` | The analysis dataset + variable (e.g., `ADAE` + `USUBJID`) |
| `analysisSetId` | References an `AnalysisSet` |
| `dataSubsetId` | References a `DataSubset` (optional) |
| `orderedGroupings` | Ordered list of `GroupingFactor` references |
| `methodId` | References an `AnalysisMethod` |
| `reason` | SPECIFIED IN PROTOCOL / SPECIFIED IN SAP / DATA DRIVEN |
| `purpose` | PRIMARY / SECONDARY / EXPLORATORY OUTCOME MEASURE |
| `documentRefs` | SAP references (with optional page numbers) |
| `programmingCode` | Code, or reference to a program file |
| `results` | The ARD — `[OperationResult]` |

**Shell row → ARM-TS Analysis field mapping**

| Annotated-shell annotation text | ARS field |
|--------------------------------|-----------|
| "USUBJID count from ADAE" | `dataset: "ADAE"`, `variable: "USUBJID"` |
| "Safety Population (SAFFL='Y')" | `analysisSetId: "AS002"` |
| "By treatment arm (TRT01A)" | `orderedGroupings: [{groupingId: "GF001", resultsByGroup: true}]` |
| "n (%)" | `methodId: "M002"` |
| "per Section 10.1 of SAP, pp. 42–44" | `documentRefs: [{referenceDocumentId: "SAP001", pageRefs: [...]}]` |
| "program: t_ae_summary.R" | `programmingCode.documentRef` |

This six-row mapping is the whole game. Every annotated-shell row decomposes into these
fields of one `Analysis` object.

---

### 3.8 Output and OutputDisplay

Defines the TFL — its file format, title, footnotes, and how analyses map into displays.

```json
{
  "id": "OUT001",
  "name": "Table 14.3.1 - Summary of Adverse Events",
  "fileSpecifications": [
    { "name": "Table 14.3.1 RTF", "fileType": { "controlledTerm": "rtf" },
      "location": "output/t_ae_summary.rtf" }
  ],
  "displays": [
    { "order": 1, "display": {
        "id": "DISP001",
        "displayTitle": "Table 14.3.1 Summary of Treatment-Emergent Adverse Events",
        "displaySections": [
          { "sectionType": "Title", "orderedSubSections": [{ "order": 1, "subSection": {
              "id": "TITLE001",
              "text": "Table 14.3.1 Summary of Treatment-Emergent Adverse Events" }}] },
          { "sectionType": "Footnote", "orderedSubSections": [{ "order": 1, "subSection": {
              "id": "FN001",
              "text": "TEAE = Treatment-Emergent Adverse Event. Subjects counted once per preferred term." }}] }
        ] } }
  ]
}
```

**Annotated-shell equivalent.** The **shell itself** — table number, title, footnotes,
abbreviations — maps directly to `Output` and `OutputDisplay`. Your title cell =
`displayTitle`. Your footnote cells = `DisplaySection` with `sectionType: "Footnote"`.

---

### 3.9 AnalysisProgrammingCodeTemplate

When a method is applied to many analyses with **minor variations** (e.g., different
`PARAMCD` values), the `codeTemplate` mechanism lets you write the code once with replacement
parameters.

```json
{
  "id": "M003",
  "name": "KM Analysis Template",
  "codeTemplate": {
    "context": "R version 4.3.1",
    "code": "km_fit <- survfit(Surv({{AVAR}}, {{CNSR_FLAG}}) ~ TRT01A, data = adtte %>% filter(PARAMCD == '{{PARAMCD}}' & ITTFL == 'Y'))",
    "parameters": [
      { "name": "AVAR",      "valueSource": "analysis.variable" },
      { "name": "PARAMCD",   "value": ["OS", "PFS", "DOR"] },
      { "name": "CNSR_FLAG", "value": ["CNSR"] }
    ]
  }
}
```

Individual analyses reference this template and supply their specific parameter values.

---

### 3.10 Naming Convention — Object IDs

Every ARM-TS object has an `id`. The LDM says only that IDs must be **strings**
and **unique within scope**; format is sponsor choice. But there is a strong
industry convention worth adopting from day one.

**Industry-standard prefix table** (used by the CDISC examples, by every
published ARM-TS the author has seen, and by this tutorial throughout):

| Object class | ID prefix | Examples |
|--------------|-----------|----------|
| `ReportingEvent` | `RE` | `RE001` |
| `ReferenceDocument` | `SAP`, `PROG`, `PROT` | `SAP001`, `PROG001` |
| `AnalysisSet` | `AS` | `AS001` (ITT), `AS002` (Safety) |
| `DataSubset` | `DS` | `DS001` |
| `GroupingFactor` | `GF` | `GF001` (Treatment Arm) |
| `Group` | `G` | `G001` (Drug A), `G002` (Placebo) |
| `AnalysisMethod` | `M` | `M001` (continuous), `M002` (count + %) |
| `Operation` | `O` | `O001` (n), `O002` (Mean), … `O008` (Pct) |
| `Analysis` | `AN` | `AN001`, `AN042` |
| `Output` | `OUT` | `OUT001` |
| `OutputDisplay` | `DISP` | `DISP001` |
| `DisplaySubSection` | `TITLE`, `HDR`, `FN`, `ABBR`, `FTR` | `TITLE001`, `FN001` |
| `ReferencedOperationRelationship` | `ROR` | `ROR001` |
| `SponsorTerm` | `ST` | `ST001` |

**Two equally valid naming styles:**

- **Style A — sequential code** (CDISC examples; dense, machine-friendly):
  `AS001`, `AS002`, `M001`, `AN001` … `AN045`
- **Style B — descriptive code** (longer, more readable in code review):
  `AS_ITT`, `AS_SAFETY`, `M_SUMM_CONT`, `AN_AGE_BY_TRT`

Pick one for the whole `ReportingEvent` and stick with it. Mixing styles
within one file is the hardest pattern to review.

**Validator checklist for IDs:**

1. **Presence** — every required `id` is non-empty
2. **Uniqueness** — no duplicates within scope
3. **Resolution** — every `*Id` reference resolves to an existing object
4. **Stability** — once published, do not rename IDs; bump
   `ReportingEvent.version` instead
5. **Characters** — stick to `[A-Za-z0-9_]`; no spaces, no hyphens

> Full reference (per-class scope rules, anti-patterns, conventions for
> SponsorTerm and TerminologyExtension IDs):
> [`materials/08_Object_ID_Conventions.md`](materials/08_Object_ID_Conventions.md).

---

### 3.11 Naming Convention — `name`, `label`, `description`

Most ARM-TS classes carry three text-content fields. They are not
interchangeable.

| Field | Purpose | Typical length | Example |
|-------|---------|----------------|---------|
| `name` | Canonical human-readable name. Goes in the annotated shell, in tooling lookups, on output displays. | 3–60 chars | `"Safety Population"` |
| `label` | Short display label for tight spaces (tooltips, column headers). | ≤ 30 chars | `"Safety Pop"` |
| `description` | Free-text explanation of intent / definition / nuance. | Sentence to paragraph | `"All subjects who received at least one dose of study drug (SAFFL='Y')"` |

**Rules of thumb:**

- `name` is **required on almost every class**. Omitting it is a schema violation.
- `label` is optional but recommended when displays have width constraints.
- `description` is optional but **highly recommended** for audit trail.
  Validators do not check it; reviewers absolutely do.
- Use **Title Case** for `name` and `label`. Sentence case is fine for `description`.

---

### 3.12 JSON Field Syntax — camelCase, References, Plurals

The ARS LDM uses LinkML serialized as JSON. Three rules apply universally.

**Rule 1 — camelCase field names.** Every field is `lowerCamelCase`. Class
names (in documentation) are `UpperCamelCase` to distinguish.

```json
"analysisSets":      [...]   ✅
"AnalysisSets":      [...]   ❌ PascalCase
"analysis_sets":     [...]   ❌ snake_case
"analysissets":      [...]   ❌ no boundary
```

**Rule 2 — Singular references end in `Id`; plural references end in `Ids`.**

```json
"analysisSetId":   "AS002"            ← single reference (string)
"categoryIds":     ["C001", "C003"]   ← multi-reference (array of strings)
```

A field ending `Id` (string) is a single reference; a field ending `Ids`
(array) is a list. **References use IDs, never nested objects** — this is
how the standard enforces Principle 1 (Reuse over repetition).

**Rule 3 — Collections are plural; elements are singular classes.**

```json
"analyses":    [Analysis, Analysis, ...]
"methods":     [AnalysisMethod, ...]
"operations":  [Operation, ...]
"groups":      [Group, ...]
"results":     [OperationResult, ...]
```

**Mutual-exclusion slot pairs** — exactly one of the two must be present,
never both:

| Class | Mutually exclusive slots |
|-------|--------------------------|
| `AnalysisSet`, `DataSubset`, `Group` | `condition` *or* `compoundExpression` |
| `ExtensibleTerminologyTerm` | `controlledTerm` *or* `sponsorTermId` |
| `PageRef` | `pageNumbers` / (`firstPage` + `lastPage`) *or* `pageNames` |

> Full reference (required-by-class table, ordered slots, data types,
> anti-patterns): [`materials/09_JSON_Field_Syntax_Reference.md`](materials/09_JSON_Field_Syntax_Reference.md).

---

### 3.13 `resultPattern` Grammar

`Operation.resultPattern` controls how `rawValue` becomes `formattedValue`.
The LDM says only "the default pattern or format" — the **`X`-grammar**
below is industry convention, used by every CDISC example and by `{cards}`
/ `{gtsummary}`.

| Character | Meaning |
|-----------|---------|
| `X` | One digit position |
| `.` | The decimal point |
| `(`, `)`, `%`, `:` | Literal characters appearing as-is in the formatted value |

**Counting rule:** the number of `X`s **after the decimal point** equals
the number of decimal places.

| Pattern | Decimals | Example raw → formatted |
|---------|----------|------------------------|
| `XXX` | 0 | `45` → `45` |
| `XX.X` | 1 | `84.44444` → `84.4` |
| `XXX.XX` | 2 | `5.234` → `5.23` |
| `X.XXX` | 3 | `0.0234` → `0.023` |
| `X.XXXX` | 4 | `0.00031` → `0.0003` |
| `(XX.X)` | 1, parenthesized | `84.4` → `(84.4)` |
| `XX.X%` | 1, percent-suffixed | `84.4` → `84.4%` |

`arsbridge::validate_ard()` Check #5 uses this rule: parse `rawValue` as a
double, count the decimal `X`s, `sprintf("%.<n>f", value)`, and compare to
`formattedValue`. Mismatches are flagged.

> Full grammar including edge cases (negative numbers, integer rawValue +
> decimal pattern, composite "n (%)" patterns):
> [`materials/10_resultPattern_and_ControlledTerms.md`](materials/10_resultPattern_and_ControlledTerms.md).

---

### 3.14 Controlled Terminology and Extensibility

ARS has **closed enumerations** (fixed, not extensible) and **extensible
enumerations** (CDISC values + sponsor extensions). All values are
case-sensitive and must match verbatim.

**Closed enumerations** — value goes directly in the slot:

| Enum | Values |
|------|--------|
| `ConditionComparatorEnum` | `EQ`, `NE`, `LT`, `LE`, `GT`, `GE`, `IN`, `NOTIN` |
| `ExpressionLogicalOperatorEnum` | `AND`, `OR`, `NOT` |
| `DisplaySectionTypeEnum` | `Title`, `Header`, `Rowlabel-Header`, `Body`, `Footnote`, `Abbreviation`, `Footer` |
| `PageRefTypeEnum` | `PhysicalRef`, `NamedDestination` |

**Extensible enumerations** — value goes in `controlledTerm` (CDISC value)
or `sponsorTermId` (sponsor extension):

| Enum | CDISC values |
|------|--------------|
| `AnalysisReasonEnum` | `SPECIFIED IN PROTOCOL`, `SPECIFIED IN SAP`, `DATA DRIVEN`, `REQUESTED BY REGULATORY AGENCY` |
| `AnalysisPurposeEnum` | `PRIMARY OUTCOME MEASURE`, `SECONDARY OUTCOME MEASURE`, `EXPLORATORY OUTCOME MEASURE` |
| `OperationRoleEnum` | `NUMERATOR`, `DENOMINATOR` |
| `OutputFileTypeEnum` | `pdf`, `rtf`, `txt` |

**Extending an enum** — add a `TerminologyExtension`:

```json
{
  "terminologyExtensions": [{
    "id": "TE001",
    "enumeration": "AnalysisReasonEnum",
    "sponsorTerms": [{
      "id": "ST001",
      "submissionValue": "PER PROTOCOL DEVIATION ANALYSIS",
      "description": "Added in response to protocol-deviation review"
    }]
  }]
}
```

Then on the consuming class:

```json
"reason": { "sponsorTermId": "ST001" }   ← instead of "controlledTerm"
```

> Every value, anti-patterns ("eq" lowercase, "Mean" instead of `Mean`),
> and the full extension pattern:
> [`materials/10_resultPattern_and_ControlledTerms.md`](materials/10_resultPattern_and_ControlledTerms.md).

---

## Hands-On: Build ARM-TS Objects

### What's in the sample `re_json`?

```{r ch3-preview, exercise=FALSE, echo=TRUE}
# AnalysisSets defined in the sample
sapply(re_json$analysisSets, function(x) paste0(x$id, " — ", x$name))
```

```{r ch3-preview-method, exercise=FALSE, echo=TRUE}
# Operations within Method M001
sapply(re_json$methods[[1]]$operations, function(o)
  paste0(o$id, " — ", o$name, " (", o$resultPattern, ")"))
```

---

## Exercise 3.1 — Read an AnalysisSet WhereClause {data-progressive=FALSE}

Extract the Safety Population (`AS002`) from the sample ARM-TS and print its filter condition
as a human-readable string of the form `"DATASET.VARIABLE COMPARATOR 'VALUE'"`. This is the
core logic of `arsbridge::ars_to_shell()`.

```{r ch3-ex1, exercise=TRUE, exercise.lines=12}
# Step 1: index into analysisSets for AS002 (Safety Population)
as_safe <- re_json$analysisSets[[___]]   # which index?

# Step 2: assemble the readable string from the condition
glue::glue(
  "{as_safe$condition$dataset}.",
  "{as_safe$condition$___} ",            # variable
  "{as_safe$condition$___} ",            # comparator
  "'{as_safe$condition$value[[1]]}'"
)
```

```{r ch3-ex1-hint-1}
# AS001 is ITT (index 1), AS002 is Safety (index 2)
# The condition has: dataset, variable, comparator, value
```

```{r ch3-ex1-solution}
as_safe <- re_json$analysisSets[[2]]
glue::glue(
  "{as_safe$condition$dataset}.",
  "{as_safe$condition$variable} ",
  "{as_safe$condition$comparator} ",
  "'{as_safe$condition$value[[1]]}'"
)
# "ADSL.SAFFL EQ 'Y'"
# This is the core logic of arsbridge::ars_to_shell().
```

```{r ch3-ex1-check}
grade_result(
  pass_if(~ as.character(.result) == "ADSL.SAFFL EQ 'Y'",
          "Correct — that's the Safety Population WhereClause as a human-readable string. arsbridge::ars_to_shell() does exactly this for every AnalysisSet, DataSubset, and Group in the ARM-TS."),
  fail_if(~ TRUE,
          "The Safety Population is AS002 (index 2). Make sure you used $variable and $comparator inside the glue() string.")
)
```

---

## Exercise 3.2 — Build an AnalysisMethod by Hand {data-progressive=FALSE}

Create an R list representing the **"n, mean, SD, median, min, max"** continuous summary method.
Follow the ARS object structure exactly — `id`, `name`, `operations`, and each operation has
`id`, `name`, `order`, `resultPattern`.

```{r ch3-ex2, exercise=TRUE, exercise.lines=20}
m_continuous <- list(
  id   = "M001",
  name = "Summary Statistics - Continuous",
  operations = list(
    list(id = "O001", name = "n",      order = 1L, resultPattern = "XXX"),
    list(id = "O002", name = ___,      order = 2L, resultPattern = "XXX.X"),    # fill in
    list(id = "O003", name = "SD",     order = 3L, resultPattern = ___),         # 2 dp
    list(id = "O004", name = "Median", order = 4L, resultPattern = "XXX.X"),
    list(id = "O005", name = ___,      order = 5L, resultPattern = "XXX"),       # minimum
    list(id = "O006", name = "Max",    order = 6L, resultPattern = "XXX")
  )
)

# How many operations does this method have?
length(m_continuous$operations)
```

```{r ch3-ex2-hint-1}
# The six standard continuous statistics are:
#   n, Mean, SD, Median, Min, Max
# resultPattern uses X placeholders to encode display precision:
#   "XXX"     = integer
#   "XXX.X"   = 1 decimal place
#   "XXX.XX"  = 2 decimal places
```

```{r ch3-ex2-solution}
m_continuous <- list(
  id   = "M001",
  name = "Summary Statistics - Continuous",
  operations = list(
    list(id = "O001", name = "n",      order = 1L, resultPattern = "XXX"),
    list(id = "O002", name = "Mean",   order = 2L, resultPattern = "XXX.X"),
    list(id = "O003", name = "SD",     order = 3L, resultPattern = "XXX.XX"),
    list(id = "O004", name = "Median", order = 4L, resultPattern = "XXX.X"),
    list(id = "O005", name = "Min",    order = 5L, resultPattern = "XXX"),
    list(id = "O006", name = "Max",    order = 6L, resultPattern = "XXX")
  )
)
length(m_continuous$operations)
# 6 operations — this method can now be referenced by any Analysis
# that needs summary stats on a continuous variable.
```

---

## Exercise 3.3 — Build a Group with a WhereClause {data-progressive=FALSE}

You need a `Group` object representing **"Xanomeline High Dose"** within the Treatment Arm
grouping factor. The group is defined by `ADSL.TRT01A == "Xanomeline High Dose"`.

```{r ch3-ex3, exercise=TRUE, exercise.lines=15}
g_high <- list(
  id   = ___,                       # "G001"
  name = "Xanomeline High Dose",
  condition = list(
    dataset    = "ADSL",
    variable   = ___,                # which ADSL variable identifies the treatment arm?
    comparator = ___,                # which ARS comparator?
    value      = list("Xanomeline High Dose")
  )
)

str(g_high)
```

```{r ch3-ex3-solution}
g_high <- list(
  id   = "G001",
  name = "Xanomeline High Dose",
  condition = list(
    dataset    = "ADSL",
    variable   = "TRT01A",
    comparator = "EQ",
    value      = list("Xanomeline High Dose")
  )
)
str(g_high)
# Notice the Group's condition has the SAME structure as an AnalysisSet's condition.
# This is why mastering WhereClause once unlocks AnalysisSet, DataSubset, and Group.
```

---

## Exercise 3.4 — Capstone: Annotated Shell → Analysis Object {data-progressive=FALSE}

You are a lead programmer. Your annotated shell for Table 14.1.1 says:

> *"n, mean, SD, median, min, max from ADSL.AGE, Safety Population (SAFFL='Y'), by treatment
> arm (TRT01A), per SAP Section 8.1"*

Build the `Analysis` object that encodes this annotation.

```{r ch3-ex4, exercise=TRUE, exercise.lines=22}
an_age <- list(
  id      = "AN001",
  name    = "Age Summary by Treatment Arm - Safety Population",
  reason  = list(controlledTerm = "SPECIFIED IN SAP"),
  purpose = list(controlledTerm = ___),       # Primary, Secondary, or Exploratory?
  dataset  = ___,                              # which ADaM dataset?
  variable = ___,                              # which variable?
  analysisSetId = ___,                         # which AnalysisSet ID? (Safety = AS002)
  orderedGroupings = list(
    list(order = 1L, groupingId = ___, resultsByGroup = TRUE)   # which GF?
  ),
  methodId = ___                               # which Method? (n, mean, SD, median, min, max)
)

str(an_age)
```

```{r ch3-ex4-hint-1}
# - It's a demographic baseline summary → PRIMARY OUTCOME MEASURE
# - Source: ADSL.AGE
# - Population: Safety = AS002
# - Grouping: Treatment Arm = GF001
# - Method: continuous summary = M001
```

```{r ch3-ex4-solution}
an_age <- list(
  id      = "AN001",
  name    = "Age Summary by Treatment Arm - Safety Population",
  reason  = list(controlledTerm = "SPECIFIED IN SAP"),
  purpose = list(controlledTerm = "PRIMARY OUTCOME MEASURE"),
  dataset  = "ADSL",
  variable = "AGE",
  analysisSetId = "AS002",
  orderedGroupings = list(
    list(order = 1L, groupingId = "GF001", resultsByGroup = TRUE)
  ),
  methodId = "M001"
)
str(an_age)
# One row of an annotated shell → one Analysis object.
# Field-by-field, the information content is identical.
```

```{r ch3-ex4-check}
grade_result(
  pass_if(~ identical(.result$dataset, "ADSL") &&
            identical(.result$variable, "AGE") &&
            identical(.result$analysisSetId, "AS002") &&
            identical(.result$methodId, "M001") &&
            identical(.result$orderedGroupings[[1]]$groupingId, "GF001"),
          "Correct — that's a complete ARM-TS Analysis object. This is the level of structure that arsbridge::shell_to_ars() will produce from your annotation text."),
  fail_if(~ TRUE,
          "Check: dataset='ADSL', variable='AGE', analysisSetId='AS002' (Safety), groupingId='GF001' (Treatment Arm), methodId='M001'.")
)
```

---

## Exercise 3.5 — Validate ID-Prefix Conventions {data-progressive=FALSE}

Section 3.10 introduced **Style A** prefix conventions: `AS` for AnalysisSet,
`GF` for GroupingFactor, `M` for AnalysisMethod, `AN` for Analysis, `O` for
Operation. The LDM does NOT enforce these — but our project (and the CDISC
examples) does. Write a tiny validator that confirms `re_json` conforms.

```{r ch3-ex5, exercise=TRUE, exercise.lines=22}
# Helper: do all items in a list have IDs starting with the given prefix?
check_prefix <- function(items, prefix) {
  ids <- sapply(items, function(x) x$___)         # which field carries the ID?
  all(startsWith(ids, prefix))
}

# Build the conformance table for the four collections in re_json
tibble::tribble(
  ~class,           ~conforms,
  "AnalysisSet",    check_prefix(re_json$analysisSets,      ___),  # expected prefix?
  "GroupingFactor", check_prefix(re_json$analysisGroupings, ___),
  "AnalysisMethod", check_prefix(re_json$methods,           ___),
  "Analysis",       check_prefix(re_json$analyses,          ___)
)
```

```{r ch3-ex5-hint-1}
# Each ARS object exposes its id as $id.
# Expected prefixes:
#   AnalysisSet     → "AS"
#   GroupingFactor  → "GF"
#   AnalysisMethod  → "M"
#   Analysis        → "AN"
```

```{r ch3-ex5-solution}
check_prefix <- function(items, prefix) {
  ids <- sapply(items, function(x) x$id)
  all(startsWith(ids, prefix))
}

tibble::tribble(
  ~class,           ~conforms,
  "AnalysisSet",    check_prefix(re_json$analysisSets,      "AS"),
  "GroupingFactor", check_prefix(re_json$analysisGroupings, "GF"),
  "AnalysisMethod", check_prefix(re_json$methods,           "M"),
  "Analysis",       check_prefix(re_json$analyses,          "AN")
)
# All four conform. In arsbridge, this is the LINT layer:
# it doesn't gate schema validity, but it surfaces inconsistencies that
# trip up downstream reviewers.
```

---

## Exercise 3.6 — Resolve a `*Id` Reference {data-progressive=FALSE}

Section 3.12 introduced the `*Id` reference pattern. An `*Id` is only
meaningful if the target object actually exists in the ReportingEvent —
that's `validate_ard()` Check #3 (reference resolution). Build the lookup
that powers it.

Given `Analysis AN001`, resolve its `analysisSetId` to the `AnalysisSet`
object it refers to, and print the resolved population name.

```{r ch3-ex6, exercise=TRUE, exercise.lines=20}
an <- re_json$analyses[[1]]
target_id <- an$___                       # which field carries the population reference?

# Build an ID -> object lookup for analysisSets
as_by_id <- setNames(
  re_json$analysisSets,
  sapply(re_json$analysisSets, function(x) x$id)
)

# Resolve and report
resolved <- as_by_id[[target_id]]
if (is.null(resolved)) {
  cat("UNRESOLVED reference:", target_id, "\n")
} else {
  cat("Reference", target_id, "resolves to AnalysisSet:", resolved$___, "\n")
}
```

```{r ch3-ex6-hint-1}
# An Analysis points at its population via the analysisSetId field.
# Each AnalysisSet exposes a $name field for the human-readable name.
```

```{r ch3-ex6-solution}
an <- re_json$analyses[[1]]
target_id <- an$analysisSetId

as_by_id <- setNames(
  re_json$analysisSets,
  sapply(re_json$analysisSets, function(x) x$id)
)

resolved <- as_by_id[[target_id]]
if (is.null(resolved)) {
  cat("UNRESOLVED reference:", target_id, "\n")
} else {
  cat("Reference", target_id, "resolves to AnalysisSet:", resolved$name, "\n")
}
# "Reference AS002 resolves to AnalysisSet: Safety Population"
# Run this same pattern over methodId, groupingId, groupId, operationId,
# referenceDocumentId — and you have validate_ard() Check #3.
```

---

## Chapter 3 Knowledge Check

```{r ch3-quiz, echo=FALSE}
quiz(
  caption = "Chapter 3 — Check Your Understanding",

  question(
    "Where in ARM-TS is the link from an Analysis to a specific SAP section recorded?",
    answer("In ReportingEvent.description"),
    answer("In Analysis.documentRefs, pointing to a ReferenceDocument", correct = TRUE),
    answer("In AnalysisMethod.codeTemplate"),
    answer("In Output.fileSpecifications"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Your shell says: 'PARAMCD = OS in ADTTE'. Which ARM-TS object encodes this?",
    answer("AnalysisSet"),
    answer("DataSubset", correct = TRUE),
    answer("GroupingFactor"),
    answer("AnalysisMethod"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "When `dataDriven = false` on a GroupingFactor, what does that imply?",
    answer("The groups must be pre-specified as `groups` with WhereClauses", correct = TRUE),
    answer("The groups are derived from the data at runtime"),
    answer("Only one group is allowed"),
    answer("The grouping factor is not used in any analysis"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "An AnalysisMethod's `resultPattern` of 'XXX.XX' means what?",
    answer("The operation produces 3 integer digits and 2 decimal digits in the formatted output", correct = TRUE),
    answer("There are 5 operations in the method"),
    answer("The result is censored at 3 significant figures"),
    answer("It is a placeholder for a percentage symbol"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "In Method M002 (Count and Percentage), why does the Percentage operation carry a `referencedOperationRelationships` block?",
    answer(
      "To formally encode that the Percentage operation depends on the Denominator operation (NUMERATOR / DENOMINATOR roles)",
      correct = TRUE
    ),
    answer("To indicate the operation is optional"),
    answer("To specify which programmer wrote the code"),
    answer("To version-control the operation"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "The annotated-shell phrase 'By treatment arm (TRT01A)' maps to which field of an Analysis object?",
    answer("dataset"),
    answer("orderedGroupings (referencing a GroupingFactor)", correct = TRUE),
    answer("analysisSetId"),
    answer("documentRefs"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "What is the purpose of AnalysisProgrammingCodeTemplate?",
    answer(
      "To write code ONCE with replacement parameters (e.g., PARAMCD) and reuse it across many analyses that vary only in those parameters",
      correct = TRUE
    ),
    answer("To store the SAS macro that produced the table"),
    answer("To define the table title and footnote text"),
    answer("To encode the resultPattern for each operation"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which Output sub-object holds the table TITLE and FOOTNOTE text?",
    answer("Output.fileSpecifications"),
    answer("OutputDisplay.displaySections (with sectionType 'Title' / 'Footnote')", correct = TRUE),
    answer("Analysis.documentRefs"),
    answer("ReportingEvent.label"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "In the sample re_json (and every example in this tutorial), 'AS' prefixes AnalysisSet IDs (AS001, AS002), 'M' prefixes Method IDs, 'AN' prefixes Analysis IDs. What does this convention represent?",
    answer(
      "An industry convention used by the CDISC examples — NOT mandated by the ARS LDM",
      correct = TRUE
    ),
    answer("A strict requirement enforced by the ARS schema validator"),
    answer("An FDA submission requirement that overrides sponsor naming"),
    answer("A reserved keyword that learnr converts to the corresponding class"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which field name is VALID in an ARM-TS JSON file?",
    answer("analysisSetId", correct = TRUE),
    answer("AnalysisSetId"),
    answer("analysis_set_id"),
    answer("AnalysisSet"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "An AnalysisSet has BOTH a `condition` field AND a `compoundExpression` field populated in the same object. What is the correct behavior?",
    answer(
      "Schema violation — exactly one of `condition` or `compoundExpression` must be present, never both",
      correct = TRUE
    ),
    answer("Both are honored; condition is applied first, then the compoundExpression"),
    answer("Allowed only when the conditions match"),
    answer("Allowed but only at the ReportingEvent level"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "An Operation has resultPattern `XXX.XX`. What does this imply for formattedValue?",
    answer("Up to three integer digits and exactly 2 decimal places", correct = TRUE),
    answer("Three operations chained together"),
    answer("A range from 0 to 99,999.99"),
    answer("Currency format with two cents shown"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "An Analysis has `reason = { 'controlledTerm': 'specified in sap' }`. What is the issue?",
    answer(
      "AnalysisReasonEnum values are case-sensitive; the correct value is 'SPECIFIED IN SAP' (uppercase exactly)",
      correct = TRUE
    ),
    answer("controlledTerm must be a number, not a string"),
    answer("Must use sponsorTermId instead — controlledTerm is for sponsor extensions"),
    answer("The value must be wrapped in an array"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

---

## Chapter 3 Summary {.congratulations}

You have now seen every major ARM-TS object with:

- **Full field definitions** for `ReportingEvent`, `ReferenceDocument`, `AnalysisSet`,
  `DataSubset`, `GroupingFactor` / `Group`, `AnalysisMethod` / `Operation`, `Analysis`,
  `Output` / `OutputDisplay`, and `AnalysisProgrammingCodeTemplate`
- **JSON examples** grounded in a realistic AE summary table
- **Explicit field-by-field mapping** to what you already write in annotated shells
- **Hands-on practice** building an `AnalysisMethod`, a `Group`, and a full `Analysis` object

And the **rules and conventions** that make an ARM-TS file machine-validatable:

- **Object ID conventions** (§3.10) — the Style A prefix system (`AS`, `GF`, `M`,
  `AN`, …) used by CDISC examples; uniqueness, resolution, and stability rules
- **`name` / `label` / `description` semantics** (§3.11) — what goes where, and why
- **JSON field syntax** (§3.12) — camelCase universally, `*Id` for single references,
  `*Ids` for multi-references, plural collections, mutual-exclusion slot pairs
  (`condition` xor `compoundExpression`, `controlledTerm` xor `sponsorTermId`)
- **`resultPattern` grammar** (§3.13) — the `X`-digit convention used by every CDISC
  example, by `{cards}` / `{gtsummary}`, and by `arsbridge::validate_ard()` Check #5
- **Controlled terminology** (§3.14) — every closed enum value verbatim, the four
  extensible enums, and the `TerminologyExtension` / `SponsorTerm` extension pattern

You also prototyped two of the validation checks `arsbridge::validate_ard()`
will perform: prefix-convention lint (Exercise 3.5) and `*Id` reference
resolution (Exercise 3.6).

**The key insight:** everything in an annotated shell has an ARM-TS equivalent. The information
content is identical; the representation is **machine-readable JSON vs. human-readable
annotation text**.

**Next:** Chapter 4 — ARD Deep Dive (what goes inside `Analysis.results` when analyses are
actually run, and how `{cards}` generates flat ARD data frames from ADaM data)
