---
title: "CDISC Analysis Results Standard (ARS): Interactive Training"
subtitle: "Chapter 5: The Traceability Chain"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Chapter 5 walks one analysis end-to-end: SAP language, ARM-TS objects,
  ADaM data, R code, ARD results, display cell. You will re-derive a stored
  ARD result from ADaM, find a deliberate traceability break by switching the
  population, and prototype a format-consistency check (rawValue vs.
  formattedValue).
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(jsonlite)
library(glue)

tutorial_options(exercise.timelimit = 60)
gradethis_setup()

# Load pharmaverse ADaM datasets
data("adsl",  package = "pharmaverseadam")
data("adae",  package = "pharmaverseadam")
# pharmaverseadam 1.4+ replaced the generic "adtte" with TA-specific
# datasets; use the oncology one and join subject-level vars in from
# ADSL. pharmaverseadam ADSL lacks ITTFL, so derive it from ARM (a
# subject is "randomised / ITT" iff ARM is non-empty).
data("adtte_onco", package = "pharmaverseadam")
adtte <- adtte_onco %>%
  left_join(
    adsl %>%
      mutate(ITTFL = ifelse(!is.na(ARM) & ARM != "", "Y", "N")) %>%
      select(USUBJID, TRT01A, ITTFL),
    by = "USUBJID"
  )
data("adlb",  package = "pharmaverseadam")

adsl_safe <- adsl %>% filter(SAFFL == "Y")
adsl_itt  <- adsl %>% filter(!is.na(ARM) & ARM != "")
adae_teae <- adae %>% filter(TRTEMFL == "Y")

# Analysis dataset for AE counts (Safety pop, any-TEAE flag)
analysis_ae <- adsl_safe %>%
  select(USUBJID, TRT01A) %>%
  left_join(
    adae_teae %>% distinct(USUBJID) %>% mutate(any_teae = 1L),
    by = "USUBJID"
  ) %>%
  mutate(any_teae = dplyr::coalesce(any_teae, 0L))

re_json <- jsonlite::read_json("data/reporting_event.json")
```

---

## Chapter 5: The Traceability Chain

### 5.1 The Chain

Everything you have learned so far comes together in this chapter. ARS enables a formal,
**machine-navigable traceability chain**:

```
Protocol / SAP
      ↓
  ARM-TS  (analysis specification)
      ↓
  ADaM    (analysis data)
      ↓
  R code  (execution)
      ↓
  ARD     (computed results)
      ↓
  Output Display  (table / figure)
```

Every link should be verifiable. The problem today is that **without ARS, the chain is
broken between the SAP and everything downstream**. The only link is a human reading the
SAP and deciding what the code should do. Every check on that link is a person comparing
prose to code.

---

### 5.2 Complete Worked Example

We will trace one specific analysis from SAP language through to a display cell. The case
study is the **Age summary by treatment arm in the Safety Population**, the same analysis
encoded in your sample `re_json`. You have met every object involved; now watch them work
as a chain.

**Step 1: the SAP.** Section 8.1 of the SAP says:

> *"Baseline age (years) will be summarized using n, mean, standard deviation, median,
> minimum, and maximum by treatment arm, in the Safety Population (subjects who received at
> least one dose of study drug, SAFFL='Y' in ADSL)."*

Today, a lead programmer reads this and writes an annotation. With ARS, this text is
**formally encoded** as the ARM-TS objects below.

**Step 2: the ARM-TS objects.** Encoded directly in `re_json`:

- `ReferenceDocument` `"SAP001"`: the SAP PDF
- `AnalysisSet` `"AS002"`: Safety Population (`ADSL.SAFFL EQ 'Y'`)
- `GroupingFactor` `"GF001"`: Treatment Arm (`TRT01A`), with two pre-specified `Group`s:
  `G001` Xanomeline High Dose and `G002` Placebo
- `AnalysisMethod` `"M001"`: n, Mean, SD, Median, Min, Max (operations `O001` to `O006`)
- `Analysis` `"AN001"`: ties it all together: `dataset = "ADSL"`, `variable = "AGE"`,
  `analysisSetId = "AS002"`, `orderedGroupings[[1]].groupingId = "GF001"`, `methodId = "M001"`

**Step 3: the ADaM data.** ADSL, one row per subject, with `TRT01A`, `AGE`, `SAFFL`.

**Step 4: the R code.** The ARM-TS tells the programmer exactly what to do. Notice how
each line of code traces to a specific object:

```r
adsl_safe <- adsl %>% filter(SAFFL == "Y")   # from AS002
adsl_safe %>%
  group_by(TRT01A) %>%                       # from GF001
  summarise(
    n      = sum(!is.na(AGE)),               # O001
    mean   = round(mean(AGE),       1),      # O002, resultPattern "XXX.X"
    sd     = round(sd(AGE),         2),      # O003, resultPattern "XXX.XX"
    median = round(median(AGE),     1),      # O004
    min    = min(AGE),                       # O005
    max    = max(AGE)                        # O006
  )
```

**Step 5: the ARD.** The stored result in `re_json$analyses[[1]]$results` says:

| operationId | groupId | rawValue | formattedValue |
|-------------|---------|----------|----------------|
| `O002` (Mean) | `G001` (Xanomeline High Dose) | `74.38095` | `74.4` |
| `O002` (Mean) | `G002` (Placebo)              | `75.20930` | `75.2` |

**Step 6: the display.** A table cell reads `74.4` under the Xanomeline High Dose column,
and that value is fully traceable: `O002.formattedValue`, for `G001` in `GF001`, in the
Safety Population `AS002`, on `ADSL.AGE`, per SAP Section 8.1.

Every number is **machine-navigable to its source**. That sentence is the entire value
proposition of ARS, and you just watched it hold for a real cell.

---

### 5.3 What `arsbridge::validate_ard()` Will Check

Given the ARM-TS and the ARD, `arsbridge` validates six things:

| Check | How |
|-------|-----|
| 1. Does every Analysis have ARD rows? | Match analysisId across ARM-TS and ARD |
| 2. Does the ARD match what you'd recompute from ADaM? | Re-run using ARM-TS parameters, compare to stored result |
| 3. Does the AnalysisSet match the ARD denominator? | Apply `AnalysisSet.condition` to ADaM, count, compare |
| 4. Are all expected group combinations present? | Cross-reference `ResultGroup.groupId` against `Group` |
| 5. Does `formattedValue` match `rawValue + resultPattern`? | Apply pattern to `rawValue`, compare |
| 6. Does the SAP page contain the analysis description? | Structural check; partially automatable with PDF parsing |

You will prototype Check #2 (Exercise 5.1) and Check #5 (Exercise 5.3) below, and you will
see what a Check #3 failure looks like in Exercise 5.2.

---

### 5.4 The Master Annotated-Shell to ARM-TS Mapping

This is the design reference for `arsbridge::shell_to_ars()`. Keep it; you will use it in
Exercise 5.4 and again in Chapter 7.

| Annotated-shell element | ARM-TS object | Key fields |
|-------------------------|---------------|------------|
| Table number & title | `Output` / `OutputDisplay` | `id`, `displayTitle` |
| Population note (e.g., "Safety Pop, SAFFL=Y") | `AnalysisSet` | `id`, `condition` |
| Data filter (e.g., "PARAMCD='OS'") | `DataSubset` | `id`, `condition` |
| Dataset reference (e.g., "from ADTTE") | `Analysis.dataset` | `dataset` |
| Variable reference (e.g., "AVAL") | `Analysis.variable` | `variable` |
| Column grouping (e.g., "by TRT01A") | `GroupingFactor` + `OrderedGroupingFactor` | `groupingVariable`, `groupingId` |
| Row grouping (e.g., "by AVISIT") | Second `GroupingFactor` + `OrderedGroupingFactor` | |
| Statistic description (e.g., "n, mean, SD") | `AnalysisMethod` + `[Operation]` | `methodId`, `operations` |
| Format string (e.g., "XX.X") | `Operation.resultPattern` | `resultPattern` |
| n (%) denominator note | `ReferencedOperationRelationship` (DENOMINATOR) | |
| SAP section reference | `DocumentReference` + `PageRef` | `referenceDocumentId`, `pageRefs` |
| Program file reference | `programmingCode.documentRef` | |
| Pre-specified / data-driven | `Analysis.reason` | SPECIFIED IN SAP vs. DATA DRIVEN |
| Primary / secondary / exploratory | `Analysis.purpose` | PRIMARY / SECONDARY / EXPLORATORY |
| Footnote text | `DisplaySection` (`sectionType: Footnote`) | `text` |
| Abbreviation text | `DisplaySection` (`sectionType: Abbreviation`) | `text` |

---

## Hands-On: Verify a Stored Result End-to-End

### Pull the stored mean from the ARD

First, look at what the ARD actually stores. The table below lists every OperationResult
in the sample, in storage order. Note the positions: the Mean (`O002`) for `G001`
(Xanomeline High Dose) sits at `results[[2]]`, and the Mean for `G002` (Placebo) at
`results[[4]]`. Exercise 5.1 indexes into exactly those two.

```{r ch5-preview, exercise=FALSE, echo=TRUE}
# What the ARM-TS says happened: mean AGE per treatment arm, Safety Pop
stored <- re_json$analyses[[1]]$results
data.frame(
  operationId    = sapply(stored, function(r) r$operationId),
  groupId        = sapply(stored, function(r) r$resultGroups[[1]]$groupId),
  rawValue       = sapply(stored, function(r) r$rawValue),
  formattedValue = sapply(stored, function(r) r$formattedValue),
  stringsAsFactors = FALSE
)
```

---

## Exercise 5.1: Re-derive the Stored Mean from ADSL {data-progressive=FALSE}

**Step 1: the annotated shell.**

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; text-align:center;">Table 14.1.1: Demographic Characteristics</div>
<div style="text-align:center; font-family:sans-serif; margin-bottom:8px;">Safety Population &nbsp;(<span style="color:#b3400e; font-weight:bold;">ADSL.SAFFL='Y'</span>)</div>
<table style="border-collapse:collapse; width:100%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left; width:44%;"></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Xanomeline High Dose<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Placebo<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX</span></th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">Mean Age &nbsp;<span style="color:#b3400e; font-weight:bold;">ADSL.AGE</span></td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x</td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">Grouping: <span style="color:#b3400e;">TRT01A</span> &nbsp;|&nbsp; Format: <span style="color:#b3400e;">XXX.X</span> &nbsp;|&nbsp; SAP §8.1</div>
</div>

**Step 2: the ARM-TS encoding.** Spec plus stored results, side by side in `AN001`:

```json
{ "id": "AN001", "dataset": "ADSL", "variable": "AGE",
  "analysisSetId": "AS002",
  "methodId": "M001",
  "results": [
    { "operationId": "O002", "resultGroups": [{ "groupId": "G001" }],
      "rawValue": "74.38095", "formattedValue": "74.4" },
    { "operationId": "O002", "resultGroups": [{ "groupId": "G002" }],
      "rawValue": "75.20930", "formattedValue": "75.2" }
  ] }
```

**Step 3: your turn.** This is Check #2 by hand. Apply the spec exactly as written
(`ADSL.AGE`, Safety population, by `TRT01A`) and re-compute the mean age per arm. Then
compare your computed values to the stored `formattedValue`s. If they match, the chain
holds.

```{r ch5-ex1, exercise=TRUE, exercise.lines=18}
# Stored values from the ARD (results[[2]] = Mean for G001, results[[4]] = Mean for G002)
stored_mean_hd <- re_json$analyses[[1]]$results[[2]]$formattedValue
stored_mean_pl <- re_json$analyses[[1]]$results[[4]]$formattedValue

# Re-compute the means from ADSL, Safety population, by treatment arm
computed_means <- adsl_safe %>%
  group_by(___) %>%                                  # which variable groups the result?
  summarise(mean_age = round(mean(___), 1))          # which variable is being summarised?

computed_means
cat("\nStored mean (Xanomeline High Dose):", stored_mean_hd, "\n")
cat("Stored mean (Placebo):              ", stored_mean_pl, "\n")
```

```{r ch5-ex1-hint-1}
# group_by(TRT01A) and summarise(mean_age = round(mean(AGE), 1))
# The resultPattern for Mean is "XXX.X" -> 1 decimal place
```

```{r ch5-ex1-solution}
stored_mean_hd <- re_json$analyses[[1]]$results[[2]]$formattedValue
stored_mean_pl <- re_json$analyses[[1]]$results[[4]]$formattedValue

computed_means <- adsl_safe %>%
  group_by(TRT01A) %>%
  summarise(mean_age = round(mean(AGE), 1))

computed_means
cat("\nStored mean (Xanomeline High Dose):", stored_mean_hd, "\n")
cat("Stored mean (Placebo):              ", stored_mean_pl, "\n")
# If the computed values match the stored formattedValues, traceability
# is verified. This is exactly what arsbridge::validate_ard() will automate.
```

---

## Exercise 5.2: Find a Deliberate Traceability Break {data-progressive=FALSE}

Now break the chain on purpose, so you know what a break looks like.

Suppose the stored ARD says mean age for Placebo = `75.2`, but the ARM-TS spec says the
population should be **Safety** (`SAFFL='Y'`). What if a programmer accidentally used the
**ITT** subset instead? Would the result still match?

This scenario is a **population traceability break**: the stored result was computed
against a different population than the ARM-TS specifies. It is Check #3 territory, and it
is one of the most dangerous bugs in clinical reporting because the numbers usually look
plausible.

```{r ch5-ex2, exercise=TRUE, exercise.lines=18}
# Compute mean age in ITT (what a buggy program might do)
itt_means <- adsl_itt %>%
  group_by(TRT01A) %>%
  summarise(mean_age_ITT = round(mean(AGE), 1))

# Compute mean age in Safety (what the ARM-TS specifies)
safe_means <- adsl_safe %>%
  group_by(TRT01A) %>%
  summarise(mean_age_Safety = round(mean(AGE), 1))

# Join them and look at the differences side by side
left_join(itt_means, safe_means, by = "TRT01A") %>%
  mutate(diff = mean_age_Safety - mean_age_ITT)
```

```{r ch5-ex2-solution}
itt_means <- adsl_itt %>%
  group_by(TRT01A) %>%
  summarise(mean_age_ITT = round(mean(AGE), 1))

safe_means <- adsl_safe %>%
  group_by(TRT01A) %>%
  summarise(mean_age_Safety = round(mean(AGE), 1))

left_join(itt_means, safe_means, by = "TRT01A") %>%
  mutate(diff = mean_age_Safety - mean_age_ITT)
# In CDISCPILOT01 the difference is small (a few subjects who randomised but
# did not dose are excluded from Safety). In a real study with screen-failures
# or major withdrawals, this difference would be large, and would be a flag.
# validate_ard() catches this by always re-applying the ARM-TS AnalysisSet
# rather than trusting the ARD's stored denominator alone.
```

---

## Exercise 5.3: Prototype a Format-Consistency Check {data-progressive=FALSE}

**Step 1: the annotated shell.** The format column of the shell:

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<table style="border-collapse:collapse; width:60%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left;">Statistic &nbsp;<span style="color:#b3400e; font-weight:normal; font-size:0.9em;">ADSL.AGE</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center; color:#b3400e;">Format (resultPattern)</th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">Mean</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center; color:#b3400e; font-weight:bold;">XXX.X</td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">One X after decimal = 1 decimal place: rawValue "74.38095" → formattedValue "<span style="color:#b3400e;">74.4</span>"</div>
</div>

**Step 2: the ARM-TS encoding.** The pattern lives on the Operation; the values live in
the result. Check #5 verifies these two agree:

```json
"operations": [ { "id": "O002", "name": "Mean", "resultPattern": "XXX.X" } ]

"results":    [ { "operationId": "O002",
                  "rawValue": "74.38095", "formattedValue": "74.4" } ]
```

**Step 3: your turn.** Does `formattedValue` match what `rawValue` would produce under
`"XXX.X"`? The pattern means **1 decimal place** (count the `X`s after the decimal point,
per the Chapter 3 grammar).

The R tool for controlled rounding-to-string is `sprintf()`: the format `"%.1f"` means "a
fixed-point number with 1 decimal", so `sprintf("%.1f", 74.38095)` gives `"74.4"`. Building
the format string dynamically is just `paste0("%.", n_decimals, "f")`.

Write a small helper `check_format()` that takes a raw and formatted value plus a decimal
count, and returns `TRUE` if they agree.

```{r ch5-ex3, exercise=TRUE, exercise.lines=22}
check_format <- function(raw_value, formatted_value, n_decimals) {
  # 1. Convert raw_value (a string from the ARD) to numeric
  raw_num <- as.numeric(___)

  # 2. Format it to n_decimals using sprintf
  expected <- sprintf(paste0("%.", n_decimals, "f"), ___)

  # 3. Compare to the stored formatted value
  expected == formatted_value
}

# Run it on the Mean for Xanomeline High Dose (O002, G001)
res <- re_json$analyses[[1]]$results[[2]]
check_format(res$rawValue, res$formattedValue, n_decimals = 1)
```

```{r ch5-ex3-hint-1}
# Mean for G001:
#   rawValue       = "74.38095"
#   formattedValue = "74.4"
# With n_decimals = 1, sprintf("%.1f", 74.38095) == "74.4" -> TRUE
```

```{r ch5-ex3-solution}
check_format <- function(raw_value, formatted_value, n_decimals) {
  raw_num  <- as.numeric(raw_value)
  expected <- sprintf(paste0("%.", n_decimals, "f"), raw_num)
  expected == formatted_value
}

res <- re_json$analyses[[1]]$results[[2]]
check_format(res$rawValue, res$formattedValue, n_decimals = 1)
# TRUE: formattedValue is consistent with rawValue + resultPattern.
# In arsbridge, the n_decimals is derived from Operation.resultPattern
# by counting the X's after the decimal point in "XXX.X" -> 1.
```

```{r ch5-ex3-check}
grade_result(
  pass_if(~ isTRUE(.result),
          "Correct. That's Check #5 of validate_ard(). In the real package, n_decimals is derived from Operation.resultPattern by counting X's after the decimal."),
  fail_if(~ TRUE,
          "Use as.numeric(raw_value) and sprintf('%.1f', raw_num).")
)
```

---

## Exercise 5.4: Map an Annotated-Shell Row to ARM-TS Objects (capstone) {data-progressive=FALSE}

**Step 1: the annotated shell.** You're handed this row:

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; text-align:center;">Table 14.3.1: Summary of Treatment-Emergent Adverse Events</div>
<div style="text-align:center; font-family:sans-serif; margin-bottom:8px;">Safety Population &nbsp;(<span style="color:#b3400e; font-weight:bold;">ADSL.SAFFL='Y'</span>)</div>
<table style="border-collapse:collapse; width:100%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left; width:44%;"></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Xanomeline High Dose<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX n (%)</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Xanomeline Low Dose<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX n (%)</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Placebo<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX n (%)</span></th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">Subjects with any TEAE &nbsp;<span style="color:#b3400e; font-weight:bold;">ADAE.USUBJID</span><br><span style="color:#b3400e; font-size:0.88em;">ADAE.TRTEMFL='Y'</span></td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx (xx.x)</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx (xx.x)</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx (xx.x)</td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">Grouping: <span style="color:#b3400e;">TRT01A</span> &nbsp;|&nbsp; Format: <span style="color:#b3400e;">XXX (XX.X)</span> &nbsp;|&nbsp; SAP §10.1 pp. 42-44 &nbsp;|&nbsp; Decompose into 6 ARS objects</div>
</div>

**Step 2: the ARM-TS target.** Six objects, connected only by IDs:

```json
"analysisSets":      [ { "id": "AS002", "...": "..." } ],
"analysisGroupings": [ { "id": "GF001", "...": "..." } ],
"methods":           [ { "id": "M002",  "...": "..." } ],
"analyses":          [ { "id": "AN010",
                         "analysisSetId": "AS002",
                         "methodId": "M002", "...": "..." } ],
"referenceDocuments":[ { "id": "SAP001", "...": "..." } ],
"outputs":           [ { "id": "OUT001", "...": "..." } ]
```

**Step 3: your turn.** Decompose the shell row into those six objects, exactly what
`arsbridge::shell_to_ars()` would emit. Use the Section 5.4 mapping table as your
reference. Fill in the missing pieces.

```{r ch5-ex4, exercise=TRUE, exercise.lines=22}
shell_row_mapping <- tribble(
  ~ars_object,        ~field,              ~value_from_shell,

  "AnalysisSet",      "id",                "AS002",
  "AnalysisSet",      "condition",         "ADSL.SAFFL EQ 'Y'",

  "GroupingFactor",   "id",                "GF001",
  "GroupingFactor",   "groupingVariable",  ___,                          # which ADSL var?

  "AnalysisMethod",   "id",                "M002",
  "AnalysisMethod",   "name",              "Count and Percentage",

  "Analysis",         "dataset",           ___,                          # which ADaM dataset?
  "Analysis",         "variable",          "USUBJID",
  "Analysis",         "analysisSetId",     ___,                          # which AnalysisSet?
  "Analysis",         "methodId",          ___,                          # which Method?

  "DocumentReference","referenceDocumentId","SAP001",
  "DocumentReference","pageRefs",          "firstPage=42, lastPage=44",

  "Output",           "displayTitle",      "Table 14.3.1: Subjects with any TEAE"
)

shell_row_mapping
```

```{r ch5-ex4-solution}
shell_row_mapping <- tribble(
  ~ars_object,        ~field,              ~value_from_shell,
  "AnalysisSet",      "id",                "AS002",
  "AnalysisSet",      "condition",         "ADSL.SAFFL EQ 'Y'",
  "GroupingFactor",   "id",                "GF001",
  "GroupingFactor",   "groupingVariable",  "TRT01A",
  "AnalysisMethod",   "id",                "M002",
  "AnalysisMethod",   "name",              "Count and Percentage",
  "Analysis",         "dataset",           "ADAE",
  "Analysis",         "variable",          "USUBJID",
  "Analysis",         "analysisSetId",     "AS002",
  "Analysis",         "methodId",          "M002",
  "DocumentReference","referenceDocumentId","SAP001",
  "DocumentReference","pageRefs",          "firstPage=42, lastPage=44",
  "Output",           "displayTitle",      "Table 14.3.1: Subjects with any TEAE"
)
shell_row_mapping
# One shell row decomposes into six referenced objects. That decomposition
# is the entire shell_to_ars() function.
```

---

## Chapter 5 Knowledge Check

Review map: the chain and its weak link (5.1), re-derivation (5.2 and Exercise 5.1), the
six validation checks (5.3), break types (Exercise 5.2), shell decomposition (5.4).

```{r ch5-quiz, echo=FALSE}
quiz(
  caption = "Chapter 5: Check Your Understanding",

  question(
    "In the traceability chain, which link is currently the WEAKEST in industry today (without ARS)?",
    answer("ADaM → R code"),
    answer("R code → ARD"),
    answer("SAP → ARM-TS / annotated shell", correct = TRUE),
    answer("ARD → display"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "A re-computed mean from ADSL (Safety pop) matches the ARD's stored rawValue. What does this verify?",
    answer(
      "That the stored result is consistent with applying the ARM-TS specification to the ADaM data",
      correct = TRUE
    ),
    answer("That the statistical method is appropriate for the variable"),
    answer("That the SAP language is unambiguous"),
    answer("That the ARM-TS schema is valid against ars_ldm.yaml"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "If the stored ARD was computed against the ITT population but the ARM-TS specifies Safety, what KIND of traceability break is this?",
    answer("A format-consistency break (rawValue vs. formattedValue)"),
    answer("A population traceability break (AnalysisSet not applied as specified)", correct = TRUE),
    answer("A grouping completeness break (a Group is missing from the ARD)"),
    answer("A coverage break (Analysis defined in ARM-TS has no ARD rows)"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Check #5 of validate_ard() compares formattedValue against what?",
    answer("The value in the rendered RTF table"),
    answer("rawValue, after applying the operation's resultPattern", correct = TRUE),
    answer("The denominator from a different Analysis"),
    answer("The result re-computed from ADaM"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which ARM-TS objects does ONE row of an annotated shell typically decompose into?",
    answer("Only an Analysis"),
    answer("An AnalysisSet, GroupingFactor, AnalysisMethod, Analysis, DocumentReference, and Output", correct = TRUE),
    answer("Only an Output and OutputDisplay"),
    answer("A single OperationResult"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Why does the chain need to be MACHINE-navigable rather than just documented in the SAP?",
    answer(
      "Machine-navigability lets a regulator (or arsbridge) traverse from a single result cell back to its full context automatically, with no manual SAP reading required",
      correct = TRUE
    ),
    answer("Because SAPs cannot be written in English"),
    answer("Because R cannot read PDF files"),
    answer("Because Define-XML already provides this for ADaM"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

---

## Chapter 5 Summary {.congratulations}

You can now trace a single cell in a clinical trial table all the way back to:

- The **SAP language** that specified it
- The **ARM-TS objects** that formally encode that specification
- The **ADaM data** it operates on
- The **R code** that computed it
- The **ARD** that stores the result
- The **display format** that renders it

And you have prototyped two of the six validation checks `arsbridge::validate_ard()` will
perform: **re-derivation from ADaM** (Check #2) and **format consistency** (Check #5). You
also watched a **population traceability break** (Check #3) happen in slow motion. The
shell-row mapping table in Section 5.4 is your primary design reference for the
`shell_to_ars()` function in Chapter 7.

**Next:** Chapter 6, R Ecosystem and Tools: what `{cards}`, `{cardx}`, `{gtsummary}`, and
`{datasetjson}` cover, and exactly where the open-source ARM-TS gap is
