---
title: "CDISC Analysis Results Standard (ARS) — Interactive Training"
subtitle: "Chapter 1: Why ARS Exists"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  A hands-on tutorial covering CDISC ARS v1.0 using real pharmaverse datasets.
  Chapter 1 covers the five chronic failures ARS was built to solve, the historical
  context, and how ARM-TS maps field-by-field to the annotated shell you already write.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(jsonlite)
library(glue)

# learnr global options
tutorial_options(
  exercise.timelimit = 60
)

# gradethis setup
gradethis_setup()

# Load pharmaverse ADaM datasets
data("adsl",  package = "pharmaverseadam")
data("adae",  package = "pharmaverseadam")
data("adtte", package = "pharmaverseadam")
data("adlb",  package = "pharmaverseadam")

# Pre-filter datasets used throughout the tutorial
# Note: pharmaverseadam adsl only has SAFFL; use ARM != "" as a proxy for randomised/ITT
adsl_safe <- adsl %>% filter(SAFFL == "Y")
adsl_itt  <- adsl %>% filter(!is.na(ARM) & ARM != "")   # randomised subjects
adae_teae <- adae %>% filter(TRTEMFL == "Y")

# Load sample ARM-TS JSON
re_json <- jsonlite::read_json("data/reporting_event.json")
```

---

## Chapter 1: Why ARS Exists

### 1.1 The Problem ARS Was Built to Solve

Every clinical trial generates hundreds of Tables, Figures, and Listings (TFLs). Each TFL is
produced by a programmer, reviewed by a statistician, checked by an independent programmer,
and eventually packaged into a Clinical Study Report (CSR) for submission to the FDA or EMA.

This process has worked for decades. It has also been broken in the same ways for decades.

### The Five Chronic Failures

**1. Static, unnavigable outputs**

TFLs are delivered as RTF or PDF files. A single submission may contain 300–800 tables. They
cannot be searched programmatically. A regulator who wants to understand how Table 14.2.1 was
derived must read the SAP narrative, find the relevant section, interpret the English text, and
mentally connect it to the code. Every step is manual.

**2. Results that are expensive to generate and impossible to reuse**

A TFL produced for one study cannot be reused for another study, even if the analysis is
identical. The numbers live in a formatted cell in an RTF file. There is no structured
representation of the result itself — only its formatted display.

**3. No machine-readable link between intent and execution**

The SAP describes the analysis in English. The programmer interprets that English and writes
code. The code produces results. The results are formatted into a table. At no point in this
chain does a machine verify that the execution matches the intent. The only check is a human
reviewer comparing the code to the SAP — a slow, error-prone, and expensive process.

**4. Unnecessary variation between sponsors**

Every sponsor, every CRO, every programming team develops their own annotation conventions,
shell formats, and metadata structures. There is no standard for what a "primary efficacy
analysis" looks like as a data artifact. Two submissions covering the same endpoint will be
structured completely differently.

**5. Limited regulatory use cases**

Regulators receive the same PDF tables as everyone else. They cannot ask: "Show me every
analysis using the ITT population." They cannot re-run an analysis to verify a result. They
cannot compare across studies. All of this is theoretically possible from the underlying data
but practically impossible from the delivered outputs.

---

### 1.2 The Historical Attempts

It is worth understanding that ARS is not the first attempt to solve this problem.

**ARM v1.0 (Analysis Results Metadata, circa 2012–2018)**

ARM described metadata about analysis displays and results at a high level — essentially, it
told you what analyses existed and what displays they appeared in. It did not define a formal
results data structure. It was adopted unevenly and never gained regulatory traction.

**Define-XML 2.x**

Define-XML provides metadata about datasets and variables. It is submission-required and
well-understood. But it describes the *input* data structure, not the analysis results. You can
read a Define-XML and understand what ADTTE contains, but you cannot learn what the KM estimate
for the primary endpoint was.

**The Gap ARM Left**

ARM told you "Analysis A0001 appears in Table 14.2.1." It did not tell you: what dataset, what
variable, what population filter, what statistical method, or what the result was. The formal
model for encoding all of that — machine-readably — did not exist until ARS.

---

### 1.3 The ARS Solution

CDISC formally launched the Analysis Results Standard (ARS) project in 2021. Version 1.0 was
released in April 2024.

ARS solves the problem with a two-layer model:

**Layer 1 — Analysis Results Metadata Technical Specification (ARM-TS)**

A machine-readable specification of *what was planned* and *how analyses were performed*. It encodes:

- Which analyses exist
- What datasets and variables they operate on
- What populations and data subsets they use
- What statistical methods they apply
- How results appear in displays
- Links back to the SAP

This is the formal replacement for the annotated shell. It encodes exactly what a lead
programmer writes in annotations — but in JSON, not English.

**Layer 2 — Analysis Results Dataset (ARD)**

A structured data format for storing the actual computed results. Each row in an ARD represents
one result value from one analysis — the KM estimate, the p-value, the mean, the hazard ratio —
with full context: which analysis, which population, which group, which operation.

Together, these two layers create a machine-navigable chain: **what was planned → what was
computed → what was displayed.**

---

### 1.4 The Regulatory Motivation

The FDA is actively pushing adoption. Three concrete signals:

**1. R Consortium Pilot 4 (fall 2025)**

The R Consortium Submissions Working Group successfully submitted ADaM datasets in Dataset-JSON
format to the FDA, demonstrating that the new JSON-based exchange format is viable for
regulatory submissions.

**2. FDA Request for Public Comment (April 2025)**

The FDA published a Federal Register notice requesting public comment on CDISC Dataset-JSON v1.1
as a replacement for SAS XPT as the standard exchange format. This is a direct signal that the
entire CDISC data exchange infrastructure — including ARS — is moving toward regulatory mandate.

**3. CDISC 360i (February 2025)**

CDISC launched the 360i initiative with the stated goal of "standards-driven automation from
study design through results." ARS is the results layer of that chain.

---

### 1.5 Where the Annotated Shell Fits

This is the most important conceptual bridge for you as a practitioner.

In your current workflow at a CRO:

1. A programmer lead receives the SAP and TLF shells
2. They open each shell and add annotations — e.g. *"AVAL from ADTTE where PARAMCD='OS' and MITTFL='Y', KM estimate at Week 24 per Section 9.2.3 of SAP"*
3. This annotated shell is reviewed, signed off, and becomes the traceability document for the study
4. The programmer uses this annotation to write code
5. The QC programmer uses this annotation to independently verify the output

The annotated shell is your company's proprietary format for encoding ARM-TS-equivalent
information. Every CRO has a version of it. None of them are interoperable. None of them are
machine-readable.

**ARS ARM-TS is the standardized, machine-readable version of your annotated shell.**

| Annotated Shell (Manual) | ARM-TS (Machine-Readable) |
|:-------------------------|:--------------------------|
| Table number and title | `Output.id`, `OutputDisplay.displayTitle` |
| "AVAL from ADTTE" | `Analysis.dataset = 'ADTTE'`, `Analysis.variable = 'AVAL'` |
| "where MITTFL='Y'" | `AnalysisSet` with `WhereClauseCondition: ADTTE.MITTFL EQ 'Y'` |
| "KM estimate" | `AnalysisMethod` with `Operation: KM estimate` |
| "per Section 9.2.3 of SAP" | `Analysis.documentRefs` pointing to the SAP document |
| Lead programmer's sign-off | `ReportingEvent` with version and name |

The logic is identical. The representation is different. Your existing knowledge of annotated
shells is directly transferable to ARM-TS — you just need to learn the formal object model
(Chapter 3).

---

### 1.6 What ARS Does NOT Cover

- ARS does not replace SDTM or ADaM. It operates on top of them.
- ARS does not specify *how* to perform statistical analyses. It specifies how to describe and record them.
- ARS does not replace the SAP. It encodes the analysis specification from the SAP in machine-readable form.
- ARS does not validate the statistical methods themselves.
- ARS does not currently cover all analysis types. Complex methods (mixed models, Bayesian analyses) require sponsor-defined extensions.

---

### 1.7 The Adoption Gap

Despite the clear need and regulatory momentum, ARS adoption as of 2026 remains low in industry:

1. The standard is new — v1.0 only released April 2024
2. The tooling is incomplete — generation tools exist (`{cards}`), exchange tools exist (`{datasetjson}`), but the traceability/validation layer is missing
3. Most programmers have never seen an ARM-TS file
4. Training resources are sparse

This is the gap you intend to fill — through building the `arsbridge` package and becoming one
of the few practitioners with deep working knowledge of the standard.

---

## Hands-On: The Data We Work With

### A Look at the Real Data

The exercises below use the same pharmaverse datasets used in CDISC pilot submissions. Let's
orient ourselves before we start.

```{r ch1-preview, exercise=FALSE, echo=TRUE}
# Real pharmaverse ADaM data — used in CDISC pilot submissions
# adsl: one row per subject (Subject-Level Analysis Dataset)
dplyr::glimpse(adsl)
```

### How many subjects are in each population?

```{r ch1-populations, exercise=FALSE, echo=TRUE}
# The two key analysis populations
cat("ITT Population (ITTFL='Y'):", nrow(adsl_itt), "subjects\n")
cat("Safety Population (SAFFL='Y'):", nrow(adsl_safe), "subjects\n")
```

---

## Exercise 1.1 — Explore the Scale of the Problem {data-progressive=FALSE}

The annotated shell problem becomes clear when you see how many distinct results a single
dataset implies. Let's count.

```{r ch1-ex1, exercise=TRUE, exercise.lines=8}
# How many distinct treatment arms are in the Safety Population?
# This tells you the minimum number of result columns any by-arm table must have.
# Use distinct() on TRT01A from adsl_safe

adsl_safe %>%
  ___(TRT01A)    # fill in: distinct
```

```{r ch1-ex1-hint-1}
# Use distinct() to get unique values of TRT01A
adsl_safe %>% distinct(TRT01A)
```

```{r ch1-ex1-solution}
adsl_safe %>% distinct(TRT01A)
# 3 treatment arms: Xanomeline High Dose, Xanomeline Low Dose, Placebo
# Each requires separate results — ARM-TS must formally define each group
```

```{r ch1-ex1-check}
grade_result(
  pass_if(~ nrow(.result) == 3,
          "Correct — 3 treatment arms. Each needs its own result row in the ARD, and each group must be formally defined in the ARM-TS AnalysisGrouping."),
  fail_if(~ TRUE,
          "Try: adsl_safe %>% distinct(TRT01A)")
)
```

---

## Exercise 1.2 — Count the Implied Analyses {data-progressive=FALSE}

A single demographics table (Table 14.1.1) with 10 rows × 3 treatment arms implies at least
30 individual result values. How many TEAEs are there across treatment arms — giving you a sense
of the scale of results in a typical AE table?

```{r ch1-ex2, exercise=TRUE, exercise.lines=10}
# Count the number of subjects with at least one TEAE, by treatment arm
# adae_teae already contains only treatment-emergent AEs (TRTEMFL='Y')
# Hint: count unique subjects per arm

adae_teae %>%
  ___(USUBJID, TRT01A) %>%      # keep one row per subject per arm
  ___(TRT01A) %>%                # count by arm
  ___(n_subjects = n())          # name the count column
```

```{r ch1-ex2-hint-1}
# Step 1: distinct(USUBJID, TRT01A) to deduplicate
# Step 2: group_by(TRT01A)
# Step 3: summarise(n_subjects = n())
```

```{r ch1-ex2-solution}
adae_teae %>%
  distinct(USUBJID, TRT01A) %>%
  group_by(TRT01A) %>%
  summarise(n_subjects = n())
# Each of these counts is one result value in an ARD —
# one operationId, one groupId, one rawValue
```

---

## Exercise 1.3 — The ARM-TS in Your Hands {data-progressive=FALSE}

The sample ARM-TS JSON (`reporting_event.json`) is already loaded as `re_json`. This is a
minimal but complete ReportingEvent — the root object of the ARS model.

Explore its top-level structure and find the name of this reporting event.

```{r ch1-ex3, exercise=TRUE, exercise.lines=8}
# re_json is an R list parsed from the ARM-TS JSON
# Print the name of this reporting event
# Hint: it's a top-level field called "name"

re_json$___
```

```{r ch1-ex3-hint-1}
# Access a named element of a list with $
# re_json$name
```

```{r ch1-ex3-solution}
re_json$name
# "CDISC Pilot Study CSR"
# This is the root object — everything else (analyses, methods, populations)
# hangs off this ReportingEvent
```

```{r ch1-ex3-check}
grade_result(
  pass_if(~ .result == "CDISC Pilot Study CSR",
          "Correct — you're reading a real ARM-TS object. Everything in Chapters 2–5 builds on this structure."),
  fail_if(~ TRUE,
          "Try: re_json$name")
)
```

---

## Chapter 1 Knowledge Check

```{r ch1-quiz, echo=FALSE}
quiz(
  caption = "Chapter 1 — Check Your Understanding",

  question(
    "What is the primary problem that ARS solves?",
    answer("SAS datasets are too large for FDA submissions"),
    answer(
      "Clinical trial results are static, non-reusable, and have no machine-readable link between analysis intent and execution",
      correct = TRUE
    ),
    answer("R cannot produce regulatory-grade tables without SAS"),
    answer("ADaM datasets are not standardized across sponsors"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "In your CRO annotated shell workflow, what ARS component does the shell annotation most directly correspond to?",
    answer("The ARD (Analysis Results Dataset)"),
    answer(
      "The ARM-TS (Analysis Results Metadata Technical Specification)",
      correct = TRUE
    ),
    answer("The Dataset-JSON exchange format"),
    answer("The SDTM domain"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which of the following is NOT covered by ARS?",
    answer("Encoding which dataset and variable an analysis uses"),
    answer("Linking an analysis to the SAP section that specifies it"),
    answer(
      "Specifying how to perform a mixed-model repeated-measures (MMRM) analysis",
      correct = TRUE
    ),
    answer("Recording the population filter used for an analysis"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "What is the key difference between the ARD and the ARM-TS?",
    answer(
      "ARM-TS is the machine-readable specification of what was planned; ARD stores the actual computed result values",
      correct = TRUE
    ),
    answer("ARD is used for SDTM data; ARM-TS is used for ADaM data"),
    answer("ARM-TS is for continuous analyses; ARD is for categorical analyses"),
    answer("They are two names for the same object"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "According to the chapter, why is ARS adoption still low in industry as of 2026?",
    answer("The FDA has not expressed interest in ARS"),
    answer("ARS only covers SDTM, not ADaM"),
    answer(
      "The standard is new (v1.0 released April 2024), tooling is incomplete, and training resources are sparse",
      correct = TRUE
    ),
    answer("SAS cannot generate ARM-TS JSON"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

---

## Chapter 1 Summary {.congratulations}

You now understand:

- The **five chronic failures** ARS was built to address: static outputs, non-reusable results,
  no machine-readable intent-to-execution link, sponsor variation, and limited regulatory utility
- The **two-layer ARS model**: ARM-TS (specification) + ARD (results)
- Why ARM v1.0 and Define-XML fell short of solving the problem
- How **ARM-TS is the machine-readable version of your annotated shell** — the mapping is direct and field-by-field
- The regulatory signals driving adoption (R Consortium Pilot 4, FDA RFI, CDISC 360i)
- Where the `arsbridge` gap sits: the traceability/validation layer is missing from the open-source ecosystem

**Next:** Chapter 2 — The ARS Data Model (the complete object hierarchy: ReportingEvent → Analysis → OperationResult)

