---
title: "CDISC Analysis Results Standard (ARS) — Interactive Training"
subtitle: "Chapter 4: ARD Deep Dive"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Chapter 4 covers the Analysis Results Dataset — the OperationResult object,
  the flat ARD format used by {cards}, the mapping between flat ARD and the
  formal ARS model, and the traceability gap that {arsbridge} will close. You
  will generate continuous, dichotomous, and Kaplan-Meier ARDs from real
  pharmaverse data.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
# cardx is optional but used for the KM example
suppressWarnings(suppressMessages({
  cardx_available <- requireNamespace("cardx", quietly = TRUE)
  if (cardx_available) library(cardx)
}))
library(jsonlite)
library(glue)

tutorial_options(exercise.timelimit = 60)
gradethis_setup()

# Load pharmaverse ADaM datasets
data("adsl",  package = "pharmaverseadam")
data("adae",  package = "pharmaverseadam")
data("adtte", package = "pharmaverseadam")
data("adlb",  package = "pharmaverseadam")

adsl_safe <- adsl %>% filter(SAFFL == "Y")
adsl_itt  <- adsl %>% filter(!is.na(ARM) & ARM != "")
adae_teae <- adae %>% filter(TRTEMFL == "Y")

# Pre-built analysis dataset for the AE exercise
analysis_ae <- adsl_safe %>%
  select(USUBJID, TRT01A) %>%
  left_join(
    adae_teae %>% distinct(USUBJID) %>% mutate(any_teae = 1L),
    by = "USUBJID"
  ) %>%
  mutate(any_teae = dplyr::coalesce(any_teae, 0L))

re_json <- jsonlite::read_json("data/reporting_event.json")
```

---

## Chapter 4: ARD Deep Dive

### 4.1 What the ARD Is

The **Analysis Results Dataset (ARD)** is the structured representation of the actual computed
results of an analysis. It answers the question: *"What number did this analysis produce?"*

In the formal ARS model, ARD lives **inside each `Analysis` object as the `results` slot** —
a list of `OperationResult` objects. In practice, the ARD is often extracted and stored as a
**flat tabular data frame** for use with tools like `{cards}` and `{gtsummary}`.

Think of it this way:

- **ARM-TS** answers *"what was planned and how?"*
- **ARD** answers *"what number did it produce?"*

Together, they encode the complete picture of an analysis.

---

### 4.2 The OperationResult Object

Each `OperationResult` represents the result of **one statistical operation** for **one
specific combination of groups**.

| Field | Description |
|-------|-------------|
| `operationId` | References the `Operation` that produced this result (e.g., "Mean") |
| `resultGroups` | Which group combination (one entry per grouping factor) |
| `rawValue` | The unformatted computed value (full precision) — stored as **string** |
| `formattedValue` | The value formatted per the operation's `resultPattern` — stored as **string** |

**`ResultGroup` fields** — one per grouping factor:

| Field | Description |
|-------|-------------|
| `groupingId` | References the `GroupingFactor` (e.g., Treatment Arm = "GF001") |
| `groupId` | References the pre-defined `Group` (e.g., Placebo = "G002") |
| `groupValue` | The data value, used for data-driven groupings |

---

### 4.3 Worked Example — AE Count Table Results

Continuing the running case study from Chapter 3: our analysis counts subjects with any
TEAE by treatment arm using method `M002` (Count + Percentage with Denominator).

**The analysis has:**

- Method `M002` with operations `O007` (Count), `O008` (Percentage), `O009` (Denominator)
- Grouping factor `GF001` (Treatment Arm) with groups `G001` (Drug XYZ 10mg) and `G002` (Placebo)
- `resultsByGroup: true` — separate results per arm

**The results (ARD) for this analysis:**

```json
{
  "results": [
    { "operationId": "O009", "resultGroups": [{"groupingId":"GF001","groupId":"G001"}],
      "rawValue": "45", "formattedValue": "45" },
    { "operationId": "O009", "resultGroups": [{"groupingId":"GF001","groupId":"G002"}],
      "rawValue": "44", "formattedValue": "44" },
    { "operationId": "O007", "resultGroups": [{"groupingId":"GF001","groupId":"G001"}],
      "rawValue": "38", "formattedValue": "38" },
    { "operationId": "O007", "resultGroups": [{"groupingId":"GF001","groupId":"G002"}],
      "rawValue": "30", "formattedValue": "30" },
    { "operationId": "O008", "resultGroups": [{"groupingId":"GF001","groupId":"G001"}],
      "rawValue": "84.44444", "formattedValue": "84.4" },
    { "operationId": "O008", "resultGroups": [{"groupingId":"GF001","groupId":"G002"}],
      "rawValue": "68.18182", "formattedValue": "68.2" }
  ]
}
```

**Reading any `OperationResult`** is a four-step lookup:

1. `operationId` → find the `Operation` → get the statistic name and `resultPattern`
2. `resultGroups[i].groupingId` → find the `GroupingFactor` → the dimension
3. `resultGroups[i].groupId` → find the `Group` → the specific level
4. `rawValue` / `formattedValue` → the actual number

So the first row says: *"The denominator (O009) for the Drug XYZ 10mg arm (G001 within GF001)
is 45 subjects."*

---

### 4.4 The Flat ARD Format (as used by `{cards}`)

The `{cards}` R package implements a **flat tabular representation** of the ARD. Rather than
nesting results inside Analysis objects, it produces a data frame where each row is one
result with full context columns.

**Standard ARD data-frame columns**

| Column | Description | Example |
|--------|-------------|---------|
| `group1` | First grouping variable name | `"TRT01A"` |
| `group1_level` | Value of first grouping variable | `"Xanomeline High Dose"` |
| `group2` / `group2_level` | Second grouping variable (if any) | `"SEX"` / `"M"` |
| `variable` | Analysis variable name | `"AGE"` |
| `variable_level` | For categorical: the level | `NA` or `"Yes"` |
| `stat_name` | Name of the statistic | `"mean"`, `"sd"`, `"n"` |
| `stat_label` | Display label for the stat | `"Mean"`, `"Std. Dev."` |
| `stat` | The computed value | `74.4` |
| `context` | Source/context string | `"continuous"` |
| `fmt_fn` | Formatting function | `"xx.x"` |
| `warning` / `error` | Any messages | `NULL` / `NULL` |

This is what `{gtsummary}` consumes to render tables. The mapping between the flat ARD and
the nested ARS model is handled by the tooling.

---

### 4.5 Flat ARD ↔ Formal ARS Model

Understanding how the flat ARD maps to the formal model is **critical for `arsbridge`**.

| Flat ARD column | ARS model path |
|----------------|----------------|
| `group1` | `GroupingFactor.groupingVariable` |
| `group1_level` | `Group.name` or `ResultGroup.groupValue` |
| `variable` | `Analysis.variable` |
| `stat_name` | `Operation.name` (lowercase) |
| `stat_label` | `Operation.name` (display form) |
| `stat` | `OperationResult.rawValue` |
| *(formatted)* | `OperationResult.formattedValue` |

The flat ARD **loses some ARS-model context** — specifically, it does **not** carry the
`analysisSetId`, `dataSubsetId`, `methodId`, or `documentRefs`. **This is the traceability gap**:
a flat ARD by itself cannot tell you which population was used or which SAP section it came
from. That information lives in the **ARM-TS**.

---

### 4.6 Generating an ARD with `{cards}` and `{cardx}`

`{cards}` covers descriptive statistics (continuous, categorical, dichotomous). `{cardx}`
extends it with t-tests, Wilcoxon, Kaplan-Meier estimates, and regression models.

```r
library(cards)

# Continuous summary
ard_age <- ard_continuous(
  data = adsl_safe,
  by = TRT01A,
  variables = AGE,
  statistic = list(AGE = continuous_summary_fns(
    c("N", "mean", "sd", "median", "min", "max")
  ))
)
```

```r
library(cardx)

# Kaplan-Meier estimates at fixed timepoints
ard_km <- ard_survival_survfit(
  data = adtte %>% filter(PARAMCD == "OS" & ITTFL == "Y"),
  by = TRT01A,
  variables = c(time = "AVAL", event = "1-CNSR"),
  times = c(6, 12, 24)
)
```

You'll generate all three (continuous, dichotomous, KM) in the exercises below.

---

### 4.7 What the ARD Cannot Tell You Alone

A flat ARD data frame contains the results. It does **not** contain:

- **Which population** was used (ITT? Safety?)
- **Which data subset** was applied (PARAMCD = OS? AVISIT = Week 24?)
- **What statistical method** was used (raw count? adjusted estimate?)
- **Where in the SAP** this analysis is specified
- **What display context** this belongs to (which table?)

To answer any of these, you need the **ARM-TS**. The ARM-TS is the context layer; the ARD is
the result layer. Only together do they form a complete, traceable record.

This is why **a validator that cross-checks ARD against ARM-TS** is the most useful tool that
currently doesn't exist — and is exactly what `arsbridge::validate_ard()` will build.

---

### 4.8 `rawValue` vs `formattedValue`

One important detail:

- `rawValue` = full-precision computed value, e.g. `"84.44444444"`
- `formattedValue` = display string, e.g. `"84.4"` (per the operation's `resultPattern`)

Both are stored as **strings** in the ARS model — not numerics. This is deliberate:

- It preserves the **exact display representation** (no surprise rounding on read)
- It keeps the underlying **precision available** for downstream use — e.g., a regulator
  verifying rounding logic

When building `arsbridge`:

- **Traceability validation** compares `rawValue` against a re-computed numeric result
- **Display validation** compares `formattedValue` against what `resultPattern` would produce
  from `rawValue`

---

## Hands-On: Generate Real ARDs

### A look at the analysis dataset

```{r ch4-preview, exercise=FALSE, echo=TRUE}
# adsl_safe = Safety Population (SAFFL='Y')
# adae_teae = treatment-emergent AEs only
cat("Safety Population subjects:", nrow(adsl_safe), "\n")
cat("TEAE records:               ", nrow(adae_teae), "\n")
cat("Subjects with any TEAE:     ", n_distinct(adae_teae$USUBJID), "\n")
```

```{r ch4-preview-ae, exercise=FALSE, echo=TRUE}
# Pre-built analysis dataset:
#   one row per subject in Safety Population
#   any_teae = 1 if subject had at least one TEAE, else 0
head(analysis_ae)
```

---

## Exercise 4.1 — Generate a Continuous ARD {data-progressive=FALSE}

Generate an ARD for **AGE** by treatment arm in the Safety Population using `ard_continuous()`.
Expected output: **6 statistics × 3 treatment arms = 18 rows**.

```{r ch4-ex1, exercise=TRUE, exercise.lines=12}
# Fill in: `by` = grouping variable, `variables` = analysis variable
ard_age <- ard_continuous(
  data      = adsl_safe,
  by        = ___,
  variables = ___,
  statistic = list(AGE = continuous_summary_fns(
    c("N", "mean", "sd", "median", "min", "max")
  ))
)

nrow(ard_age)
```

```{r ch4-ex1-hint-1}
# by = TRT01A   (the treatment arm in ADSL)
# variables = AGE
```

```{r ch4-ex1-solution}
ard_age <- ard_continuous(
  data      = adsl_safe,
  by        = TRT01A,
  variables = AGE,
  statistic = list(AGE = continuous_summary_fns(
    c("N", "mean", "sd", "median", "min", "max")
  ))
)
nrow(ard_age)
# 18 rows = 6 stats × 3 treatment arms.
# Each row maps to one ARS OperationResult.
```

---

## Exercise 4.2 — Pull a Specific Statistic Out of the ARD {data-progressive=FALSE}

The ARD is a tabular dataset — you can filter it like any other data frame. Pull the **mean
AGE for Xanomeline High Dose** out of `ard_age` (already built in your environment).

```{r ch4-ex2-setup, include=FALSE}
ard_age <- ard_continuous(
  data      = adsl_safe,
  by        = TRT01A,
  variables = AGE,
  statistic = list(AGE = continuous_summary_fns(
    c("N", "mean", "sd", "median", "min", "max")
  ))
)
```

```{r ch4-ex2, exercise=TRUE, exercise.lines=10, exercise.setup="ch4-ex2-setup"}
# Filter ard_age to the Mean for Xanomeline High Dose
ard_age %>%
  filter(group1_level == ___,
         stat_name    == ___) %>%      # the stat name as it appears in cards
  pull(stat)
```

```{r ch4-ex2-hint-1}
# group1_level: "Xanomeline High Dose"
# stat_name:    "mean"  (lowercase, as cards stores it)
```

```{r ch4-ex2-solution}
ard_age %>%
  filter(group1_level == "Xanomeline High Dose",
         stat_name    == "mean") %>%
  pull(stat)
# A single numeric — the same value that, in the ARS model,
# would be stored as OperationResult.rawValue (as a string).
```

---

## Exercise 4.3 — Generate a Dichotomous ARD for TEAE {data-progressive=FALSE}

Goal: count subjects with any TEAE by treatment arm. The analysis dataset `analysis_ae` is
already built (one row per safety-population subject, `any_teae` = 0/1). Use `ard_dichotomous()`
and then pull the **percentage** for Xanomeline High Dose.

```{r ch4-ex3, exercise=TRUE, exercise.lines=18}
ard_ae <- ard_dichotomous(
  data      = analysis_ae,
  by        = ___,                        # grouping variable
  variables = ___,                        # the 0/1 flag
  value     = list(any_teae = 1L)         # the "positive" value
)

# Pull the 'p' (percentage) row for Xanomeline High Dose
ard_ae %>%
  filter(group1_level == ___,
         stat_name    == "p") %>%
  pull(stat)
```

```{r ch4-ex3-hint-1}
# by = TRT01A, variables = any_teae
# group1_level = "Xanomeline High Dose"
```

```{r ch4-ex3-solution}
ard_ae <- ard_dichotomous(
  data      = analysis_ae,
  by        = TRT01A,
  variables = any_teae,
  value     = list(any_teae = 1L)
)
ard_ae %>%
  filter(group1_level == "Xanomeline High Dose",
         stat_name    == "p") %>%
  pull(stat)
# This single percentage is one OperationResult.rawValue.
# In the ARM-TS, the corresponding Operation has resultPattern "XX.X"
# which determines the formattedValue.
```

---

## Exercise 4.4 — Bonus: A Kaplan-Meier ARD with `{cardx}` {data-progressive=FALSE}

Generate a KM ARD for time-to-first-AE from `adtte`, ITT population. (This exercise is
skipped silently if `{cardx}` isn't installed.)

```{r ch4-ex4, exercise=TRUE, exercise.lines=20}
if (!requireNamespace("cardx", quietly = TRUE)) {
  cat("Install cardx to run this exercise: install.packages('cardx')\n")
} else {
  # Pick the first available PARAMCD as the endpoint
  param <- adtte$PARAMCD[1]

  adtte_endpoint <- adtte %>%
    filter(PARAMCD == ___ & ITTFL == "Y")     # which PARAMCD?

  ard_km <- cardx::ard_survival_survfit(
    data      = adtte_endpoint,
    by        = TRT01A,
    variables = c(time = "AVAL", event = "1-CNSR"),
    times     = c(6, 12, 18, 24)
  )

  cat("KM ARD rows:", nrow(ard_km), "\n")
  head(ard_km, 6)
}
```

```{r ch4-ex4-hint-1}
# Use the variable `param` that you just defined,
# i.e. filter(PARAMCD == param & ITTFL == "Y")
# In a real analysis you would name the endpoint explicitly, e.g. "OS".
```

```{r ch4-ex4-solution}
if (!requireNamespace("cardx", quietly = TRUE)) {
  cat("Install cardx to run this exercise.\n")
} else {
  param <- adtte$PARAMCD[1]

  adtte_endpoint <- adtte %>%
    filter(PARAMCD == param & ITTFL == "Y")

  ard_km <- cardx::ard_survival_survfit(
    data      = adtte_endpoint,
    by        = TRT01A,
    variables = c(time = "AVAL", event = "1-CNSR"),
    times     = c(6, 12, 18, 24)
  )
  cat("KM ARD rows:", nrow(ard_km), "\n")
  head(ard_km, 6)
}
# Each row is one OperationResult: a KM estimate at one timepoint
# for one treatment arm.
```

---

## Exercise 4.5 — Map Flat ARD Columns to the Formal Model {data-progressive=FALSE}

This is the conceptual capstone of Chapter 4. Build a small lookup mapping each flat-ARD
column to its formal-ARS-model path. This mapping is the core knowledge `arsbridge` needs.

```{r ch4-ex5, exercise=TRUE, exercise.lines=18}
mappings <- tribble(
  ~flat_ard_column,   ~ars_model_path,
  "group1",           ___,
  "group1_level",     ___,
  "variable",         ___,
  "stat_name",        ___,
  "stat",             ___
)

mappings
```

```{r ch4-ex5-hint-1}
# Choose from these formal-model paths:
#   "GroupingFactor.groupingVariable"
#   "Group.name (or ResultGroup.groupValue)"
#   "Analysis.variable"
#   "Operation.name"
#   "OperationResult.rawValue"
```

```{r ch4-ex5-solution}
mappings <- tribble(
  ~flat_ard_column,   ~ars_model_path,
  "group1",           "GroupingFactor.groupingVariable",
  "group1_level",     "Group.name (or ResultGroup.groupValue)",
  "variable",         "Analysis.variable",
  "stat_name",        "Operation.name",
  "stat",             "OperationResult.rawValue"
)
mappings
# Note what's MISSING from the flat ARD:
#   analysisSetId, dataSubsetId, methodId, documentRefs.
# Those live only in the ARM-TS — and that's the traceability gap.
```

---

## Chapter 4 Knowledge Check

```{r ch4-quiz, echo=FALSE}
quiz(
  caption = "Chapter 4 — Check Your Understanding",

  question(
    "Where, in the formal ARS model, does the ARD physically live?",
    answer("As a separate top-level field of ReportingEvent"),
    answer("Inside Analysis.results as a list of OperationResult objects", correct = TRUE),
    answer("In a dedicated ARD file referenced by Output.fileSpecifications"),
    answer("Inside AnalysisMethod.codeTemplate"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Each OperationResult represents…",
    answer(
      "one statistical operation's result for one specific combination of groups",
      correct = TRUE
    ),
    answer("the complete output for one display table"),
    answer("all results for a given AnalysisSet"),
    answer("a single subject's value on the analysis variable"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Why are both rawValue and formattedValue stored as STRINGS in the ARS model?",
    answer(
      "To preserve the exact display representation AND retain full underlying precision — both matter for downstream validation",
      correct = TRUE
    ),
    answer("Because JSON cannot encode numeric values"),
    answer("To force every consumer to re-parse the values"),
    answer("Because R only supports character vectors"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which ARS-model field does the flat ARD column `stat` correspond to?",
    answer("Operation.name"),
    answer("OperationResult.rawValue", correct = TRUE),
    answer("Analysis.variable"),
    answer("Group.name"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "What is the 'traceability gap' that arsbridge is meant to close?",
    answer(
      "The flat ARD alone does NOT carry analysisSetId, dataSubsetId, methodId, or documentRefs — that context lives only in the ARM-TS",
      correct = TRUE
    ),
    answer("ADaM datasets cannot be linked to SDTM"),
    answer("R cannot read SAS XPT files"),
    answer("Define-XML does not describe ADaM"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which R package generates the flat ARD data frame?",
    answer("{gtsummary}"),
    answer("{cards}", correct = TRUE),
    answer("{datasetjson}"),
    answer("{admiral}"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which package extends {cards} with KM estimates, regression, and inferential statistics?",
    answer("{cardx}", correct = TRUE),
    answer("{rtables}"),
    answer("{tfrmt}"),
    answer("{admiral}"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "An OperationResult has rawValue = '84.44444' and formattedValue = '84.4'. What can you infer?",
    answer(
      "The operation's resultPattern rounds to one decimal place (e.g., 'XX.X')",
      correct = TRUE
    ),
    answer("The analysis produced two different statistics"),
    answer("The result failed validation"),
    answer("Both values are wrong — they must be identical"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

---

## Chapter 4 Summary {.congratulations}

You now understand:

- The **ARD is the result layer**: `Analysis.results` → `[OperationResult]`
- Each `OperationResult` stores **one statistic for one group combination**, with both
  `rawValue` (full precision) and `formattedValue` (display string) as strings
- The **flat ARD format** (used by `{cards}`) is a tabular extraction of this data — each row
  is one result with grouping/variable/stat columns
- The flat ARD alone is **not sufficient for traceability** — `analysisSetId`, `dataSubsetId`,
  `methodId`, and `documentRefs` live only in the ARM-TS
- Generating ARDs in R uses **`{cards}`** for descriptive statistics and **`{cardx}`** for KM
  estimates, t-tests, and regression
- You've generated **continuous**, **dichotomous**, and **Kaplan-Meier** ARDs from real
  pharmaverse data

**Next:** Chapter 5 — The Traceability Chain (re-derive a stored ARD result from ARM-TS +
ADaM and find a deliberate traceability break)
