---
title: "CDISC Analysis Results Standard (ARS): Interactive Training"
subtitle: "Chapter 4: ARD Deep Dive"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Chapter 4 covers the Analysis Results Dataset: the OperationResult object,
  the flat ARD format used by {cards}, the mapping between flat ARD and the
  formal ARS model, and the traceability gap that {arsbridge} will close. You
  will generate continuous, dichotomous, and Kaplan-Meier ARDs from real
  pharmaverse data.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
# cardx is optional but used for the KM example
suppressWarnings(suppressMessages({
  cardx_available <- requireNamespace("cardx", quietly = TRUE)
  if (cardx_available) library(cardx)
}))
library(jsonlite)
library(glue)

tutorial_options(exercise.timelimit = 60)
gradethis_setup()

# Load pharmaverse ADaM datasets
data("adsl",  package = "pharmaverseadam")
data("adae",  package = "pharmaverseadam")
# pharmaverseadam 1.4+ replaced the generic "adtte" with TA-specific
# datasets; use the oncology one and join subject-level vars in from
# ADSL. pharmaverseadam ADSL lacks ITTFL, so derive it from ARM (a
# subject is "randomised / ITT" iff ARM is non-empty).
data("adtte_onco", package = "pharmaverseadam")
adtte <- adtte_onco %>%
  left_join(
    adsl %>%
      mutate(ITTFL = ifelse(!is.na(ARM) & ARM != "", "Y", "N")) %>%
      select(USUBJID, TRT01A, ITTFL),
    by = "USUBJID"
  )
data("adlb",  package = "pharmaverseadam")

adsl_safe <- adsl %>% filter(SAFFL == "Y")
adsl_itt  <- adsl %>% filter(!is.na(ARM) & ARM != "")
adae_teae <- adae %>% filter(TRTEMFL == "Y")

# Pre-built analysis dataset for the AE exercise
analysis_ae <- adsl_safe %>%
  select(USUBJID, TRT01A) %>%
  left_join(
    adae_teae %>% distinct(USUBJID) %>% mutate(any_teae = 1L),
    by = "USUBJID"
  ) %>%
  mutate(any_teae = dplyr::coalesce(any_teae, 0L))

re_json <- jsonlite::read_json("data/reporting_event.json")
```

---

## Chapter 4: ARD Deep Dive

### 4.1 What the ARD Is

The **Analysis Results Dataset (ARD)** is the structured representation of the actual
computed results of an analysis. It answers one question: *"What number did this analysis
produce?"*

In the formal ARS model, the ARD lives **inside each `Analysis` object as the `results`
slot**, a list of `OperationResult` objects. In practice, the ARD is often extracted and
stored as a **flat tabular data frame** for use with tools like `{cards}` and
`{gtsummary}`. This chapter covers both forms and, crucially, how they map to each other.

Think of the division of labor this way:

- **ARM-TS** answers *"what was planned and how?"*
- **ARD** answers *"what number did it produce?"*

Together, they encode the complete picture of an analysis. Apart, each has a blind spot,
and Section 4.7 shows exactly where the ARD's blind spot is.

---

### 4.2 The OperationResult Object

Each `OperationResult` represents the result of **one statistical operation** for **one
specific combination of groups**. One number, fully addressed.

| Field | Description |
|-------|-------------|
| `operationId` | References the `Operation` that produced this result (e.g., "Mean") |
| `resultGroups` | Which group combination (one entry per grouping factor) |
| `rawValue` | The unformatted computed value (full precision), stored as a **string** |
| `formattedValue` | The value formatted per the operation's `resultPattern`, stored as a **string** |

**`ResultGroup` fields**, one per grouping factor:

| Field | Description |
|-------|-------------|
| `groupingId` | References the `GroupingFactor` (e.g., Treatment Arm = "GF001") |
| `groupId` | References the pre-defined `Group` (e.g., Placebo = "G002") |
| `groupValue` | The data value, used for data-driven groupings |

Notice that every field except the two values is a reference. An OperationResult never
repeats what the ARM-TS already knows; it just points at it. That is Chapter 2's
Principle 1 showing up again.

---

### 4.3 Worked Example: AE Count Table Results

Continuing the running case study from Chapter 3: our analysis counts subjects with any
TEAE by treatment arm, using method `M002` (Count + Percentage with Denominator).

**The analysis has:**

- Method `M002` with operations `O007` (Count), `O008` (Percentage), `O009` (Denominator)
- Grouping factor `GF001` (Treatment Arm) with groups `G001` (Drug XYZ 10mg) and `G002` (Placebo)
- `resultsByGroup: true`, meaning separate results per arm

**The results (ARD) for this analysis:**

```json
{
  "results": [
    { "operationId": "O009", "resultGroups": [{"groupingId":"GF001","groupId":"G001"}],
      "rawValue": "45", "formattedValue": "45" },
    { "operationId": "O009", "resultGroups": [{"groupingId":"GF001","groupId":"G002"}],
      "rawValue": "44", "formattedValue": "44" },
    { "operationId": "O007", "resultGroups": [{"groupingId":"GF001","groupId":"G001"}],
      "rawValue": "38", "formattedValue": "38" },
    { "operationId": "O007", "resultGroups": [{"groupingId":"GF001","groupId":"G002"}],
      "rawValue": "30", "formattedValue": "30" },
    { "operationId": "O008", "resultGroups": [{"groupingId":"GF001","groupId":"G001"}],
      "rawValue": "84.44444", "formattedValue": "84.4" },
    { "operationId": "O008", "resultGroups": [{"groupingId":"GF001","groupId":"G002"}],
      "rawValue": "68.18182", "formattedValue": "68.2" }
  ]
}
```

**Reading any `OperationResult`** is a four-step lookup:

1. `operationId` finds the `Operation`, giving you the statistic name and `resultPattern`
2. `resultGroups[i].groupingId` finds the `GroupingFactor`, giving you the dimension
3. `resultGroups[i].groupId` finds the `Group`, giving you the specific level
4. `rawValue` / `formattedValue` give you the actual number

So the first row says: *"The denominator (O009) for the Drug XYZ 10mg arm (G001 within
GF001) is 45 subjects."* Practice reading two or three rows this way and it becomes
automatic.

---

### 4.4 The Flat ARD Format (as used by `{cards}`)

The `{cards}` R package implements a **flat tabular representation** of the ARD. Rather
than nesting results inside Analysis objects, it produces a data frame where each row is
one result with full context columns.

**Standard ARD data-frame columns**

| Column | Description | Example |
|--------|-------------|---------|
| `group1` | First grouping variable name | `"TRT01A"` |
| `group1_level` | Value of first grouping variable | `"Xanomeline High Dose"` |
| `group2` / `group2_level` | Second grouping variable (if any) | `"SEX"` / `"M"` |
| `variable` | Analysis variable name | `"AGE"` |
| `variable_level` | For categorical: the level | `NA` or `"Yes"` |
| `stat_name` | Name of the statistic | `"mean"`, `"sd"`, `"n"` |
| `stat_label` | Display label for the stat | `"Mean"`, `"Std. Dev."` |
| `stat` | The computed value | `74.4` |
| `context` | Source/context string | `"continuous"` |
| `fmt_fn` | Formatting function | `"xx.x"` |
| `warning` / `error` | Any messages | `NULL` / `NULL` |

One convention to burn in now, because you will filter on it constantly: **`stat_name`
values are lowercase machine names** (`"n"`, `"mean"`, `"sd"`, `"p"`), while `stat_label`
holds the display form (`"Mean"`, `"Std. Dev."`). When you filter an ARD, you filter on
`stat_name`.

This flat format is what `{gtsummary}` consumes to render tables. The mapping between the
flat ARD and the nested ARS model is handled by tooling, and Section 4.5 spells that
mapping out.

---

### 4.5 Flat ARD to Formal ARS Model, and Back

Understanding how the flat ARD maps to the formal model is **critical for `arsbridge`**.

| Flat ARD column | ARS model path |
|----------------|----------------|
| `group1` | `GroupingFactor.groupingVariable` |
| `group1_level` | `Group.name` or `ResultGroup.groupValue` |
| `variable` | `Analysis.variable` |
| `stat_name` | `Operation.name` (lowercase) |
| `stat_label` | `Operation.name` (display form) |
| `stat` | `OperationResult.rawValue` |
| *(formatted)* | `OperationResult.formattedValue` |

Now look at what is *not* in that table. The flat ARD **loses ARS-model context**:
it does **not** carry the `analysisSetId`, `dataSubsetId`, `methodId`, or `documentRefs`.
**This is the traceability gap.** A flat ARD by itself cannot tell you which population
was used or which SAP section it came from. That information lives only in the ARM-TS.

---

### 4.6 Generating an ARD with `{cards}` and `{cardx}`

`{cards}` covers descriptive statistics (continuous, categorical, dichotomous). `{cardx}`
extends it with t-tests, Wilcoxon, Kaplan-Meier estimates, and regression models.

The workhorse functions share one calling pattern, so learn it once:

- `data`: the analysis dataset (already filtered to the right population)
- `by`: the grouping variable, unquoted (e.g., `TRT01A`). This becomes `group1`.
- `variables`: the analysis variable(s), unquoted (e.g., `AGE`)
- `statistic`: which statistics to compute, per variable

**Continuous summary.** `continuous_summary_fns()` is a helper that expands a character
vector of statistic names into the functions `{cards}` should run:

```r
library(cards)

ard_age <- ard_continuous(
  data = adsl_safe,
  by = TRT01A,
  variables = AGE,
  statistic = list(AGE = continuous_summary_fns(
    c("N", "mean", "sd", "median", "min", "max")
  ))
)
```

**Dichotomous (yes/no) outcomes.** `ard_dichotomous()` handles 0/1 flags. Its extra
argument `value` declares which value counts as the event (here, `any_teae = 1L`). It
produces three statistics per group: `n` (count with the event), `N` (group total, the
denominator), and `p` (the proportion, `n/N`). Those are exactly the Count, Denominator,
and Percentage operations of method `M002` from Section 4.3.

```r
ard_ae <- ard_dichotomous(
  data      = analysis_ae,
  by        = TRT01A,
  variables = any_teae,
  value     = list(any_teae = 1L)
)
```

**Kaplan-Meier estimates at fixed timepoints.** In ADaM time-to-event data, `AVAL` is the
time and `CNSR` is the censoring flag (1 = censored, 0 = event). Survival tooling wants an
*event* indicator, so the convention is `event = "1-CNSR"`, which flips the flag. `times`
asks for survival estimates at specific timepoints:

```r
library(cardx)

ard_km <- ard_survival_survfit(
  data = adtte %>% filter(PARAMCD == "OS" & ITTFL == "Y"),
  by = TRT01A,
  variables = c(time = "AVAL", event = "1-CNSR"),
  times = c(6, 12, 24)
)
```

You will generate all three (continuous, dichotomous, KM) in the exercises below.

---

### 4.7 What the ARD Cannot Tell You Alone

A flat ARD data frame contains the results. It does **not** contain:

- **Which population** was used (ITT? Safety?)
- **Which data subset** was applied (PARAMCD = OS? AVISIT = Week 24?)
- **What statistical method** was used (raw count? adjusted estimate?)
- **Where in the SAP** this analysis is specified
- **What display context** this belongs to (which table?)

To answer any of these, you need the ARM-TS. The ARM-TS is the context layer; the ARD is
the result layer. Only together do they form a complete, traceable record.

This is why **a validator that cross-checks the ARD against the ARM-TS** is the most
useful tool that currently does not exist, and it is exactly what
`arsbridge::validate_ard()` will build.

---

### 4.8 `rawValue` vs `formattedValue`

One important detail before the hands-on work:

- `rawValue` = full-precision computed value, e.g. `"84.44444444"`
- `formattedValue` = display string, e.g. `"84.4"` (per the operation's `resultPattern`)

Both are stored as **strings** in the ARS model, not numerics. This is deliberate, and it
serves two audiences at once:

- It preserves the **exact display representation**, so nothing silently re-rounds on read
- It keeps the underlying **precision available** for downstream use, for example a
  regulator verifying rounding logic

When building `arsbridge`:

- **Traceability validation** compares `rawValue` against a re-computed numeric result
- **Display validation** compares `formattedValue` against what `resultPattern` would
  produce from `rawValue`

---

## Hands-On: Generate Real ARDs

### A look at the analysis dataset

Two objects matter for the exercises. `adsl_safe` you know from earlier chapters. The new
one is `analysis_ae`: a pre-built analysis dataset with one row per Safety-population
subject and a flag `any_teae` set to 1 if the subject had at least one TEAE, else 0. This
subject-level 0/1 shape is exactly what a dichotomous analysis needs.

```{r ch4-preview, exercise=FALSE, echo=TRUE}
# adsl_safe = Safety Population (SAFFL='Y')
# adae_teae = treatment-emergent AEs only
cat("Safety Population subjects:", nrow(adsl_safe), "\n")
cat("TEAE records:               ", nrow(adae_teae), "\n")
cat("Subjects with any TEAE:     ", n_distinct(adae_teae$USUBJID), "\n")
```

```{r ch4-preview-ae, exercise=FALSE, echo=TRUE}
# Pre-built analysis dataset:
#   one row per subject in Safety Population
#   any_teae = 1 if subject had at least one TEAE, else 0
head(analysis_ae)
```

---

## Exercise 4.1: Generate a Continuous ARD {data-progressive=FALSE}

**Step 1: the annotated shell.**

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; text-align:center;">Table 14.1.1: Demographic Characteristics</div>
<div style="text-align:center; font-family:sans-serif; margin-bottom:8px;">Safety Population &nbsp;(<span style="color:#b3400e; font-weight:bold;">ADSL.SAFFL='Y'</span>)</div>
<table style="border-collapse:collapse; width:100%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left; width:44%;"></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Xanomeline High Dose<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Xanomeline Low Dose<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Placebo<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX</span></th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">Age (years) &nbsp;<span style="color:#b3400e; font-weight:bold;">ADSL.AGE</span></td>
  <td style="border:1px solid #bbb; padding:3px 10px;"></td>
  <td style="border:1px solid #bbb; padding:3px 10px;"></td>
  <td style="border:1px solid #bbb; padding:3px 10px;"></td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;n</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx</td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;Mean (SD)</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x (xxx.xx)</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x (xxx.xx)</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x (xxx.xx)</td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;Median</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx.x</td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;Min, Max</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx, xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx, xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx, xxx</td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">Grouping: <span style="color:#b3400e;">TRT01A</span> &nbsp;|&nbsp; Source: <span style="color:#b3400e;">ADSL.AGE</span> &nbsp;|&nbsp; 6 stats × 3 arms = 18 ARD rows</div>
</div>

**Step 2: the ARM-TS encoding.** The spec side you already know from Chapter 3:

```json
{ "id": "AN001", "dataset": "ADSL", "variable": "AGE",
  "analysisSetId": "AS002",
  "methodId": "M001",
  "results": [ "<-- 18 OperationResults will land here" ] }
```

**Step 3: your turn.** Produce the numbers that fill that `results` slot. Generate the ARD
with `ard_continuous()`; the calling pattern is the one from Section 4.6: `by` takes the
grouping variable, `variables` takes the analysis variable, both unquoted.

Before you run it, predict the row count: 6 statistics, 3 treatment arms, one row per
statistic per arm. That is **18 rows**, and each of them corresponds to one
`OperationResult` in the formal model.

```{r ch4-ex1, exercise=TRUE, exercise.lines=12}
# Fill in: `by` = grouping variable, `variables` = analysis variable
ard_age <- ard_continuous(
  data      = adsl_safe,
  by        = ___,
  variables = ___,
  statistic = list(AGE = continuous_summary_fns(
    c("N", "mean", "sd", "median", "min", "max")
  ))
)

nrow(ard_age)
```

```{r ch4-ex1-hint-1}
# by = TRT01A   (the treatment arm in ADSL)
# variables = AGE
```

```{r ch4-ex1-solution}
ard_age <- ard_continuous(
  data      = adsl_safe,
  by        = TRT01A,
  variables = AGE,
  statistic = list(AGE = continuous_summary_fns(
    c("N", "mean", "sd", "median", "min", "max")
  ))
)
nrow(ard_age)
# 18 rows = 6 stats x 3 treatment arms.
# Each row maps to one ARS OperationResult.
```

---

## Exercise 4.2: Pull a Specific Statistic Out of the ARD {data-progressive=FALSE}

**Step 1: the annotated shell.** The one cell you want:

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; text-align:center;">Table 14.1.1: Demographic Characteristics [excerpt]</div>
<div style="text-align:center; font-family:sans-serif; margin-bottom:8px;">Safety Population &nbsp;(<span style="color:#b3400e; font-weight:bold;">ADSL.SAFFL='Y'</span>)</div>
<table style="border-collapse:collapse; width:100%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left; width:44%;"></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Xanomeline High Dose &nbsp;<span style="color:#b3400e; font-size:0.82em; font-weight:normal;">ADSL.TRT01A='Xanomeline High Dose'</span></th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">Mean Age &nbsp;<span style="color:#b3400e; font-weight:bold;">ADSL.AGE</span></td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;"><strong>xx.x</strong></td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">This single cell = one ARD row: operationId=O002, groupId=G001, rawValue=?, formattedValue=?</div>
</div>

**Step 2: the ARM-TS encoding.** In the formal model that cell is one OperationResult:

```json
{ "operationId": "O002",
  "resultGroups": [ { "groupId": "G001" } ],
  "rawValue": "..." }
```

In the flat ARD, the same result is one row where `group1_level` plays the role of
`groupId` and `stat_name` plays the role of `operationId` (the Section 4.5 mapping).

**Step 3: your turn.** The ARD is a tabular dataset, so you can filter it like any other
data frame. That is the whole point of the flat format: results become queryable.

Pull the **mean AGE for Xanomeline High Dose** out of `ard_age` (already built in your
environment). Remember from Section 4.4 that you filter on `stat_name` using the lowercase
machine name, so the mean is `"mean"`. The final `pull(stat)` extracts the value column as
a plain vector instead of a one-row data frame.

```{r ch4-ex2-setup, include=FALSE}
ard_age <- ard_continuous(
  data      = adsl_safe,
  by        = TRT01A,
  variables = AGE,
  statistic = list(AGE = continuous_summary_fns(
    c("N", "mean", "sd", "median", "min", "max")
  ))
)
```

```{r ch4-ex2, exercise=TRUE, exercise.lines=10, exercise.setup="ch4-ex2-setup"}
# Filter ard_age to the Mean for Xanomeline High Dose
ard_age %>%
  filter(group1_level == ___,
         stat_name    == ___) %>%      # the stat name as it appears in cards
  pull(stat)
```

```{r ch4-ex2-hint-1}
# group1_level: "Xanomeline High Dose"
# stat_name:    "mean"  (lowercase, as cards stores it)
```

```{r ch4-ex2-solution}
ard_age %>%
  filter(group1_level == "Xanomeline High Dose",
         stat_name    == "mean") %>%
  pull(stat)
# A single numeric: the same value that, in the ARS model,
# would be stored as OperationResult.rawValue (as a string).
```

---

## Exercise 4.3: Generate a Dichotomous ARD for TEAE {data-progressive=FALSE}

**Step 1: the annotated shell.**

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; text-align:center;">Table 14.3.1: Summary of Treatment-Emergent Adverse Events</div>
<div style="text-align:center; font-family:sans-serif; margin-bottom:8px;">Safety Population &nbsp;(<span style="color:#b3400e; font-weight:bold;">ADSL.SAFFL='Y'</span>)</div>
<table style="border-collapse:collapse; width:100%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left; width:44%;"></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Xanomeline High Dose<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX n (%)</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Xanomeline Low Dose<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX n (%)</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Placebo<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX n (%)</span></th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">Subjects with any TEAE &nbsp;<span style="color:#b3400e; font-weight:bold;">ADAE.USUBJID</span><br><span style="color:#b3400e; font-size:0.88em;">ADAE.TRTEMFL='Y'</span></td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx (xx.x)</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx (xx.x)</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">xxx (xx.x)</td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">Grouping: <span style="color:#b3400e;">TRT01A</span> &nbsp;|&nbsp; Format: <span style="color:#b3400e;">XXX (XX.X)</span> &nbsp;|&nbsp; Denominator: N safety subjects per arm</div>
</div>

**Step 2: the ARM-TS encoding.** The "n (%)" annotation is method `M002` with its three
operations, and the "(%)" carries a formal denominator reference:

```json
{ "id": "M002", "name": "Count and Percentage",
  "operations": [
    { "id": "O007", "name": "Count",       "resultPattern": "XXX"  },
    { "id": "O008", "name": "Percentage",  "resultPattern": "XX.X" },
    { "id": "O009", "name": "Denominator", "resultPattern": "XXX"  }
  ] }
```

**Step 3: your turn.** Compute those three statistics per arm. The analysis dataset
`analysis_ae` is already built (one row per Safety-population subject, `any_teae` = 0/1).

Recall from Section 4.6 how `ard_dichotomous()` works: `value = list(any_teae = 1L)`
declares that 1 is the event, and the output carries `n` (count), `N` (denominator), and
`p` (proportion) per arm. Those map one-to-one onto O007 / O009 / O008 above.

Generate the ARD, then pull the **`p` (percentage) row** for Xanomeline High Dose.

```{r ch4-ex3, exercise=TRUE, exercise.lines=18}
ard_ae <- ard_dichotomous(
  data      = analysis_ae,
  by        = ___,                        # grouping variable
  variables = ___,                        # the 0/1 flag
  value     = list(any_teae = 1L)         # the "positive" value
)

# Pull the 'p' (percentage) row for Xanomeline High Dose
ard_ae %>%
  filter(group1_level == ___,
         stat_name    == "p") %>%
  pull(stat)
```

```{r ch4-ex3-hint-1}
# by = TRT01A, variables = any_teae
# group1_level = "Xanomeline High Dose"
```

```{r ch4-ex3-solution}
ard_ae <- ard_dichotomous(
  data      = analysis_ae,
  by        = TRT01A,
  variables = any_teae,
  value     = list(any_teae = 1L)
)
ard_ae %>%
  filter(group1_level == "Xanomeline High Dose",
         stat_name    == "p") %>%
  pull(stat)
# This single percentage is one OperationResult.rawValue.
# In the ARM-TS, the corresponding Operation has resultPattern "XX.X"
# which determines the formattedValue.
```

---

## Exercise 4.4: Bonus, a Kaplan-Meier ARD with `{cardx}` {data-progressive=FALSE}

**Step 1: the annotated shell.**

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; text-align:center;">Table 14.2.1: Kaplan-Meier Estimates of Survival</div>
<div style="text-align:center; font-family:sans-serif; margin-bottom:8px;">Intent-to-Treat Population &nbsp;(<span style="color:#b3400e; font-weight:bold;">ADSL.ITTFL='Y'</span>)</div>
<table style="border-collapse:collapse; width:100%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left; width:44%;"></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Treatment A<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Placebo<br><span style="font-weight:normal; font-size:0.88em; color:#555;">N=XXX</span></th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">Survival probability &nbsp;<span style="color:#b3400e; font-weight:bold;">ADTTE.AVAL</span> &nbsp;<span style="color:#b3400e; font-size:0.88em;">ADTTE.CNSR</span><br><span style="color:#b3400e; font-size:0.85em;">ADTTE.PARAMCD='OS'</span></td>
  <td style="border:1px solid #bbb; padding:3px 10px;"></td>
  <td style="border:1px solid #bbb; padding:3px 10px;"></td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;Month 6</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">x.xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">x.xxx</td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;Month 12</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">x.xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">x.xxx</td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;Month 18</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">x.xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">x.xxx</td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">&nbsp;&nbsp;Month 24</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">x.xxx</td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center;">x.xxx</td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">Grouping: <span style="color:#b3400e;">TRT01A</span> &nbsp;|&nbsp; DataSubset: <span style="color:#b3400e;">ADTTE.PARAMCD='OS'</span> &nbsp;|&nbsp; Format: <span style="color:#b3400e;">X.XXX</span></div>
</div>

**Step 2: the ARM-TS encoding.** One operation per requested timepoint:

```json
{ "id": "M003", "name": "KM Survival Estimates",
  "operations": [
    { "id": "O010", "name": "Surv. prob. at Month 6",  "resultPattern": "X.XXX" },
    { "id": "O011", "name": "Surv. prob. at Month 12", "resultPattern": "X.XXX" },
    "... one Operation per timepoint ..."
  ] }
```

**Step 3: your turn.** Generate the KM ARD from `adtte`, ITT population. Two reminders
from Section 4.6: `AVAL` is the time variable, and `event = "1-CNSR"` converts the ADaM
censoring flag into an event indicator. (This exercise is skipped silently if `{cardx}` is
not installed.)

```{r ch4-ex4, exercise=TRUE, exercise.lines=20}
if (!requireNamespace("cardx", quietly = TRUE)) {
  cat("Install cardx to run this exercise: install.packages('cardx')\n")
} else {
  # Pick the first available PARAMCD as the endpoint
  param <- adtte$PARAMCD[1]

  adtte_endpoint <- adtte %>%
    filter(PARAMCD == ___ & ITTFL == "Y")     # which PARAMCD?

  ard_km <- cardx::ard_survival_survfit(
    data      = adtte_endpoint,
    by        = TRT01A,
    variables = c(time = "AVAL", event = "1-CNSR"),
    times     = c(6, 12, 18, 24)
  )

  cat("KM ARD rows:", nrow(ard_km), "\n")
  head(ard_km, 6)
}
```

```{r ch4-ex4-hint-1}
# Use the variable `param` that you just defined,
# i.e. filter(PARAMCD == param & ITTFL == "Y")
# In a real analysis you would name the endpoint explicitly, e.g. "OS".
```

```{r ch4-ex4-solution}
if (!requireNamespace("cardx", quietly = TRUE)) {
  cat("Install cardx to run this exercise.\n")
} else {
  param <- adtte$PARAMCD[1]

  adtte_endpoint <- adtte %>%
    filter(PARAMCD == param & ITTFL == "Y")

  ard_km <- cardx::ard_survival_survfit(
    data      = adtte_endpoint,
    by        = TRT01A,
    variables = c(time = "AVAL", event = "1-CNSR"),
    times     = c(6, 12, 18, 24)
  )
  cat("KM ARD rows:", nrow(ard_km), "\n")
  head(ard_km, 6)
}
# Each row is one OperationResult: a KM estimate at one timepoint
# for one treatment arm.
```

---

## Exercise 4.5: Map Flat ARD Columns to the Formal Model {data-progressive=FALSE}

This is the conceptual capstone of Chapter 4. Build a small lookup mapping each flat-ARD
column to its formal-ARS-model path, exactly the table from Section 4.5, reconstructed
from memory. This mapping is the core knowledge `arsbridge` needs. (The scaffold uses
`tribble()`, the row-wise table builder you met in Chapter 2.)

```{r ch4-ex5, exercise=TRUE, exercise.lines=18}
mappings <- tribble(
  ~flat_ard_column,   ~ars_model_path,
  "group1",           ___,
  "group1_level",     ___,
  "variable",         ___,
  "stat_name",        ___,
  "stat",             ___
)

mappings
```

```{r ch4-ex5-hint-1}
# Choose from these formal-model paths:
#   "GroupingFactor.groupingVariable"
#   "Group.name (or ResultGroup.groupValue)"
#   "Analysis.variable"
#   "Operation.name"
#   "OperationResult.rawValue"
```

```{r ch4-ex5-solution}
mappings <- tribble(
  ~flat_ard_column,   ~ars_model_path,
  "group1",           "GroupingFactor.groupingVariable",
  "group1_level",     "Group.name (or ResultGroup.groupValue)",
  "variable",         "Analysis.variable",
  "stat_name",        "Operation.name",
  "stat",             "OperationResult.rawValue"
)
mappings
# Note what's MISSING from the flat ARD:
#   analysisSetId, dataSubsetId, methodId, documentRefs.
# Those live only in the ARM-TS. That's the traceability gap.
```

---

## Chapter 4 Knowledge Check

Review map for the questions below: where the ARD lives and OperationResult anatomy
(Sections 4.1 and 4.2), string storage of values (4.8), the flat-ARD column mapping (4.5),
the traceability gap (4.5 and 4.7), and the `{cards}` / `{cardx}` division of labor (4.6).

```{r ch4-quiz, echo=FALSE}
quiz(
  caption = "Chapter 4: Check Your Understanding",

  question(
    "Where, in the formal ARS model, does the ARD physically live?",
    answer("As a separate top-level field of ReportingEvent"),
    answer("Inside Analysis.results as a list of OperationResult objects", correct = TRUE),
    answer("In a dedicated ARD file referenced by Output.fileSpecifications"),
    answer("Inside AnalysisMethod.codeTemplate"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Each OperationResult represents…",
    answer(
      "one statistical operation's result for one specific combination of groups",
      correct = TRUE
    ),
    answer("the complete output for one display table"),
    answer("all results for a given AnalysisSet"),
    answer("a single subject's value on the analysis variable"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Why are both rawValue and formattedValue stored as STRINGS in the ARS model?",
    answer(
      "To preserve the exact display representation AND retain full underlying precision; both matter for downstream validation",
      correct = TRUE
    ),
    answer("Because JSON cannot encode numeric values"),
    answer("To force every consumer to re-parse the values"),
    answer("Because R only supports character vectors"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which ARS-model field does the flat ARD column `stat` correspond to?",
    answer("Operation.name"),
    answer("OperationResult.rawValue", correct = TRUE),
    answer("Analysis.variable"),
    answer("Group.name"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "What is the 'traceability gap' that arsbridge is meant to close?",
    answer(
      "The flat ARD alone does NOT carry analysisSetId, dataSubsetId, methodId, or documentRefs; that context lives only in the ARM-TS",
      correct = TRUE
    ),
    answer("ADaM datasets cannot be linked to SDTM"),
    answer("R cannot read SAS XPT files"),
    answer("Define-XML does not describe ADaM"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which R package generates the flat ARD data frame?",
    answer("{gtsummary}"),
    answer("{cards}", correct = TRUE),
    answer("{datasetjson}"),
    answer("{admiral}"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which package extends {cards} with KM estimates, regression, and inferential statistics?",
    answer("{cardx}", correct = TRUE),
    answer("{rtables}"),
    answer("{tfrmt}"),
    answer("{admiral}"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "An OperationResult has rawValue = '84.44444' and formattedValue = '84.4'. What can you infer?",
    answer(
      "The operation's resultPattern rounds to one decimal place (e.g., 'XX.X')",
      correct = TRUE
    ),
    answer("The analysis produced two different statistics"),
    answer("The result failed validation"),
    answer("Both values are wrong; they must be identical"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

---

## Chapter 4 Summary {.congratulations}

You now understand:

- The **ARD is the result layer**: `Analysis.results` holds a list of `OperationResult`
  objects
- Each `OperationResult` stores **one statistic for one group combination**, with both
  `rawValue` (full precision) and `formattedValue` (display string) as strings
- The **flat ARD format** (used by `{cards}`) is a tabular extraction of this data: each
  row is one result with grouping/variable/stat columns, filterable like any data frame
- The flat ARD alone is **not sufficient for traceability**: `analysisSetId`,
  `dataSubsetId`, `methodId`, and `documentRefs` live only in the ARM-TS
- Generating ARDs in R uses **`{cards}`** for descriptive statistics and **`{cardx}`** for
  KM estimates, t-tests, and regression
- You have generated **continuous**, **dichotomous**, and **Kaplan-Meier** ARDs from real
  pharmaverse data

**Next:** Chapter 5, The Traceability Chain: re-derive a stored ARD result from ARM-TS
plus ADaM, and find a deliberate traceability break
