---
title: "CDISC Analysis Results Standard (ARS): Interactive Training"
subtitle: "Chapter 7: Package Design Roadmap for arsbridge"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Chapter 7 lays out the v0.1 design for {arsbridge}: five functions
  (parse_ars, ars_to_shell, shell_to_ars, validate_ard, ars_summary), the
  internal WhereClause helpers, the build sequence, package infrastructure,
  and the GitHub / LinkedIn launch strategy. You will prototype the core
  helpers and sketch the parser by hand.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(dplyr)
library(jsonlite)
library(glue)

tutorial_options(exercise.timelimit = 60)
gradethis_setup()

data("adsl",  package = "pharmaverseadam")
data("adae",  package = "pharmaverseadam")

adsl_safe <- adsl %>% filter(SAFFL == "Y")
re_json   <- jsonlite::read_json("data/reporting_event.json")
```

---

## Chapter 7: Package Design Roadmap for `arsbridge`

Six chapters of concepts converge here. Every exercise in this chapter prototypes a real
piece of the package, using skills you already have: WhereClause reading (Chapter 2),
object navigation and ID lookups (Chapter 3), the flat-ARD mapping (Chapter 4), and the
validation checks (Chapter 5).

### 7.1 Package Concept

`arsbridge` is an R package with a single clear purpose:

> **Bridge the annotated-shell world and the ARS world, bidirectionally, and validate
> traceability between ARM-TS and ARD.**

It serves two audiences:

1. **Practitioners with annotated shells** who want to adopt ARS without starting from
   scratch
2. **ARS adopters** who want a human-readable view of their ARM-TS files and a validator
   for their pipeline

---

### 7.2 The Five v0.1 Functions

| Function | Direction | Input | Output |
|----------|-----------|-------|--------|
| `parse_ars()` | (foundation) | ARM-TS JSON file path | R list (ARS object tree) |
| `ars_to_shell()` | ARM-TS → Shell | ARS R list + output ID | Excel annotated shell |
| `shell_to_ars()` | Shell → ARM-TS | Excel annotated shell | ARM-TS JSON file |
| `validate_ard()` | Validate | ARS R list + ARD data frame | Validation report |
| `ars_summary()` | Explore | ARS R list | Console summary |

This is a complete, coherent v0.1. Each function is **independently useful**. Together
they cover the full bridge concept.

---

### 7.3 `parse_ars()`: The Foundation

**Purpose.** Read an ARM-TS JSON file; return a structured R list with elements indexed
by object ID for fast lookup.

**Signature**

```r
parse_ars(
  file,            # Path to ARM-TS JSON
  validate = TRUE  # Validate against ARS schema
)
```

**Returns** an S3 object of class `"ars_reporting_event"` with named-list slots:

- `$id`, `$name`, `$version`
- `$reference_documents`, indexed by ID
- `$analysis_sets`, `$data_subsets`, `$groupings`, `$methods`, `$analyses`, `$outputs`,
  each indexed by ID

**Why ID-indexed?** Because everything in ARM-TS is referenced by ID
(`analysisSetId = "AS002"`, `methodId = "M001"`, and so on). ID-indexed lists make
`re$analysis_sets[["AS002"]]` a constant-time lookup instead of a linear scan. You built
exactly this lookup by hand in Chapter 3 (Exercise 3.6, with `setNames`); `parse_ars()`
just applies it to every collection.

---

### 7.4 `ars_to_shell()`: Render ARM-TS as Annotated Excel

**Purpose.** Generate a human-readable annotated shell in Excel from an ARM-TS object,
for a specified output or for all outputs.

**Signature**

```r
ars_to_shell(
  re,                 # ARS reporting event (from parse_ars())
  output_id = NULL,   # NULL -> generate shells for all outputs
  file      = "annotated_shell.xlsx",
  style     = c("standard", "cdisc_pilot")
)
```

**Output Excel columns**, one worksheet per output:

| Column | ARM-TS source |
|--------|---------------|
| Table Number / Title | `Output.id` / `OutputDisplay.displayTitle` |
| Dataset / Variable | `Analysis.dataset` / `Analysis.variable` |
| Population (name + filter text) | `AnalysisSet.name` + generated from `condition` |
| Data Filter | Generated from `DataSubset.condition` |
| Grouping 1 / 2 | `GroupingFactor.name` (one or two) |
| Method / Operations / Format | `AnalysisMethod.name`, `[Operation.name]`, `resultPattern` |
| SAP Reference | Generated from `DocumentReference + PageRef` |
| Program | `programmingCode.documentRef` |
| Reason / Purpose | `Analysis.reason` / `Analysis.purpose` |

The **core internal helper** turning WhereClause objects into filter text:

```r
where_clause_to_text <- function(wc) {
  if (!is.null(wc$condition)) {
    cond     <- wc$condition
    val_str  <- paste0("'", unlist(cond$value), "'", collapse = ", ")
    glue::glue("{cond$dataset}.{cond$variable} {cond$comparator} {val_str}")
  } else if (!is.null(wc$compoundExpression)) {
    op      <- wc$compoundExpression$logicalOperator
    clauses <- sapply(wc$compoundExpression$whereClauses, where_clause_to_text)
    if (op == "NOT") glue::glue("NOT ({clauses[1]})")
    else paste(clauses, collapse = glue::glue(" {op} "))
  }
}
```

You will build the simple-condition core of this function yourself in Exercise 7.1.

---

### 7.5 `shell_to_ars()`: Parse Annotated Excel into ARM-TS

**Purpose.** Read an annotated Excel shell and emit an ARM-TS JSON file.

**Signature**

```r
shell_to_ars(
  file,                              # Excel shell
  study_id,                          # ReportingEvent ID
  study_name,                        # ReportingEvent name
  sap_file    = NULL,                # SAP PDF path (for ReferenceDocument)
  output_file = "reporting_event.json",
  template    = c("arsbridge_standard", "custom"),
  col_mapping = NULL                 # For custom shell formats
)
```

**The hardest internal function**, `text_to_where_clause()`, is the inverse of
`where_clause_to_text()`. It parses text like `"ADSL.SAFFL EQ 'Y'"` into a
`WhereClauseCondition`:

```r
text_to_where_clause <- function(text) {
  pattern <- "^(\\w+)\\.(\\w+)\\s+(EQ|NE|GT|GE|LT|LE|IN|NOTIN)\\s+'(.+)'$"
  m <- regmatches(text, regexec(pattern, text, perl = TRUE))[[1]]
  list(
    dataset    = m[2],
    variable   = m[3],
    comparator = m[4],
    value      = list(m[5])
  )
}
```

For **v0.1**, support simple single-condition WHERE clauses. Compound expressions
(AND/OR/NOT) go to **v0.2**. Scope discipline matters more than completeness in a first
release.

---

### 7.6 `validate_ard()`: The Traceability Validator

**Purpose.** Verify an ARD is consistent with the ARM-TS specification.

**Signature**

```r
validate_ard(
  re,              # ARS reporting event
  ard,             # ARD data frame (from {cards})
  analysis_id,     # NULL -> validate all
  adam_data = NULL # Optional: ADaM for re-computation check
)
```

**Six checks** (you prototyped #2 and #5 in Chapter 5):

1. **Coverage**: every Analysis in ARM-TS has rows in the ARD
2. **Re-derivation**: applying the ARM-TS spec to ADaM reproduces stored `rawValue`
3. **Population consistency**: `AnalysisSet` applied to ADaM matches the ARD denominator
4. **Group completeness**: every expected `Group` combination appears in the ARD
5. **Format consistency**: `formattedValue` matches `rawValue + resultPattern`
6. **SAP page exists**: `documentRef.pageRefs` resolve to real pages (optional PDF check)

**Returns** a two-part report: a `$summary` list of counts, and a `$details` data frame
of per-check rows with `status` in `{PASS, FAIL, WARN}` and a `message`.

---

### 7.7 `ars_summary()`: A Human-Readable Print Method

A `print()` for `ars_reporting_event` objects. Useful for exploring an unfamiliar ARS
file.

```
── ARS Reporting Event ────────────────────────────────────
ID:      RE001
Name:    Study XYZ-001 Final CSR
Version: 1

── Populations (AnalysisSets) ────────────────────────────
  AS001  Intent-to-Treat Population    ADSL.ITTFL EQ 'Y'
  AS002  Safety Population             ADSL.SAFFL EQ 'Y'
  AS003  Per-Protocol Population       ADSL.PPSFL EQ 'Y' AND ADSL.COMPLFL EQ 'Y'

── Methods ──────────────────────────────────────────────
  M001   Summary Statistics - Continuous   6 operations
  M002   Count and Percentage              3 operations

── Analyses ─────────────────────────────────────────────
  45 analyses defined
  38 analyses have stored results (ARD)
   7 analyses have no stored results
```

A small but high-impact function. It is how new users will **explore** an unfamiliar
ARM-TS file before doing anything else, and it is the first thing anyone sees in a demo.

---

### 7.8 Package Infrastructure

**Dependencies**

```
Imports:
  jsonlite      # JSON read/write
  openxlsx2     # Excel read/write (modern replacement for openxlsx)
  dplyr         # Data manipulation
  glue          # String interpolation
  cli           # Console output formatting
  rlang         # Error handling

Suggests:
  cards         # ARD generation examples
  cardx         # Extended ARD
  haven         # SAS dataset reading in vignettes
  testthat      # Testing
  knitr         # Vignette building
```

**File structure**

```
arsbridge/
├── R/
│   ├── parse_ars.R
│   ├── ars_to_shell.R
│   ├── shell_to_ars.R
│   ├── validate_ard.R
│   ├── ars_summary.R
│   ├── where_clause.R     # Internal helpers
│   ├── schema_objects.R   # Internal R representations of ARS classes
│   └── utils.R
├── inst/
│   ├── extdata/
│   │   ├── ars_schema/    # ARS v1.0 JSON Schema
│   │   └── templates/     # Shell Excel templates
│   └── examples/
│       └── cdisc_pilot/
├── vignettes/
│   ├── getting_started.Rmd
│   ├── shell_to_ars_workflow.Rmd
│   └── validating_ard.Rmd
├── tests/testthat/
└── data/                   # Small example ARS/ARD objects
```

---

### 7.9 Build Sequence (v0.1)

Recommended order, designed to **minimize rework**:

1. **`schema_objects.R`**: R structures mirroring ARS classes
2. **`parse_ars()`**: once you can parse, everything builds on it
3. **`ars_summary()`**: quick win to verify parsing; useful for debugging
4. **`where_clause.R` helpers**: needed by both shell functions and `validate_ard()`
5. **`ars_to_shell()`**: tests your understanding of object navigation
6. **`validate_ard()`**: Checks 1 to 4 first; add re-computation last
7. **`shell_to_ars()`**: the hardest function; build after the shell template is
   established
8. **Vignette: CDISC pilot**: walk through `parse_ars`, `ars_summary`, `ars_to_shell`,
   `validate_ard` on public CDISC pilot data

---

### 7.10 GitHub and LinkedIn Strategy

**Repo:** `arsbridge`
**Tagline:** *"R package for bidirectional translation between annotated TLF shells and
CDISC ARS ARM-TS, with ARD traceability validation."*

**Three-post LinkedIn launch series**

1. **(at repo creation)** *"The clinical programming community spent 20 years annotating
   TLF shells in Word files. CDISC ARS is the machine-readable version of exactly that.
   I'm building the R package that bridges them."*
2. **(when `parse_ars` + `ars_to_shell` work)** *"You can now take any ARS ARM-TS JSON
   file and render it as a human-readable annotated Excel shell in 3 lines of R. Here's a
   demo using the CDISC pilot study data."*
3. **(when `validate_ard` works)** *"arsbridge v0.1 is on GitHub: parse ARM-TS, generate
   annotated shells, validate ARD traceability. First open-source R tool that closes the
   loop between analysis specification and results. Feedback welcome."*

---

## Hands-On: Prototype the Core Helpers

Each exercise below is a real piece of `arsbridge` v0.1, in build-sequence order.

---

## Exercise 7.1: Prototype `where_clause_to_text()` {data-progressive=FALSE}

**Step 1: the annotated shell.** The text a human wants to read — the subtitle line on every safety table:

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:8px 14px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; text-align:center;">Table 14.3.1: Adverse Event Summary</div>
<div style="text-align:center; font-family:sans-serif; margin-bottom:4px;">Safety Population &nbsp;(<span style="color:#b3400e; font-weight:bold;">ADSL.SAFFL='Y'</span>)</div>
<div style="color:#777; font-family:sans-serif; font-size:0.85em; text-align:center;">↑ <code>where_clause_to_text(condition)</code> must reconstruct this red annotation string</div>
</div>

**Step 2: the ARM-TS encoding.** The JSON a machine stores:

```json
{ "condition": { "dataset": "ADSL", "variable": "SAFFL",
                 "comparator": "EQ", "value": ["Y"] } }
```

**Step 3: your turn.** Build the function that converts the second into the first, for
any condition. This is the core internal helper powering `ars_to_shell()` and
`ars_summary()`. You did the one-off version in Chapter 3 (Exercise 3.1); now wrap it as
a reusable function.

```{r ch7-ex1, exercise=TRUE, exercise.lines=18}
where_clause_to_text <- function(condition) {
  # Wrap each value in single quotes and comma-separate if more than one
  val_str <- paste0("'", unlist(condition$value), "'", collapse = ", ")

  # Glue together: DATASET.VARIABLE COMPARATOR 'value'
  glue::glue(
    "{condition$___}.{condition$___} ",   # dataset.variable
    "{condition$___} ",                   # comparator
    "{val_str}"
  )
}

# Test on the Safety Population AnalysisSet
safe_cond <- re_json$analysisSets[[2]]$condition
as.character(where_clause_to_text(safe_cond))

# Expected: "ADSL.SAFFL EQ 'Y'"
```

```{r ch7-ex1-hint-1}
# The condition has four fields:
#   dataset, variable, comparator, value
# You want a string of the form: "ADSL.SAFFL EQ 'Y'"
```

```{r ch7-ex1-solution}
where_clause_to_text <- function(condition) {
  val_str <- paste0("'", unlist(condition$value), "'", collapse = ", ")
  glue::glue(
    "{condition$dataset}.{condition$variable} ",
    "{condition$comparator} ",
    "{val_str}"
  )
}
safe_cond <- re_json$analysisSets[[2]]$condition
as.character(where_clause_to_text(safe_cond))
# "ADSL.SAFFL EQ 'Y'". This single function powers both ars_to_shell()
# AND the population display line in ars_summary().
```

```{r ch7-ex1-check}
grade_result(
  pass_if(~ as.character(.result) == "ADSL.SAFFL EQ 'Y'",
          "Correct. This is the core of ars_to_shell(). It composes the population/data-filter text every time arsbridge needs to render a WhereClause."),
  fail_if(~ TRUE,
          "Use condition$dataset, condition$variable, and condition$comparator. The value list is wrapped with single quotes by paste0('\\''...).")
)
```

---

## Exercise 7.2: Prototype `parse_ars()`, the ID-Indexed Lookup {data-progressive=FALSE}

**Step 1: the annotated shell.** In a shell package, you find a table by its number
("Table 14.1.1"). In ARS, you find any object by its ID.

**Step 2: the ARM-TS encoding.** IDs are defined once and referenced everywhere, so the
parser's job is to make every lookup instant:

```json
"analyses": [
  { "id": "AN001", "analysisSetId": "AS002", "methodId": "M001", "...": "..." }
]
```

**Step 3: your turn.** The defining feature of `parse_ars()` is that every collection is
**indexed by object ID**, so that lookups like `re$analyses[["AN001"]]` are O(1). Build
that indexing pattern. It is the same `setNames()` + `sapply()` combination you used to
resolve references in Chapter 3 (Exercise 3.6), promoted to a reusable helper.

```{r ch7-ex2, exercise=TRUE, exercise.lines=20}
# A small helper: take a list of ARS objects and return a named list keyed by $id
index_by_id <- function(object_list) {
  ids <- sapply(object_list, function(x) x$___)   # which field carries the ID?
  setNames(object_list, ids)
}

# Apply it to the Analyses in re_json
analyses_by_id <- index_by_id(re_json$___)         # which top-level field?

# Now access AN001 directly by ID
an001 <- analyses_by_id[["AN001"]]
cat("Analysis name:    ", an001$name,          "\n")
cat("Dataset.variable: ", an001$dataset, ".", an001$variable, "\n", sep = "")
cat("Population ID:    ", an001$analysisSetId, "\n")
```

```{r ch7-ex2-hint-1}
# Each ARS object has an $id field.
# The top-level field on re_json holding the list of analyses is $analyses.
```

```{r ch7-ex2-solution}
index_by_id <- function(object_list) {
  ids <- sapply(object_list, function(x) x$id)
  setNames(object_list, ids)
}

analyses_by_id <- index_by_id(re_json$analyses)
an001 <- analyses_by_id[["AN001"]]
cat("Analysis name:    ", an001$name,          "\n")
cat("Dataset.variable: ", an001$dataset, ".", an001$variable, "\n", sep = "")
cat("Population ID:    ", an001$analysisSetId, "\n")
# parse_ars() runs index_by_id() over EVERY object collection:
# analysisSets, dataSubsets, groupings, methods, analyses, outputs.
# That single transformation is most of what parse_ars() does.
```

---

## Exercise 7.3: Prototype `ars_summary()`, the Populations Section {data-progressive=FALSE}

**Step 1: the annotated shell.** The human output `ars_summary()` must produce — the populations section of a shell cover page:

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<div style="font-weight:bold; margin-bottom:4px; font-family:sans-serif;">── Populations (AnalysisSets) ───────────────────────────────</div>
<div>&nbsp;&nbsp;AS001 &nbsp; ITT Population &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; <span style="color:#b3400e; font-weight:bold;">ADSL.ITTFL EQ 'Y'</span></div>
<div>&nbsp;&nbsp;AS002 &nbsp; Safety Population &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; <span style="color:#b3400e; font-weight:bold;">ADSL.SAFFL EQ 'Y'</span></div>
<div style="color:#777; font-family:sans-serif; font-size:0.85em; margin-top:6px;">↑ each red annotation is what <code>where_clause_to_text(as$condition)</code> returns</div>
</div>

**Step 2: the ARM-TS encoding.** Two AnalysisSets, each with a `condition` field the function must render:

```json
"analysisSets": [
  { "id": "AS001", "name": "ITT Population",
    "condition": { "dataset": "ADSL", "variable": "ITTFL", "comparator": "EQ", "value": ["Y"] } },
  { "id": "AS002", "name": "Safety Population",
    "condition": { "dataset": "ADSL", "variable": "SAFFL", "comparator": "EQ", "value": ["Y"] } }
]
```

**Step 3: your turn.** Build a small function that prints the **Populations** section of `ars_summary()`: one
line per AnalysisSet showing `id`, `name`, and the filter text. Reuse
`where_clause_to_text()` from Exercise 7.1.

Formatting note: `sprintf("%-30s", x)` left-pads the string `x` to 30 characters, which is
what keeps the columns aligned in the console output.

```{r ch7-ex3, exercise=TRUE, exercise.lines=22}
where_clause_to_text <- function(condition) {
  val_str <- paste0("'", unlist(condition$value), "'", collapse = ", ")
  glue::glue(
    "{condition$dataset}.{condition$variable} ",
    "{condition$comparator} ",
    "{val_str}"
  )
}

print_populations <- function(re) {
  cat("── Populations (AnalysisSets) ───────────────────────────\n")
  for (as in re$___) {                       # which top-level field?
    filter_text <- where_clause_to_text(as$condition)
    cat(sprintf("  %s  %-30s  %s\n", as$___, as$___, filter_text))
  }
}

print_populations(re_json)
```

```{r ch7-ex3-hint-1}
# Iterate over re$analysisSets
# Each AnalysisSet has $id, $name, and $condition
```

```{r ch7-ex3-solution}
where_clause_to_text <- function(condition) {
  val_str <- paste0("'", unlist(condition$value), "'", collapse = ", ")
  glue::glue(
    "{condition$dataset}.{condition$variable} ",
    "{condition$comparator} ",
    "{val_str}"
  )
}

print_populations <- function(re) {
  cat("── Populations (AnalysisSets) ───────────────────────────\n")
  for (as in re$analysisSets) {
    filter_text <- where_clause_to_text(as$condition)
    cat(sprintf("  %s  %-30s  %s\n", as$id, as$name, filter_text))
  }
}
print_populations(re_json)
# Two lines for AS001 (ITT) and AS002 (Safety), populated entirely from
# the ARM-TS, with no hard-coded knowledge of CDISCPILOT01.
# ars_summary() composes 5 print_* functions like this one: populations,
# data subsets, groupings, methods, analyses.
```

---

## Exercise 7.4: Sketch `validate_ard()` Check #1, Coverage {data-progressive=FALSE}

**Step 1: the annotated shell.** The validation report a QC programmer reads — which ARM-TS analyses are covered by the ARD and which are missing:

<div style="font-family:'Courier New',monospace; border:1px solid #aaa; background:#fff; padding:10px 16px; font-size:0.86em; line-height:1.5; margin-bottom:6px;">
<table style="border-collapse:collapse; width:100%;">
<tr style="background:#f0f0f0;">
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left;">Analysis ID</th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left;">ARM-TS name &nbsp;<span style="color:#b3400e; font-weight:normal; font-size:0.9em;">DATASET.VARIABLE</span></th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:center;">Status</th>
  <th style="border:1px solid #bbb; padding:3px 10px; text-align:left;">Message</th>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">AN001</td>
  <td style="border:1px solid #bbb; padding:3px 10px;">Age Summary &nbsp;<span style="color:#b3400e;">ADSL.AGE</span></td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center; color:green; font-weight:bold;">PASS</td>
  <td style="border:1px solid #bbb; padding:3px 10px;">ARD rows present</td>
</tr>
<tr>
  <td style="border:1px solid #bbb; padding:3px 10px;">AN042</td>
  <td style="border:1px solid #bbb; padding:3px 10px;">KM PFS Analysis &nbsp;<span style="color:#b3400e;">ADTTE.AVAL</span></td>
  <td style="border:1px solid #bbb; padding:3px 10px; text-align:center; color:#b3400e; font-weight:bold;">FAIL</td>
  <td style="border:1px solid #bbb; padding:3px 10px;">No ARD rows found</td>
</tr>
</table>
<div style="color:#777; font-family:sans-serif; font-size:0.82em; margin-top:4px;">Check #1 of <code>validate_ard()</code>: every ARM-TS analysis must have results in the ARD</div>
</div>

**Step 2: the ARM-TS source.** IDs from the ARM-TS drive the left column; the ARD's coverage drives the status:

```json
"analyses": [ { "id": "AN001", "dataset": "ADSL", "variable": "AGE", "...": "..." } ]
```

**Step 3: your turn.** Write the coverage check.

Check #1 asks: **does every Analysis in the ARM-TS have at least one row in the ARD?**
Write a small coverage check that returns a data frame of `analysis_id` + `status` +
`message`.

The membership test is R's `%in%` operator: `x %in% y` returns `TRUE` for each element of
`x` found anywhere in `y`. Vectorized, so one call checks every analysis at once.

```{r ch7-ex4, exercise=TRUE, exercise.lines=22}
# Simulate an ARD subset: in this sample re_json there's only AN001, with results.
# Pretend the ARD covers only the analyses listed below.
ard_analysis_ids <- c("AN001")             # what's actually in the ARD
ars_analysis_ids <- sapply(re_json$analyses, function(a) a$id)

coverage_check <- function(ars_ids, ard_ids) {
  tibble(
    analysis_id = ars_ids,
    status      = ifelse(ars_ids %in% ___, "PASS", "FAIL"),   # what to compare against?
    message     = ifelse(ars_ids %in% ard_ids,
                         "ARD rows present",
                         "No ARD rows found")
  )
}

coverage_check(ars_analysis_ids, ard_analysis_ids)
```

```{r ch7-ex4-hint-1}
# You're testing whether each ARM-TS analysis_id is found in the set of
# analysis IDs that the ARD covers.
```

```{r ch7-ex4-solution}
ard_analysis_ids <- c("AN001")
ars_analysis_ids <- sapply(re_json$analyses, function(a) a$id)

coverage_check <- function(ars_ids, ard_ids) {
  tibble(
    analysis_id = ars_ids,
    status      = ifelse(ars_ids %in% ard_ids, "PASS", "FAIL"),
    message     = ifelse(ars_ids %in% ard_ids,
                         "ARD rows present",
                         "No ARD rows found")
  )
}
coverage_check(ars_analysis_ids, ard_analysis_ids)
# In a real study with 45 ARM-TS analyses and 38 ARD-covered, you'd see
# 7 FAIL rows, each one a specific analysis that was specified but never
# computed. That's exactly the kind of gap a validator catches.
```

---

## Exercise 7.5: Capstone, Map Each Function to the Pipeline Step {data-progressive=FALSE}

Build a small lookup that links each `arsbridge` function to the pipeline step it serves
and the pieces it depends on. This is the conceptual map you'll explain to anyone who
asks *"what does arsbridge do?"*

```{r ch7-ex5, exercise=TRUE, exercise.lines=15}
arsbridge_map <- tribble(
  ~function_name,    ~pipeline_step,                                    ~depends_on,
  "parse_ars",       "Read ARM-TS JSON into R",                         "(none)",
  "ars_summary",     "Human-readable exploration of an ARS object",     ___,         # depends on?
  "ars_to_shell",    "Render ARM-TS as annotated Excel shell",          ___,
  "shell_to_ars",    "Translate annotated Excel shell to ARM-TS JSON",  "where_clause helpers (text_to_where_clause)",
  "validate_ard",    "Cross-check ARD against ARM-TS specification",    ___
)

arsbridge_map
```

```{r ch7-ex5-hint-1}
# ars_summary, ars_to_shell, and validate_ard all consume an already-parsed
# ReportingEvent. What function produces that?
```

```{r ch7-ex5-solution}
arsbridge_map <- tribble(
  ~function_name,    ~pipeline_step,                                    ~depends_on,
  "parse_ars",       "Read ARM-TS JSON into R",                         "(none)",
  "ars_summary",     "Human-readable exploration of an ARS object",     "parse_ars",
  "ars_to_shell",    "Render ARM-TS as annotated Excel shell",          "parse_ars + where_clause_to_text",
  "shell_to_ars",    "Translate annotated Excel shell to ARM-TS JSON",  "where_clause helpers (text_to_where_clause)",
  "validate_ard",    "Cross-check ARD against ARM-TS specification",    "parse_ars + where_clause helpers"
)
arsbridge_map
# Note: 4 of the 5 functions depend on parse_ars().
# That's why parse_ars() is item #2 on the build sequence, right after
# defining the R schema objects.
```

---

## Chapter 7 Knowledge Check

Review map: function roster and ID indexing (7.2 and 7.3), the two shell functions and
their helpers (7.4 and 7.5), the six checks (7.6), infrastructure (7.8), build order and
scope decisions (7.9).

```{r ch7-quiz, echo=FALSE}
quiz(
  caption = "Chapter 7: Check Your Understanding",

  question(
    "Which arsbridge function is the foundation that almost every other function depends on?",
    answer("ars_to_shell()"),
    answer("parse_ars()", correct = TRUE),
    answer("validate_ard()"),
    answer("shell_to_ars()"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Why does parse_ars() return ID-indexed lists rather than ordered arrays?",
    answer(
      "Because every ARS object is referenced by ID elsewhere; ID-indexed lists give O(1) lookups for analysisSetId, methodId, etc.",
      correct = TRUE
    ),
    answer("Because the ARS schema requires it"),
    answer("Because R cannot store ordered lists"),
    answer("Because JSON is intrinsically unordered"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which function is the INVERSE of ars_to_shell()?",
    answer("validate_ard()"),
    answer("shell_to_ars()", correct = TRUE),
    answer("parse_ars()"),
    answer("ars_summary()"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which internal helper is reused by both ars_to_shell() and ars_summary()?",
    answer("text_to_where_clause()"),
    answer("where_clause_to_text()", correct = TRUE),
    answer("index_by_id()"),
    answer("coverage_check()"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which validate_ard() check requires ADaM data as an input?",
    answer("Check #1 (Coverage)"),
    answer("Check #2 (Re-derivation from ADaM)", correct = TRUE),
    answer("Check #4 (Group completeness)"),
    answer("Check #5 (Format consistency)"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Why is shell_to_ars() built LAST in the recommended build sequence?",
    answer(
      "Because parsing free-form annotation text into structured WhereClauses is the hardest part, and the Excel shell template must be stable first",
      correct = TRUE
    ),
    answer("Because it depends on validate_ard()"),
    answer("Because it requires a paid CRAN account"),
    answer("Because it has no test coverage"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which two packages handle Excel and JSON read/write in arsbridge?",
    answer("readxl + jsonify"),
    answer("openxlsx2 + jsonlite", correct = TRUE),
    answer("xlsx + RJSONIO"),
    answer("haven + arrow"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "What is the v0.1 scope decision for compound WhereClause expressions (AND/OR/NOT) in shell_to_ars()?",
    answer("Fully supported in v0.1"),
    answer("Deferred to v0.2; v0.1 supports simple single-condition WHERE only", correct = TRUE),
    answer("Not planned for any version"),
    answer("Implemented only for AND, never OR or NOT"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

---

## Tutorial Completion {.congratulations}

You've completed the **ARS Interactive Training**.

### What you now know

- **Why ARS exists** and the five chronic problems it addresses (Chapter 1)
- The complete **ARS object hierarchy**, ReportingEvent down to OperationResult
  (Chapter 2)
- **Every major ARM-TS object** with field-by-field annotated-shell mappings (Chapter 3)
- How to **generate ARDs** with `{cards}` and `{cardx}` on real pharmaverse data, and the
  mapping between the flat ARD and the formal model (Chapter 4)
- The **end-to-end traceability chain** from SAP through ARM-TS, ADaM, R, and ARD to the
  display, plus how to detect a population traceability break and a format-consistency
  break (Chapter 5)
- Exactly **where the open-source ecosystem stops**: `{cards}`, `{cardx}`, `{gtsummary}`,
  and `{datasetjson}` cover their layers, but the ARM-TS layer is uncovered (Chapter 6)
- A complete **v0.1 design for `arsbridge`**: five functions, internal helpers, build
  sequence, package infrastructure, and launch plan (Chapter 7)

### Next steps

1. **Build `arsbridge`.** Start with `parse_ars()`; everything else builds on it. Follow
   the build sequence in Section 7.9.
2. **Validate against the CDISC Pilot study.** The CDISC sdtm-adam-pilot-project on
   GitHub has public ADaM data. Build the vignette `getting_started.Rmd` against it.
3. **Publish.** Use the three-post LinkedIn series in Section 7.10 to tie each milestone
   to a public announcement.

### Key resources

- [ARS v1.0 Specification](https://cdisc-org.github.io/analysis-results-standard/)
- [ARS GitHub (model YAML, examples)](https://github.com/cdisc-org/analysis-results-standard)
- [CDISC Pilot Study data](https://github.com/cdisc-org/sdtm-adam-pilot-project)
- [cards](https://insightsengineering.github.io/cards/) ·
  [cardx](https://insightsengineering.github.io/cardx/) ·
  [gtsummary](https://www.danieldsjoberg.com/gtsummary/) ·
  [datasetjson](https://atorus-research.github.io/datasetjson/)
- [pharmaverse](https://pharmaverse.org)

You have everything you need. Go build it.
