---
title: "CDISC Analysis Results Standard (ARS) — Interactive Training"
subtitle: "Chapter 3: ARM-TS Deep Dive"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Chapter 3 walks through every major ARM-TS object with full JSON examples and
  the equivalent annotated-shell annotation. You will dissect the running case
  study — a safety AE summary by treatment arm — and build ARM-TS objects (an
  AnalysisMethod, an Analysis) by hand from a shell annotation.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(jsonlite)
library(glue)

tutorial_options(exercise.timelimit = 60)
gradethis_setup()

# Load pharmaverse ADaM datasets
data("adsl",  package = "pharmaverseadam")
data("adae",  package = "pharmaverseadam")
data("adtte", package = "pharmaverseadam")
data("adlb",  package = "pharmaverseadam")

adsl_safe <- adsl %>% filter(SAFFL == "Y")
adsl_itt  <- adsl %>% filter(!is.na(ARM) & ARM != "")
adae_teae <- adae %>% filter(TRTEMFL == "Y")

re_json <- jsonlite::read_json("data/reporting_event.json")
```

---

## Chapter 3: ARM-TS Deep Dive

This chapter walks through every major ARM-TS object with full JSON examples. Each section
includes the equivalent annotated-shell annotation so you can see exactly how your existing
knowledge maps to the formal model.

The examples use a running case study: **a safety table showing adverse event summary
statistics by treatment arm in the Safety Population** — a table type you have programmed many
times.

---

### 3.1 ReportingEvent

The **root container**. Represents one reporting deliverable — typically one CSR or one
interim analysis.

**Key fields**

| Field | Required | Description |
|-------|----------|-------------|
| `id`, `name` | Yes | Unique identifier and display name |
| `version` | No | Version number (increment on updates) |
| `mainListOfContents` | Yes | Structured table of contents |
| `referenceDocuments` | No | External documents (SAP, programs) |
| `analysisSets`, `dataSubsets`, `analysisGroupings`, `methods`, `analyses`, `outputs` | No | The object collections |

**Annotated-shell equivalent.** The annotated-shell header block — study number, protocol
title, SAP version, date. The `ReportingEvent` is the machine-readable version of that header.

---

### 3.2 ReferenceDocument

Links the reporting event to **external files**. This is how ARS creates a formal link to
the SAP.

```json
{
  "id": "SAP001",
  "name": "Statistical Analysis Plan v2.0",
  "location": "documents/XYZ001_SAP_v2.0.pdf",
  "description": "Final SAP approved 15-Jan-2025"
}
```

**Annotated-shell equivalent.** The *"per Section X.X of SAP"* text you write. In ARM-TS this
becomes a formal `DocumentReference` (with optional `pageRefs`) inside the `Analysis` object,
pointing to the `ReferenceDocument` you define here.

---

### 3.3 AnalysisSet

Defines a **subject population**. Your population flags in ADSL (`ITTFL`, `SAFFL`, `PPSFL`)
become `AnalysisSet` objects.

**Simple condition (single flag)**

```json
{
  "id": "AS002",
  "name": "Safety Population",
  "condition": {
    "dataset": "ADSL",
    "variable": "SAFFL",
    "comparator": "EQ",
    "value": ["Y"]
  }
}
```

**Compound expression (multiple flags)**

```json
{
  "id": "AS003",
  "name": "Per-Protocol Population",
  "compoundExpression": {
    "logicalOperator": "AND",
    "whereClauses": [
      { "condition": { "dataset": "ADSL", "variable": "PPSFL",   "comparator": "EQ", "value": ["Y"] } },
      { "condition": { "dataset": "ADSL", "variable": "COMPLFL", "comparator": "EQ", "value": ["Y"] } }
    ]
  }
}
```

Exactly one of `condition` or `compoundExpression` should be present.

**Annotated-shell equivalent.** *"SAFFL='Y' from ADSL"* or *"Safety Population"*. You define
it **once** as `AS002`, and every analysis referencing the Safety Population just sets
`"analysisSetId": "AS002"` — this is **Principle 1 (reuse)** in action.

---

### 3.4 DataSubset

A filter on **data records** (rows in an analysis dataset), not on subjects.

**Key distinction**

- `AnalysisSet` = *who* is in the analysis (subject-level, usually from ADSL)
- `DataSubset` = *which records* to use (row-level, on the analysis dataset itself)

```json
{
  "id": "DS001",
  "name": "Overall Survival Parameter",
  "condition": {
    "dataset": "ADTTE",
    "variable": "PARAMCD",
    "comparator": "EQ",
    "value": ["OS"]
  }
}
```

**Annotated-shell equivalent.** *"PARAMCD='OS'"* or *"AVISIT='Week 24'"*. The distinction
matters: your population flag selects **subjects**; your `PARAMCD` filter selects **records
within those subjects' data**.

---

### 3.5 GroupingFactor and Group

Defines how to subdivide the analysis. Treatment arm, sex, age group, visit — these are all
grouping factors.

**GroupingFactor — key fields**

| Field | Description |
|-------|-------------|
| `groupingDataset`, `groupingVariable` | Where the grouping variable lives |
| `dataDriven` | `true` = groups derived from data; `false` = groups pre-specified |
| `groups` | Pre-specified groups (when `dataDriven = false`) |

**Pre-specified (treatment arm)**

```json
{
  "id": "GF001",
  "name": "Treatment Arm",
  "groupingDataset": "ADSL",
  "groupingVariable": "TRT01A",
  "dataDriven": false,
  "groups": [
    { "id": "G001", "name": "Drug XYZ 10mg",
      "condition": { "dataset": "ADSL", "variable": "TRT01A", "comparator": "EQ", "value": ["Drug XYZ 10mg"] } },
    { "id": "G002", "name": "Placebo",
      "condition": { "dataset": "ADSL", "variable": "TRT01A", "comparator": "EQ", "value": ["Placebo"] } }
  ]
}
```

**Data-driven (visit)** — groups come from the data:

```json
{
  "id": "GF002",
  "name": "Analysis Visit",
  "groupingDataset": "ADLB",
  "groupingVariable": "AVISIT",
  "dataDriven": true
}
```

**Annotated-shell equivalent.** The **column headers** in your tables. A table with columns
"Drug XYZ 10mg" and "Placebo" reflects a `GroupingFactor` with two pre-specified `Groups`.
Each stored `OperationResult` carries a `resultGroup` linking it back to `G001` or `G002`.

---

### 3.6 AnalysisMethod

Defines one or more **statistical operations** that can be applied to any analysis variable.
Methods are reusable across analyses.

**Operations — key fields**

| Field | Description |
|-------|-------------|
| `id`, `name`, `order` | Identifier, name, position within the method |
| `resultPattern` | Display format (e.g., `"XX.X"`, `"XX.XX"`) |
| `referencedOperationRelationships` | How this operation uses results from another (e.g., NUMERATOR / DENOMINATOR for %) |

**Continuous summary**

```json
{
  "id": "M001",
  "name": "Summary Statistics - Continuous",
  "operations": [
    { "id": "O001", "name": "n",      "order": 1, "resultPattern": "XXX"    },
    { "id": "O002", "name": "Mean",   "order": 2, "resultPattern": "XXX.X"  },
    { "id": "O003", "name": "SD",     "order": 3, "resultPattern": "XXX.XX" },
    { "id": "O004", "name": "Median", "order": 4, "resultPattern": "XXX.X"  },
    { "id": "O005", "name": "Min",    "order": 5, "resultPattern": "XXX"    },
    { "id": "O006", "name": "Max",    "order": 6, "resultPattern": "XXX"    }
  ]
}
```

**Count + percentage (uses ReferencedOperationRelationship)**

```json
{
  "id": "M002",
  "name": "Count and Percentage",
  "operations": [
    { "id": "O007", "name": "Count",      "order": 1, "resultPattern": "XXX" },
    { "id": "O008", "name": "Percentage", "order": 2, "resultPattern": "XX.X",
      "referencedOperationRelationships": [
        { "id": "ROR001",
          "referencedOperationRole": { "controlledTerm": "DENOMINATOR" },
          "operationId": "O009",
          "analysisId": "AN_DENOM_001" }
      ] },
    { "id": "O009", "name": "Denominator", "order": 3, "resultPattern": "XXX" }
  ]
}
```

**Annotated-shell equivalent.** *"n (%), denominator = N in safety population"*. The
`AnalysisMethod` encodes this formally; the `resultPattern` encodes your format string;
the `ReferencedOperationRelationship` with `DENOMINATOR` encodes where the N comes from.

---

### 3.7 Analysis — The Central Object

Each `Analysis` ties together: a dataset, a variable, a population, optional data filter,
optional grouping factors, and a method.

**Key fields**

| Field | Description |
|-------|-------------|
| `dataset`, `variable` | The analysis dataset + variable (e.g., `ADAE` + `USUBJID`) |
| `analysisSetId` | References an `AnalysisSet` |
| `dataSubsetId` | References a `DataSubset` (optional) |
| `orderedGroupings` | Ordered list of `GroupingFactor` references |
| `methodId` | References an `AnalysisMethod` |
| `reason` | SPECIFIED IN PROTOCOL / SPECIFIED IN SAP / DATA DRIVEN |
| `purpose` | PRIMARY / SECONDARY / EXPLORATORY OUTCOME MEASURE |
| `documentRefs` | SAP references (with optional page numbers) |
| `programmingCode` | Code, or reference to a program file |
| `results` | The ARD — `[OperationResult]` |

**Shell row → ARM-TS Analysis field mapping**

| Annotated-shell annotation text | ARS field |
|--------------------------------|-----------|
| "USUBJID count from ADAE" | `dataset: "ADAE"`, `variable: "USUBJID"` |
| "Safety Population (SAFFL='Y')" | `analysisSetId: "AS002"` |
| "By treatment arm (TRT01A)" | `orderedGroupings: [{groupingId: "GF001", resultsByGroup: true}]` |
| "n (%)" | `methodId: "M002"` |
| "per Section 10.1 of SAP, pp. 42–44" | `documentRefs: [{referenceDocumentId: "SAP001", pageRefs: [...]}]` |
| "program: t_ae_summary.R" | `programmingCode.documentRef` |

This six-row mapping is the whole game. Every annotated-shell row decomposes into these
fields of one `Analysis` object.

---

### 3.8 Output and OutputDisplay

Defines the TFL — its file format, title, footnotes, and how analyses map into displays.

```json
{
  "id": "OUT001",
  "name": "Table 14.3.1 - Summary of Adverse Events",
  "fileSpecifications": [
    { "name": "Table 14.3.1 RTF", "fileType": { "controlledTerm": "rtf" },
      "location": "output/t_ae_summary.rtf" }
  ],
  "displays": [
    { "order": 1, "display": {
        "id": "DISP001",
        "displayTitle": "Table 14.3.1 Summary of Treatment-Emergent Adverse Events",
        "displaySections": [
          { "sectionType": "Title", "orderedSubSections": [{ "order": 1, "subSection": {
              "id": "TITLE001",
              "text": "Table 14.3.1 Summary of Treatment-Emergent Adverse Events" }}] },
          { "sectionType": "Footnote", "orderedSubSections": [{ "order": 1, "subSection": {
              "id": "FN001",
              "text": "TEAE = Treatment-Emergent Adverse Event. Subjects counted once per preferred term." }}] }
        ] } }
  ]
}
```

**Annotated-shell equivalent.** The **shell itself** — table number, title, footnotes,
abbreviations — maps directly to `Output` and `OutputDisplay`. Your title cell =
`displayTitle`. Your footnote cells = `DisplaySection` with `sectionType: "Footnote"`.

---

### 3.9 AnalysisProgrammingCodeTemplate

When a method is applied to many analyses with **minor variations** (e.g., different
`PARAMCD` values), the `codeTemplate` mechanism lets you write the code once with replacement
parameters.

```json
{
  "id": "M003",
  "name": "KM Analysis Template",
  "codeTemplate": {
    "context": "R version 4.3.1",
    "code": "km_fit <- survfit(Surv({{AVAR}}, {{CNSR_FLAG}}) ~ TRT01A, data = adtte %>% filter(PARAMCD == '{{PARAMCD}}' & ITTFL == 'Y'))",
    "parameters": [
      { "name": "AVAR",      "valueSource": "analysis.variable" },
      { "name": "PARAMCD",   "value": ["OS", "PFS", "DOR"] },
      { "name": "CNSR_FLAG", "value": ["CNSR"] }
    ]
  }
}
```

Individual analyses reference this template and supply their specific parameter values.

---

## Hands-On: Build ARM-TS Objects

### What's in the sample `re_json`?

```{r ch3-preview, exercise=FALSE, echo=TRUE}
# AnalysisSets defined in the sample
sapply(re_json$analysisSets, function(x) paste0(x$id, " — ", x$name))
```

```{r ch3-preview-method, exercise=FALSE, echo=TRUE}
# Operations within Method M001
sapply(re_json$methods[[1]]$operations, function(o)
  paste0(o$id, " — ", o$name, " (", o$resultPattern, ")"))
```

---

## Exercise 3.1 — Read an AnalysisSet WhereClause {data-progressive=FALSE}

Extract the Safety Population (`AS002`) from the sample ARM-TS and print its filter condition
as a human-readable string of the form `"DATASET.VARIABLE COMPARATOR 'VALUE'"`. This is the
core logic of `arsbridge::ars_to_shell()`.

```{r ch3-ex1, exercise=TRUE, exercise.lines=12}
# Step 1: index into analysisSets for AS002 (Safety Population)
as_safe <- re_json$analysisSets[[___]]   # which index?

# Step 2: assemble the readable string from the condition
glue::glue(
  "{as_safe$condition$dataset}.",
  "{as_safe$condition$___} ",            # variable
  "{as_safe$condition$___} ",            # comparator
  "'{as_safe$condition$value[[1]]}'"
)
```

```{r ch3-ex1-hint-1}
# AS001 is ITT (index 1), AS002 is Safety (index 2)
# The condition has: dataset, variable, comparator, value
```

```{r ch3-ex1-solution}
as_safe <- re_json$analysisSets[[2]]
glue::glue(
  "{as_safe$condition$dataset}.",
  "{as_safe$condition$variable} ",
  "{as_safe$condition$comparator} ",
  "'{as_safe$condition$value[[1]]}'"
)
# "ADSL.SAFFL EQ 'Y'"
# This is the core logic of arsbridge::ars_to_shell().
```

```{r ch3-ex1-check}
grade_result(
  pass_if(~ as.character(.result) == "ADSL.SAFFL EQ 'Y'",
          "Correct — that's the Safety Population WhereClause as a human-readable string. arsbridge::ars_to_shell() does exactly this for every AnalysisSet, DataSubset, and Group in the ARM-TS."),
  fail_if(~ TRUE,
          "The Safety Population is AS002 (index 2). Make sure you used $variable and $comparator inside the glue() string.")
)
```

---

## Exercise 3.2 — Build an AnalysisMethod by Hand {data-progressive=FALSE}

Create an R list representing the **"n, mean, SD, median, min, max"** continuous summary method.
Follow the ARS object structure exactly — `id`, `name`, `operations`, and each operation has
`id`, `name`, `order`, `resultPattern`.

```{r ch3-ex2, exercise=TRUE, exercise.lines=20}
m_continuous <- list(
  id   = "M001",
  name = "Summary Statistics - Continuous",
  operations = list(
    list(id = "O001", name = "n",      order = 1L, resultPattern = "XXX"),
    list(id = "O002", name = ___,      order = 2L, resultPattern = "XXX.X"),    # fill in
    list(id = "O003", name = "SD",     order = 3L, resultPattern = ___),         # 2 dp
    list(id = "O004", name = "Median", order = 4L, resultPattern = "XXX.X"),
    list(id = "O005", name = ___,      order = 5L, resultPattern = "XXX"),       # minimum
    list(id = "O006", name = "Max",    order = 6L, resultPattern = "XXX")
  )
)

# How many operations does this method have?
length(m_continuous$operations)
```

```{r ch3-ex2-hint-1}
# The six standard continuous statistics are:
#   n, Mean, SD, Median, Min, Max
# resultPattern uses X placeholders to encode display precision:
#   "XXX"     = integer
#   "XXX.X"   = 1 decimal place
#   "XXX.XX"  = 2 decimal places
```

```{r ch3-ex2-solution}
m_continuous <- list(
  id   = "M001",
  name = "Summary Statistics - Continuous",
  operations = list(
    list(id = "O001", name = "n",      order = 1L, resultPattern = "XXX"),
    list(id = "O002", name = "Mean",   order = 2L, resultPattern = "XXX.X"),
    list(id = "O003", name = "SD",     order = 3L, resultPattern = "XXX.XX"),
    list(id = "O004", name = "Median", order = 4L, resultPattern = "XXX.X"),
    list(id = "O005", name = "Min",    order = 5L, resultPattern = "XXX"),
    list(id = "O006", name = "Max",    order = 6L, resultPattern = "XXX")
  )
)
length(m_continuous$operations)
# 6 operations — this method can now be referenced by any Analysis
# that needs summary stats on a continuous variable.
```

---

## Exercise 3.3 — Build a Group with a WhereClause {data-progressive=FALSE}

You need a `Group` object representing **"Xanomeline High Dose"** within the Treatment Arm
grouping factor. The group is defined by `ADSL.TRT01A == "Xanomeline High Dose"`.

```{r ch3-ex3, exercise=TRUE, exercise.lines=15}
g_high <- list(
  id   = ___,                       # "G001"
  name = "Xanomeline High Dose",
  condition = list(
    dataset    = "ADSL",
    variable   = ___,                # which ADSL variable identifies the treatment arm?
    comparator = ___,                # which ARS comparator?
    value      = list("Xanomeline High Dose")
  )
)

str(g_high)
```

```{r ch3-ex3-solution}
g_high <- list(
  id   = "G001",
  name = "Xanomeline High Dose",
  condition = list(
    dataset    = "ADSL",
    variable   = "TRT01A",
    comparator = "EQ",
    value      = list("Xanomeline High Dose")
  )
)
str(g_high)
# Notice the Group's condition has the SAME structure as an AnalysisSet's condition.
# This is why mastering WhereClause once unlocks AnalysisSet, DataSubset, and Group.
```

---

## Exercise 3.4 — Capstone: Annotated Shell → Analysis Object {data-progressive=FALSE}

You are a lead programmer. Your annotated shell for Table 14.1.1 says:

> *"n, mean, SD, median, min, max from ADSL.AGE, Safety Population (SAFFL='Y'), by treatment
> arm (TRT01A), per SAP Section 8.1"*

Build the `Analysis` object that encodes this annotation.

```{r ch3-ex4, exercise=TRUE, exercise.lines=22}
an_age <- list(
  id      = "AN001",
  name    = "Age Summary by Treatment Arm - Safety Population",
  reason  = list(controlledTerm = "SPECIFIED IN SAP"),
  purpose = list(controlledTerm = ___),       # Primary, Secondary, or Exploratory?
  dataset  = ___,                              # which ADaM dataset?
  variable = ___,                              # which variable?
  analysisSetId = ___,                         # which AnalysisSet ID? (Safety = AS002)
  orderedGroupings = list(
    list(order = 1L, groupingId = ___, resultsByGroup = TRUE)   # which GF?
  ),
  methodId = ___                               # which Method? (n, mean, SD, median, min, max)
)

str(an_age)
```

```{r ch3-ex4-hint-1}
# - It's a demographic baseline summary → PRIMARY OUTCOME MEASURE
# - Source: ADSL.AGE
# - Population: Safety = AS002
# - Grouping: Treatment Arm = GF001
# - Method: continuous summary = M001
```

```{r ch3-ex4-solution}
an_age <- list(
  id      = "AN001",
  name    = "Age Summary by Treatment Arm - Safety Population",
  reason  = list(controlledTerm = "SPECIFIED IN SAP"),
  purpose = list(controlledTerm = "PRIMARY OUTCOME MEASURE"),
  dataset  = "ADSL",
  variable = "AGE",
  analysisSetId = "AS002",
  orderedGroupings = list(
    list(order = 1L, groupingId = "GF001", resultsByGroup = TRUE)
  ),
  methodId = "M001"
)
str(an_age)
# One row of an annotated shell → one Analysis object.
# Field-by-field, the information content is identical.
```

```{r ch3-ex4-check}
grade_result(
  pass_if(~ identical(.result$dataset, "ADSL") &&
            identical(.result$variable, "AGE") &&
            identical(.result$analysisSetId, "AS002") &&
            identical(.result$methodId, "M001") &&
            identical(.result$orderedGroupings[[1]]$groupingId, "GF001"),
          "Correct — that's a complete ARM-TS Analysis object. This is the level of structure that arsbridge::shell_to_ars() will produce from your annotation text."),
  fail_if(~ TRUE,
          "Check: dataset='ADSL', variable='AGE', analysisSetId='AS002' (Safety), groupingId='GF001' (Treatment Arm), methodId='M001'.")
)
```

---

## Chapter 3 Knowledge Check

```{r ch3-quiz, echo=FALSE}
quiz(
  caption = "Chapter 3 — Check Your Understanding",

  question(
    "Where in ARM-TS is the link from an Analysis to a specific SAP section recorded?",
    answer("In ReportingEvent.description"),
    answer("In Analysis.documentRefs, pointing to a ReferenceDocument", correct = TRUE),
    answer("In AnalysisMethod.codeTemplate"),
    answer("In Output.fileSpecifications"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Your shell says: 'PARAMCD = OS in ADTTE'. Which ARM-TS object encodes this?",
    answer("AnalysisSet"),
    answer("DataSubset", correct = TRUE),
    answer("GroupingFactor"),
    answer("AnalysisMethod"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "When `dataDriven = false` on a GroupingFactor, what does that imply?",
    answer("The groups must be pre-specified as `groups` with WhereClauses", correct = TRUE),
    answer("The groups are derived from the data at runtime"),
    answer("Only one group is allowed"),
    answer("The grouping factor is not used in any analysis"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "An AnalysisMethod's `resultPattern` of 'XXX.XX' means what?",
    answer("The operation produces 3 integer digits and 2 decimal digits in the formatted output", correct = TRUE),
    answer("There are 5 operations in the method"),
    answer("The result is censored at 3 significant figures"),
    answer("It is a placeholder for a percentage symbol"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "In Method M002 (Count and Percentage), why does the Percentage operation carry a `referencedOperationRelationships` block?",
    answer(
      "To formally encode that the Percentage operation depends on the Denominator operation (NUMERATOR / DENOMINATOR roles)",
      correct = TRUE
    ),
    answer("To indicate the operation is optional"),
    answer("To specify which programmer wrote the code"),
    answer("To version-control the operation"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "The annotated-shell phrase 'By treatment arm (TRT01A)' maps to which field of an Analysis object?",
    answer("dataset"),
    answer("orderedGroupings (referencing a GroupingFactor)", correct = TRUE),
    answer("analysisSetId"),
    answer("documentRefs"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "What is the purpose of AnalysisProgrammingCodeTemplate?",
    answer(
      "To write code ONCE with replacement parameters (e.g., PARAMCD) and reuse it across many analyses that vary only in those parameters",
      correct = TRUE
    ),
    answer("To store the SAS macro that produced the table"),
    answer("To define the table title and footnote text"),
    answer("To encode the resultPattern for each operation"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),

  question(
    "Which Output sub-object holds the table TITLE and FOOTNOTE text?",
    answer("Output.fileSpecifications"),
    answer("OutputDisplay.displaySections (with sectionType 'Title' / 'Footnote')", correct = TRUE),
    answer("Analysis.documentRefs"),
    answer("ReportingEvent.label"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

---

## Chapter 3 Summary {.congratulations}

You have now seen every major ARM-TS object with:

- **Full field definitions** for `ReportingEvent`, `ReferenceDocument`, `AnalysisSet`,
  `DataSubset`, `GroupingFactor` / `Group`, `AnalysisMethod` / `Operation`, `Analysis`,
  `Output` / `OutputDisplay`, and `AnalysisProgrammingCodeTemplate`
- **JSON examples** grounded in a realistic AE summary table
- **Explicit field-by-field mapping** to what you already write in annotated shells
- **Hands-on practice** building an `AnalysisMethod`, a `Group`, and a full `Analysis` object

**The key insight:** everything in an annotated shell has an ARM-TS equivalent. The information
content is identical; the representation is **machine-readable JSON vs. human-readable
annotation text**.

**Next:** Chapter 4 — ARD Deep Dive (what goes inside `Analysis.results` when analyses are
actually run, and how `{cards}` generates flat ARD data frames from ADaM data)
